### 2017-01-29 paper/paper.tex
push all changes

\documentclass[12pt,titlepage]{article}

% \usepackages21{fancyhdr}

\usepackage[printwatermark]{xwatermark}

\usepackage{grffile}
\usepackage{xcolor}
\usepackage{lipsum}
\usepackage{times}
\usepackage{soul}
\usepackage{epsfig}
\usepackage{rotating}
\usepackage{url}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{longtable}
\usepackage{keyval}
\usepackage{caption,subcaption}
\usepackage{arydshln}
%%\usepackage[hyphenbreaks]{breakurl}

%% allow urls to get broken on hyphens
\usepackage{hyperref}
\def\UrlBreaks{\do\/\do-}

%% APSR submission: no commas in citations between name and year
%% See http://merkel.zoneo.net/Latex/natbib.php
\bibpunct{(}{)}{;}{author-year}{}{;}

% the opening bracket symbol, default = (
% the closing bracket symbol, default = )
% the punctuation between multiple citations, default = ;
% the letter `n' for numerical style, or `s' for numerical superscript style, any other letter for author-year, default = author-year;
% the punctuation that comes between the author names and the year
% the punctuation that comes between years or numbers when common author lists are suppressed (default = ,);

\usepackage{footmisc}
\renewcommand{\footnotelayout}{\doublespacing} % set spacing in footnotes
\newlength{\myfootnotesep}
\setlength{\myfootnotesep}{\baselineskip}
\addtolength{\myfootnotesep}{-\footnotesep}
\setlength{\footnotesep}{\myfootnotesep} % set spacing between footnotes

% make footnote font size same as regular font size in text
\renewcommand{\footnotesize}{\normalsize}

%% Use this for a "DRAFT" watermark
%% \newwatermark[allpages,color=pink!50,angle=45,scale=5,xpos=-25,ypos=40]{DRAFT}

%% List all locations for graphics here
\graphicspath{ {../plots/} }

\begin{document}
\sloppy
\thispagestyle{empty}

%% APSR submission requires double-spaced footnotes
%%\newcommand{\footnote}[1]{\footnote{\doublespacing #1}} %% <-- note \doublespacing here.

\renewcommand{\topfraction}{.85}
\renewcommand{\bottomfraction}{.7}
\renewcommand{\textfraction}{.15}
\renewcommand{\floatpagefraction}{.66}
\renewcommand{\dbltopfraction}{.66}
\renewcommand{\dblfloatpagefraction}{.66}

% \urldef\myurlncsl1\url{foo%.com}
% \begin{document}
% text\footnote{WWW: \myurl}

\title{Was simultaneous marathon finish in Rio coincidental?}\author{David Cottrell\thanks{Postdoctoral Research Fellow,
Program in Quantitative Social Science, Dartmouth College.  6108
Silsby Hall, Hanover, NH 03755(\texttt{david.cottrell@dartmouth.edu}).} \and Michael C.\
Herron\thanks{Visiting Scholar, Hertie School of Governance, Berlin,
Germany, and Professor of Government, Dartmouth College. 6108 Silsby
Hall, Hanover, NH 03755-3547
(\texttt{michael.c.herron@dartmouth.edu}).}}

\maketitle \doublespacing

%\begin{abstract} 
% \noindent
%The abstract
%\end{abstract}

%\newpage

\begin{quote}
 \emph{``I invested all I had and 300 meters before the finish line, I was next to Lisa. It was a magical moment that we could finish this marathon together. We did not think about what we were doing.''€ - Anna Hahner} 
\end{quote}

\section*{Introduction}

On August 14, 2016 at 8:30 in the morning (check) the Women's marathon kicked off in Rio De Janeiro, Brazil, when 156 olympic runners from 80 countries around the world took off from the starting line en route to their destination 42.195 kilometers away.  Two hours, twenty-four minutes, and four seconds later, Jemima Sumgong of Kenya would be the first to cross the finish line and take home gold.  She was just 3.5 minutes behind her personal best.   Approximately 21 minutes later, Germany's twin marathoners,  Anna and Lisa Hahner, would cross the finish line together, holding hands and celebrating a personal victory.  Although they would finish 81st and 82nd respectively and record a time slightly more than 18 minutes behind their personal bests, Anna Hahner would describe the event as a ``magical moment.'' \href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}   The media quickly picked up on the story as an image of the two twins finishing hand-in-hand with beaming smiles captured a public audience.  Many believed that the moment was a reflection of the Olympic spirit.

However, not everyone agreed.  The sports director of the German Athletics Federation, Thomas Kurschilgen, immediately stirred up controversy when he suggested that the twins' photo-finish was no coincidence.  He stated that their main goal was to create a spectacle and ``generate media attention,'' using the fact that they were at least 18 minutes behind their personal best times as evidence that they were slowing their pace in order to finish simultaneously.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}    Of course, these accusations were denied by Anna and Lisa Hahner, who claimed that their simultaneous finish was simply an unintended coincidence.

We wondered what the data said about this likelihood.  Did Anna and Lisa Hahner finish the race together by chance or was it something more calculated?

\section*{The probability of an unintentional simultaneous finish}

Thomas Kurschilgen's accusations of an intentional finish stems from two observations about the race.  First, the twins finished slower than he expected, recording a time that is more than 18 minutes behind their personal best.  Second, the twins finished together at the exact same time, which seemed like an unlikely coincidence.   Kurschilgen clearly believed that neither of these events would have occurred had Anna and Lisa run the marathon independently, absent any coordination. Instead, the twins' would have run at a faster pace and they would not have finished simultaneously.

On the other hand, only a handful of runners completed the marathon in Rio with times that were faster than their recorded best time. Therefore, 18 minutes behind their personal record may not be the outlier that Kurschilgen claimed it to be.  Moreover, there is good reason to suspect that an unintended simultaneous finish is more likely for Anna and Lisa than it would be for any other set of runners. After all, they are twins with similar abilities.  Not only do they train together, but their record best times are less than two minutes apart.  While Anna may be slightly faster than Lisa, given their personal bests, we might expect the difference in their result to be just as close as the difference in the their recorded bests.   And given random variation in finishing - some days are better than others - a simultaneous finish might not be out of the ordinary.

To test the assertion that Anna and Lisa Hahner intentionally slowed their pace to finish simultaneously, we simply need to compute the probability that such a result would have occurred unintentionally.  Hence we need to know the probability distribution of Anna and Lisa's finishing times given that their runs were completely independent of each other.    The challenge, of course, is that this distribution is unknown.  Therefore it needs to be estimated.

\section*{Modeling}

%% THIS SECTION IS A MESS.  I JUST WANTED TO PUT SOMETHING IN WRITING.

We know that Anna and Lisa Hahner finished the marathon in Rio simultaneously.  We don't know, however, if the finish was intentional; either the twins acted independently $I$ or they acted in coordination $C$.  These are the two possible states of the race $\psi = \{I, C\}$.  Given that we observed Anna and Lisa's final times, $Y_A =  y_A$ and $Y_L = y_L$, we want to know the probability that they ran independently, as they say they did. Hence, we are looking to determine,

$$P(\psi = I \mid y_A  \cap y_L )$$

However, to determine this, we need to have some understanding of the likelihood function.  We want to know the likelihood of Anna's and Lisa's final times given independence $P(y_A  \cap y_L \mid \psi = I )$.  
We can estimate this function by beginning with a few assumptions.  First, we assume that under independence, any given runner's final time $Y_i$ is conditional on his/her running ability plus noise.   Specifically, we assume that the $Y_i$ is a linear function of the runner's ability $X_i$ plus a normally distributed error term $e_i \sim N(0, \sigma)$.   Second, we assume that every runner shares the same linear relationship - meaning the slope and intercept remain constant across runners.   Third, we assume that the error term is constant across runners as well.  Hence, luck and misfortune are drawn from the same distribution.  Therefore,

$$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e, \sigma)$$

We also assume that a runner's ability $X_i$ can be measured precisely by their best marathon performance leading up to the Olympics. Any measurement error must therefore be negligible.

If Anna and Lisa intentionally slowed down as a result of coordination then we would likely observe $y_A  >  E(Y_A \mid \psi = I) $ and  $y_L  >  E(Y_L\mid \psi = I )$.  In other words, the final times that Anna and Lisa recorded in the race would be greater than we would expect if they had run independently.

Moreover, if Anna and Lisa coordinated to finish simultaneously, then the difference between the two sisters' final times would be less that the expected difference had they run independently.  Therefore, under a coordinated finish we would expect $\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid  \psi = I )$.

\section*{Did Anna and Lisa intentionally slow down?}
According to Kurschilgen, Anna and Lisa underperformed in the Rio marathon. He claimed that because their goal was to finish simultaneously rather than finish at their fastest pace, their times were slower than they otherwise would have been.  He claimed that the twins were simply trading speed for a photo-finish.

If this were the case, the function generating Anna and Lisa's final times would deviate from the function that generated everyone else's final times.  Given everyone else would draw their times from the independent distribution  $Y_i \sim N(\beta_0 + X_{i}\beta_1 + e, \sigma)$,  Anna and Lisa would draw from a distribution of times that are slower in expectation. Hence, they would lie well-above the line that links a runner's performance in Rio to their previous best performance.

\begin{figure}[!ht]
 \caption{Relationship between Personal Best and Result}
 \label{fig:scatter}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
 \end{subfigure}
 \begin{subfigure}{.5\textwidth}
 %\label{fig:Studentized Residuals}
 \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
 \end{subfigure}
\end{figure}

The scatter plot in Figure \ref{fig:scatter} displays the relationship between a runner's personal best and their final result.   Of the 156 runners that competed in the marathon event, 133 finished the race.  These individuals are plotted as grey points.  The 23 that did not finish are plotted below along the x-axis as grey tick-marks.   The dashed black line with grey confidence intervals represent OLS estimates of the linear relationship between the two variables.   The Hahner twins are marked by orange points.

One can see that there is a clear linear relationship between a runner's performance in the olympics and her final result.  Moreover, there is a significant amount of noise in the relationship.  This suggests that despite one's innate abilities, luck plays a major role in olympic finishes.    Still, the twins' results are well above the fitted line, even accounting for typical error.  They are much slower than the underlying linear relationship suggests.  In fact, if we order each point by the size of the studentized residual (see the second plot), we can see that the Hahner twins are in the tail-end of the distribution.  Anna Hahner's residual deviation is greater than 91.7\% of the runners who completed the race while Lisa Hahner's deviation is greater than 86.5\%.  Hence, the twins appear to have finished at a much slower pace than expectation, which is what we might expect if they were coordinating their runs.\footnote{It is important to note that the North Korean twins finished much faster than expected.  XXXX explain PRK}

\section*{Did Anna and Lisa intentionally finish together?}

\begin{figure}[!ht]
 \caption{Relationship between Difference in Personal Best and Difference in Result}
 \label{fig:diffdiff}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_1.pdf}
 \end{subfigure}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_2.pdf}
 \end{subfigure}
\end{figure}

\subsection*{Simulate the run}

\begin{enumerate}
\item Removing the twins from the sample, estimate a linear model that
predicts a runner?s final time (y) based on her personal best time (x).
\item Extract the coefficients and the variance-covariance matrix from this
model.
\item For each runner, draw $\beta_0$ and $\beta_1$ from a multivariate normal
distribution with mean equal to the coefficients and variance equal to the
variance-covariance matrix.
\item For each runner, draw an error term ($e$) from a normal distribution with
mean equal to zero and a standard deviation equal to the standard
deviation of the model?s residuals.
\item For each runner, predict the final result by combining the randomly
generated beta coefficients and error terms, $\hat{y} = \beta_0 + \beta_1 + e$.
\item Eliminate each runner from the race with a probability equal to the
percent of the total runners who did not finish in the actual race.  This
account for the likelihood of a DNF.
\item Calculate the runner?s rank relative to the other runners.
\item From the rankings, determine if each set of twins finished
consecutively.
\item Repeat steps 2 through 8 ten thousand times.
\item Calculate the percent of the simulation in which the twins finished
consecutively
\end{enumerate}

\begin{figure}[!ht]
 \caption{The distribution of simulated results}
 \label{fig:diffdiff}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{simulated_time.pdf}
 \end{subfigure}
 \begin{subfigure}{.5\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
 \end{subfigure}
\end{figure}

\end{document}

### 2017-01-29 2-simulate.R

@@ -1,10 +1,114 @@
 library(MASS)
 library(tidyverse)
 library(lubridate)
-df <- read.csv("data/times.csv", stringsAsFactors = FALSE)
+library(stringr)
+df <- read.csv("data/times.csv", stringsAsFactors = FALSE) %>% tbl_df()
+
+# Remove Irvette Van Zyl, since she did not start the race due to injury.  Her bib number is 1172.
+df <- df[df$BIB != 1172,]
+
+pdf("plots/scatter_plot.pdf", height = 5,  width = 5)
+ggplot(aes(y=FINAL, x=PB), data = df) + 
+  geom_point(size = 1, colour = "gray") + 
+  geom_point(data = df[!is.na(df$TWINS),], aes( col = TWINS)) +
+  geom_smooth(method = "lm", colour = "black", alpha = .5, size = .5, linetype = "dashed") + 
+  geom_abline(intercept = 0, slope =1) +
+  geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
+  geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
+  coord_fixed(xlim = c(8000,12500), ylim = c(8000,12500)) +
+  ylab("Result (Sec)\n") + 
+  xlab("\nPersonal Best (Sec)") + 
+  ggtitle("Distribution of Y | X") +
+  scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
+  theme_bw() +
+  theme(legend.position = c(.8,.2))
+dev.off()
+
+nms <- df$ATHLETE
+pb <- df$PB
+result <- df$FINAL
+rank <- min_rank(df$FINAL)
+name_i <- vector()
+name_j <- vector()
+pb_diff <- vector()
+result_diff <- vector()
+consecutive <- vector()
+k = 1
+a = 1
+for(i in 1:length(pb)){
+  for(j in a:length(result)){
+    name_i[k] = nms[i]
+    name_j[k] = nms[j]
+    pb_diff[k] = pb[i] - pb[j]
+    result_diff[k] = result[i] - result[j]
+    consecutive[k] = ifelse(abs(rank[i] - rank[j]) == 1, T, F) 
+    k = k + 1
+  }
+  a = a + 1 
+}
+
+
[…]

### 2017-01-29 paper/paper.tex
update

@@
 
-\subsection*{Simulate the run}
+\newpage
+\subsection*{Steps of the simulation}
 
@@
 \item Removing the twins from the sample, estimate a linear model that
-predicts a runner?s final time (y) based on her personal best time (x).
+predicts a runner?s final time ($Y_i$) based on her personal best time ($X_i$).
 \item Extract the coefficients and the variance-covariance matrix from this
@@
 mean equal to zero and a standard deviation equal to the standard
-deviation of the model?s residuals.
+deviation of the model?s residuals ($\sigma$).
 \item For each runner, predict the final result by combining the randomly
@@
 account for the likelihood of a DNF.
-\item Calculate the runner?s rank relative to the other runners.
-\item From the rankings, determine if each set of twins finished
-consecutively.
-\item Repeat steps 2 through 8 ten thousand times.
-\item Calculate the percent of the simulation in which the twins finished
-consecutively
+\item Calculate the difference in time between Anna Hahner and Lisa Hahner.
+\item Calculate the difference in ranking between Anna Hahner and Lisa Hahner.
+\item Repeat steps 3 through 8 ten thousand times.
+\item Plot a histogram of the results in Figure \ref{fig:simdiff}
 \end{enumerate}
@@
  \caption{The distribution of simulated results}
- \label{fig:diffdiff}
+ \label{fig:simdiff}
  \begin{subfigure}{.5\textwidth}

### 2017-01-29 paper/paper.tex
some paper edits

@@
 
+\newcommand{\yi}{\ensuremath{Y_i}}
+
 % \urldef\myurlncsl1\url{foo%.com}
@@
 
-\title{Was simultaneous marathon finish in Rio coincidental?}\author{David Cottrell\thanks{Postdoctoral Research Fellow,
-Program in Quantitative Social Science, Dartmouth College.  6108
-Silsby Hall, Hanover, NH 03755(\texttt{david.cottrell@dartmouth.edu}).} \and Michael C.\
+\title{\Large{All in the family:\\Anna and Lisa Hahner's finishing
+    times in\\the 2016 women's Olympic
+  marathon}}\author{David Cottrell\thanks{Postdoctoral Research
+  Fellow, Program in Quantitative Social Science, Dartmouth College,
+    6108 Silsby Hall, Hanover, NH
+    03755 (\texttt{david.cottrell@dartmouth.edu}).} \and Michael C.\
   Herron\thanks{Visiting Scholar, Hertie School of Governance, Berlin,
-Germany, and Professor of Government, Dartmouth College. 6108 Silsby
-Hall, Hanover, NH 03755-3547
+    Germany, and Professor of Government, Dartmouth College, 6108
+    Silsby Hall, Hanover, NH 03755
     (\texttt{michael.c.herron@dartmouth.edu}).}}
@@
 
-On August 14, 2016 at 8:30 in the morning (check) the Women's marathon kicked off in Rio De Janeiro, Brazil, when 156 olympic runners from 80 countries around the world took off from the starting line en route to their destination 42.195 kilometers away.
-Two hours, twenty-four minutes, and four seconds later, Jemima Sumgong of Kenya would be the first to cross the finish line and take home gold.
-She was just 3.5 minutes behind her personal best.
-Approximately 21 minutes later, Germany's twin marathoners,  Anna and Lisa Hahner, would cross the finish line together, holding hands and celebrating a personal victory.
-Although they would finish 81st and 82nd respectively and record a time slightly more than 18 minutes behind their personal bests, Anna Hahner would describe the event as a ``magical moment.'' \href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}   The media quickly picked up on the story as an image of the two twins finishing hand-in-hand with beaming smiles captured a public audience.
-Many believed that the moment was a reflection of the Olympic spirit.
+On August 14, 2016, at 9:30 in the morning, the Women's Olympic
+marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
+80 countries across the world took off from the starting line en route
+to their destination 42.195 kilometers away. Two hours, twenty-four
+minutes, and four seconds later, Jemima Sumgong of Kenya would be the
+first to cross the finishline and take home gold; Sumgong was just 3.5
+minutes behind her personal best time in the marathon. Approximately
+21 minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
+would cross the finishline together, holding hands and celebrating a
+personal victory. Although the Hahners would finish 81st and 82nd,
+respectively, with times slightly more than 18 minutes slower than
+corresponding personal bests, Anna Hahner would describe their joint
+finish as a ``magical moment.''
+\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
+The media quickly picked up on the Hanher story as an image of the
+beaming twins finishing hand-in-hand captured a public audience. Many
+believed the moment was a reflection of the Olympic spirit.
 
-However, not everyone agreed.
-The sports director of the German Athletics Federation, Thomas Kurschilgen, immediately stirred up controversy when he suggested that the twins' photo-finish was no coincidence.
-He stated that their main goal was to create a spectacle and ``generate media attention,'' using the fact that they were at least 18 minutes behind their personal best times as evidence that they were slowing their pace in order to finish simultaneously.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}    Of course, these accusations were denied by Anna and Lisa Hahner, who claimed that their simultaneous finish was simply an unintended coincidence.
+Not everyone agreed with this rosy interpretation. The twins' happy
+facial expressions at the finish were portrayed as a bit contrived
+(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
+cookie in the shape of a horse), and the sports director of the German
+Athletics Federation, Thomas Kurschilgen, stirred up controversy when
+he suggested that the Hahners' photo-finish was no coincidence.
+Kurschilgen averred that the twins slowed down so as to finish
+simultaneously and sought to create a spectacle and ``generate media
+attention.'' Kurschilgen justified his charge with the fact that the
+twins were at least 18 minutes behind their personal best
+times.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
+Not surprisingly, these accusations were denied by the Hahners, who
+claimed that their simultaneous finish was simply an unintended
+coincidence.
 
-We wondered what the data said about this likelihood.  Did Anna and Lisa Hahner finish the race together by chance or was it something more calculated?
+%% Honigkuchenpferd quote from here:
+%%
+%% https://www.welt.de/sport/olympia/article157669264/Das-falsche-Laecheln-der-deutschen-Lauf-Zwillinge.html
+
+What happened in the women's Olympic marathon, and how might we assess
+whether the Hahner finish was coincidental or intentional? These two
+interpretations are clearly at odds. If the former, then the Hahner
+twins are to be celebrated and their finish treated as an expression
+of the spirit behind the Olympic games. If the latter, though, then
+the Hahner twins may have violated this spirt by not trying hard
+enough. It is perhaps too easy for us to write such a glib
+sentence---neither of us cannot fathom being able to complete a
+marathon anywhere in the vicinity of two and a half hours---but we
+nonetheless want to know what the data from the Olympic marathon data
+tell us. Was the women's Olympic marathon a lovely coincidence or
+something else?
+
+Among female Olympic marathoners, the Hahner twins were not alone in
+their familial ties, and we draw on this in the analysis that follows.
+The marathon also featured twins from North Korea, Kim Hye-song and
+Kim Hye-gyong, who had identical times and finished 10th and 11th,
+respectively. The Kim finish, unlike the Hahner finish, appears devoid
+of post-race controversy. Moreover, three triplets from Estonia
+competed in the Rio marathon, although only two, Lily Luik and Leila
+Luik, finished it, in 97th and 114th place, respectively. The third
+Estonia triplet, Liina Luik, recorded what is known as a DNF---an
+abbreviation that means did not finish, a term that we will use
+throughout this article.
+
+\section*{Marathon data}
+
+For each participant who started the women's Olympic marathon, we know
+several things: personal best marathon time prior to the 2016 Olympic
+games; split times from the Rio marathon course at 5 kilometers, 10
+kilometers, and so forth; and, finishing time. We cannot directly
+observe the effort that an individual put into the race, and we do not
+know why some runners have DNF results; some runners may have injured
+themselves on the course and accordingly dropped out, and others may
+have dropped out, uninjured, in anticipation of an unsatisfactory
+result. Of the 156 marathon starters, 133 completed the race and 23
+DNFed at various locations throughout the course. The overall DNF rate
+was thus $\frac{23}{156} \approx 0.15$, and the relatively small
+sample size at our disposal---156 runners---means that a 95\%
+confidence for this rate is fairly wide, namely,
+$\left(0.098, 0.22\right)$.
+
+Key in the analysis that follows is the deviation between a runner's
+Rio time and her prior personal best time in the marathon. Of course
+there is variance across runners in these deviations, and we want to
+know if patterns in the deviations support an honest effort by the
+Hanher twins or perhaps something else.
 
@@
 
-Thomas Kurschilgen's accusations of an intentional finish stems from two observations about the race.
-First, the twins finished slower than he expected, recording a time that is more than 18 minutes behind their personal best.
-Second, the twins finished together at the exact same time, which seemed like an unlikely coincidence.
-Kurschilgen clearly believed that neither of these events would have occurred had Anna and Lisa run the marathon independently, absent any coordination.
-Instead, the twins' would have run at a faster pace and they would not have finished simultaneously.
+Thomas Kurschilgen's accusations of an intentional finish stems from
+two observations. First, the Hahner twins finished slower than he
+expected, as we already noted recording times more than 18 minutes
+slower than respective personal bests. Second, the twins finished
+together at the exactly same time. Kurschilgen clearly believed that
+neither of these events would have occurred had both Hahner run the
+marathon independently, absent coordination. By his logic, the German
+twins should have run faster and not have finished simultaneously.
 
-On the other hand, only a handful of runners completed the marathon in Rio with times that were faster than their recorded best time.
-Therefore, 18 minutes behind their personal record may not be the outlier that Kurschilgen claimed it to be.
-Moreover, there is good reason to suspect that an unintended simultaneous finish is more likely for Anna and Lisa than it would be for any other set of runners.
-After all, they are twins with similar abilities.
-Not only do they train together, but their record best times are less than two minutes apart.
-While Anna may be slightly faster than Lisa, given their personal bests, we might expect the difference in their result to be just as close as the difference in the their recorded bests.
-And given random variation in finishing - some days are better than others - a simultaneous finish might not be out of the ordinary.
+On the other hand, only a handful of runners completed the Rio
+marathon with times that were faster than their recorded best times.
+Thus, 18 minutes behind a personal record may not be the outlier that
+Kurschilgen claimed it to be. Moreover, there is good reason to
+suspect that an unintended simultaneous finish was more likely for the
+Hahner twin than it would be for any other set of runners. After all,
+these two women are twins with presumably similar abilities. Not only
+do they train together, but their pre-Rio best times are less than two
+minutes apart. While Anna may be slightly faster than Lisa, measured
+by personal bests, we might expect the difference in their Rio result
+to be just as close as the differences in their recorded bests. And
+given random variation in finishing times, a simultaneous finish might
+not be out of the ordinary.
 
-To test the assertion that Anna and Lisa Hahner intentionally slowed their pace to finish simultaneously, we simply need to compute the probability that such a result would have occurred unintentionally.
-Hence we need to know the probability distribution of Anna and Lisa's finishing times given that their runs were completely independent of each other.
-The challenge, of course, is that this distribution is unknown.
-Therefore it needs to be estimated.
+Even though they seem similar, finishing at very similar times and
+finishing together are different phenomena. If, for example, both
+Hahners were of similar ability to each other and also to many other
+runners, then we might expect similar finishing times yet not
+necessarily similar placements. The latter will be a function of the
+extent to which all runners on the Rio course have similar talent
+levels.  This point is an important one and will be evident in the
+results that follow.
+
+To test the assertion that the Hahner twins paced themselves to finish
+at the same time and with back-to-back placements, we simply need to
+compute the probability that such a result would have occurred
+unintentionally. Hence we need to know the probability distribution of
+Anna and Lisa's finishing times if their runs had been independent of
+each other. The challenge, of course, is that this distribution is
+unknown.  ^^ Therefore it needs to be estimated.
 
@@
 
-We know that Anna and Lisa Hahner finished the marathon in Rio simultaneously.
-We don't know, however, if the finish was intentional; either the twins acted independently $I$ or they acted in coordination $C$.
-These are the two possible states of the race $\psi = \{I, C\}$.
-Given that we observed Anna and Lisa's final times, $Y_A =  y_A$ and $Y_L = y_L$, we want to know the probability that they ran independently, as they say they did.
-Hence, we are looking to determine,
+%% Ok.  Not sure what all the notation means.
+
+% We know that Anna and Lisa Hahner finished the marathon in Rio
+% simultaneously. We don't know, however, if the finish was intentional;
+
+Suppose that Hahner the twins acted independently $I$ or they acted in
+coordination $C$; these are the two possible states of the race
+$\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
+times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
+that they ran independently, as they say they did. Hence, we are
+looking to determine,  %% y_a versus Y_a.  Former realized?
 
@@
 
-However, to determine this, we need to have some understanding of the likelihood function.
-We want to know the likelihood of Anna's and Lisa's final times given independence $P(y_A  \cap y_L \mid \psi = I )$.
-We can estimate this function by beginning with a few assumptions.
-First, we assume that under independence, any given runner's final time $Y_i$ is conditional on his/her running ability plus noise.
-Specifically, we assume that the $Y_i$ is a linear function of the runner's ability $X_i$ plus a normally distributed error term $e_i \sim N(0, \sigma)$.
-Second, we assume that every runner shares the same linear relationship - meaning the slope and intercept remain constant across runners.
-Third, we assume that the error term is constant across runners as well.
-Hence, luck and misfortune are drawn from the same distribution.
-Therefore,
+However, to determine this, we need to have some understanding of a
+likelihood function that specifies Rio finishing times. We want to
+know the likelihood of Anna's and Lisa's final times given
+independence $P(y_A \cap y_L \mid \psi = I )$. We can estimate this
+function with a few assumptions. First, we assume that under
+independence, any given runner's final time $Y_i$ is conditional on
+his/her running ability plus noise. Specifically, we assume that the
+$Y_i$ is a linear function of the runner's ability $X_i$ plus a
+normally distributed error term $e_i \sim N(0, \sigma)$. Second, we
+assume that every runner shares the same linear relationship - meaning
+the slope and intercept remain constant across runners. Third, we
+assume that the error term is drawn from a common distribution across
+runners. Hence, luck and misfortune are drawn from the same
+distribution. Therefore,
 
-$$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e, \sigma)$$
+$$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e_i, \sigma)$$
 
-We also assume that a runner's ability $X_i$ can be measured precisely by their best marathon performance leading up to the Olympics. Any measurement error must therefore be negligible.
+We also assume that a runner's ability $X_i$ can be measured precisely
+by her best marathon performance leading up to the Olympics.
 
-If Anna and Lisa intentionally slowed down as a result of coordination then we would likely observe $y_A  >  E(Y_A \mid \psi = I) $ and  $y_L  >  E(Y_L\mid \psi = I )$.
-In other words, the final times that Anna and Lisa recorded in the race would be greater than we would expect if they had run independently.
+%% Any measurement error must therefore be negligible.
 
-Moreover, if Anna and Lisa coordinated to finish simultaneously, then the difference between the two sisters' final times would be less that the expected difference had they run independently.
-Therefore, under a coordinated finish we would expect $\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid  \psi = I )$.
+If Anna and Lisa intentionally slowed down as a result of coordination
+then we would likely observe $y_A > E(Y_A \mid \psi = I) $ and
+$y_L > E(Y_L\mid \psi = I )$. In other words, the final times that
+Anna and Lisa recorded in the race would be greater than we would
+expect if they had run independently.
+
+Moreover, if Anna and Lisa coordinated to finish simultaneously, then
+the difference between the two sisters' final times would be less that
+the expected difference had they run independently. Therefore, under a
+coordinated finish we would expect
+$\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid \psi = I )$.
 
 \section*{Did Anna and Lisa intentionally slow down?}
+
 According to Kurschilgen, Anna and Lisa underperformed in the Rio marathon.
@@
 \begin{figure}[!ht]
- \caption{Relationship between Personal Best and Result}
+  \caption{Relationship between Personal Best and Result  -- better label?}
   \label{fig:scatter}
   \begin{subfigure}{.5\textwidth}
- \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
+    \includegraphics[width=\textwidth,
+    keepaspectratio]{scatter_plot.pdf}
+    \caption{Rio finishing times and personal best times}
+    \label{fig:A}
   \end{subfigure}
   \begin{subfigure}{.5\textwidth}
- %\label{fig:Studentized Residuals}
+    % \label{fig:Studentized Residuals}
     \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
+    \caption{A}
+    \label{fig:B}
   \end{subfigure}

### 2017-01-29 2-simulate.R
some paper and code edits

@@ -16,9 +16,9 @@ ggplot(aes(y=FINAL, x=PB), data = df) +
   geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
   geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
   coord_fixed(xlim = c(8000,12500), ylim = c(8000,12500)) +
-  ylab("Result (Sec)\n") + 
-  xlab("\nPersonal Best (Sec)") + 
-  ggtitle("Distribution of Y | X") +
+  ylab("Olympic result in seconds\n") + 
+  xlab("\nPersonal Best in seconds") + 
+    ##  ggtitle("Distribution of Y | X") +
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +
   theme(legend.position = c(.8,.2))
@@ -64,7 +64,7 @@ ggplot(df2b, aes(PERCENTILE,diff_in_diff)) +
   geom_point(data = na.omit(df2b), aes( col = twins)) + 
   geom_text(data = df2b[df2b$twins == "Hahner Twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_y = 100, colour = "orange") +
   geom_text(data = df2b[df2b$twins == "Kim Twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_y = 100, colour = "red") +
-  ggtitle("Distribution of Difference in Differences (All Diads)") +
+  ggtitle("Distribution of Difference in Differences (All Dyads)") +
   ylab("Difference in Difference (sec)\n") +
   xlab("\nPrecentile") + 
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
@@ -99,8 +99,8 @@ ggplot(dta, aes(PERCENTILE,STUD_RESID)) +
   geom_point(data = na.omit(dta), aes( col = TWINS)) + 
   geom_text(data = dta[dta$TWINS == "Hahner Twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_x = -10, colour = "orange") +
   geom_text(data = dta[dta$TWINS == "Kim Twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_x = 10, colour = "red") +
-  ggtitle("Distribution of Studentized Residuals") +
-  ylab("Studentized Residuals\n") +
[…]

### 2017-01-29 seconds-behind.R

@@ -1,0 +1,59 @@
+library(dplyr)
+
+dta <- read.csv (file = "data/times.csv", header = TRUE)
+
+pdf (file = "plots/seconds-behind.pdf")
+
+plot (0,0, cex = 0, xlim = c(1, 10), ylim = c(0, 2000),
+      xlab = "Split", ylab = "Seconds behind",
+      xaxt = "n")
+
+for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
+    dta2use <- dta[!is.na(dta[,col2use]), ]
+
+    ## set colors and shapes; have to do this for each split since the
+    ## rows for the matrix being printed change per split when runners
+    ## drop out
+    cex2use <- rep(0.12, times = nrow(dta))
+    cex2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- 1
+    cex2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- 1
+    cex2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- 1
+    
+    pch2use <- rep(19, times = nrow(dta2use))
+    pch2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- 19
+    pch2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- 19
+    pch2use[dta2use$BIB == 633  | dta2use$BIB == 634 | dta2use$BIB == 635] <- 19
+
+    color2use <- rep("grey50", times = nrow(dta2use))
+    color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- "orange"
+    color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- "red"
+    color2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- "blue"
+
+    points (x = rep (match(col2use, cols2use), times = nrow(dta2use)), y = dta2use[, col2use] - min(dta2use[, col2use]),
+           col = color2use,
+           cex = cex2use,
+           pch = pch2use            
+           )
+}
+
+axis (side = 1, at = c(1:10),
+      labels = c("5K", "10K", "15K", "20K", "Half", "25K", "30K", "35K", "40K", "Final"))
+
+legend (x = 1, y = 2000,
+        col = c("blue", "orange", "red", "grey50"),
+        pch = 19,
[…]

### 2017-01-29 paper/paper.tex
other edits

@@
 \begin{quote}
- \emph{``I invested all I had and 300 meters before the finish line, I was next to Lisa.
-It was a magical moment that we could finish this marathon together.
-We did not think about what we were doing.''€ - Anna Hahner}
+  \emph{``I invested all I had and 300 meters before the finish line,
+    I was next to Lisa. It was a magical moment that we could finish
+    this marathon together. We did not think about what we were
+    doing.'' -- Anna Hahner}
 \end{quote}
@@
 
+As a brief overview of the race, Figure \ref{fig:secondsbehind}
+describes each runner's status at the various split times on the
+marathon course.  Each dot in the figure depicts a recorded split and
+the number of seconds each runner was behind the race leader at the
+time.  There are more dots at earlier splits due to the accumulation
+of DNFs.  Regarding twins and triplets, the Estonian Luik triplets are
+denoted with blue dots; there are three blue dots through the 20
+kilometer split, but the third triplet did not record a half-marathon
+split.  The orange dots in the figure represent the Hahner twins; Lisa
+Hahner was ahead of her sister through 15 kilometers at which point
+Anna surged ahead.  There are two red dots representing the North
+Korean Kim twins although this is not visually apparent; the Kim twins
+had identical split times during the marathon.
+
+\begin{figure}[!ht]
+  \centering
+  \caption{Runner status by split}
+  \label{fig:secondsbehind}
+  \includegraphics[scale = 1]{seconds-behind.pdf}
+  \begin{flushleft}
+    \emph{Note: each dot represents one runner at a split.  DNFs are
+      not pictured, and splits are not to scale.}
+  \end{flushleft}
+\end{figure}
+
+Figure \ref{fig:secondsbehind} does not directly shed any light on our
+motivating research question, but it does show that the Hahner twins
+finished amidst many other runners.  We will come back to this point
+later.
+
 \section*{The probability of an unintentional simultaneous finish}
@@
 two observations. First, the Hahner twins finished slower than he
-expected, as we already noted recording times more than 18 minutes
-slower than respective personal bests. Second, the twins finished
+expected given respective personal bests. Second, the twins finished
 together at the exactly same time. Kurschilgen clearly believed that
@@
 do they train together, but their pre-Rio best times are less than two
-minutes apart. While Anna may be slightly faster than Lisa, measured
-by personal bests, we might expect the difference in their Rio result
-to be just as close as the differences in their recorded bests. And
-given random variation in finishing times, a simultaneous finish might
-not be out of the ordinary.
+minutes apart. While Anna may be slightly faster than Lisa measured by
+personal bests, we might expect the difference in their Rio result to
+be just as close as the differences in their recorded bests. And given
+random variation in finishing times, a simultaneous finish might not
+be out of the ordinary.
 
@@
 To test the assertion that the Hahner twins paced themselves to finish
-at the same time and with back-to-back placements, we simply need to
-compute the probability that such a result would have occurred
+at the same time and with back-to-back placements, we need to compute
+the probability that such a result would have occurred
 unintentionally. Hence we need to know the probability distribution of
@@
 each other. The challenge, of course, is that this distribution is
-unknown.  ^^ Therefore it needs to be estimated.
+unknown.  % Therefore it needs to be estimated.
 
@@
 
-%% Ok.  Not sure what all the notation means.
+%% Ok!  Not sure what all the notation means.
 
@@
 
-Suppose that Hahner the twins acted independently $I$ or they acted in
-coordination $C$; these are the two possible states of the race
+The Hahner twins either acted independently $I$ or coordinated their
+times $C$; these are the two possible states of the race
 $\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
@@
 \begin{figure}[!ht]
-  \caption{Relationship between Personal Best and Result  -- better label?}
+  \caption{Relationship between Personal Best and Result  -- caption?}
   \label{fig:scatter}
   \begin{subfigure}{.5\textwidth}
+    \centering
     \includegraphics[width=\textwidth,
@@
     \caption{Rio finishing times and personal best times}
-    \label{fig:A}
+    \label{fig:45degreeplot}
   \end{subfigure}
   \begin{subfigure}{.5\textwidth}
+    \centering
     % \label{fig:Studentized Residuals}
     \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
-    \caption{A}
-    \label{fig:B}
+    \caption{Studentized residuals}
+    \label{fig:studentizedresiduals}
   \end{subfigure}
@@
 
-The scatter plot in Figure \ref{fig:scatter} displays the relationship between a runner's personal best and their final result.
-Of the 156 runners that competed in the marathon event, 133 finished the race.
-These individuals are plotted as grey points.
-The 23 that did not finish are plotted below along the x-axis as grey tick-marks.
-The dashed black line with grey confidence intervals represent OLS estimates of the linear relationship between the two variables.
-The Hahner twins are marked by orange points.
+Figure \ref{fig:45degreeplot} displays the relationship between a
+runner's personal best (horizontal axis) and her final result
+(vertical axis).  The 133 finishers are plotted as grey points, and
+the 23 DNFs are depicted below along the horizontal axis with
+tick-marks.  The dashed black line with grey confidence intervals
+represent ordinarily least squares estimates of the linear
+relationship between racer's finishing and personal best times.
 
-One can see that there is a clear linear relationship between a runner's performance in the olympics and her final result.
-Moreover, there is a significant amount of noise in the relationship.
-This suggests that despite one's innate abilities, luck plays a major role in olympic finishes.
-Still, the twins' results are well above the fitted line, even accounting for typical error.
-They are much slower than the underlying linear relationship suggests.
-In fact, if we order each point by the size of the studentized residual (see the second plot), we can see that the Hahner twins are in the tail-end of the distribution.
-Anna Hahner's residual deviation is greater than 91.7\% of the runners who completed the race while Lisa Hahner's deviation is greater than 86.5\%.
-Hence, the twins appear to have finished at a much slower pace than expectation, which is what we might expect if they were coordinating their runs.\footnote{It is important to note that the North Korean twins finished much faster than expected.
-XXXX explain PRK}
+One can see in Figure \ref{fig:scatter} that there is a clear
+relationship between a runner's personal best time and her performance
+in the Olympics.  While there is a significant amount of noise in the
+relationship, the Hahner twins' dots are well above the regression's
+fitted line notwithstanding this point.  They finished the Rio
+marathon much slower than they should have, per their pre-Olympic
+personal best times.  In fact, if we sort studentized residual, we can
+see in Figure \ref{fig:studentizedresiduals} that the Hahner twins are
+in the tail-end of the residual distribution.  Anna Hahner's residual
+deviation is greater than 91.7\% of the runners who completed the
+marathon while Lisa Hahner's deviation is greater than 86.5\%.  Hence,
+the twins appear to have finished at a much slower pace than expected,
+which is what we might expect if they were coordinating their runs.
+The North Korean Kim twins are also in the tail of the residual
+distribution---but the left tail.  These twins finished much faster
+than expected.
 
@@
 \begin{enumerate}
-\item Removing the twins from the sample, estimate a linear model that
-predicts a runner?s final time ($Y_i$) based on her personal best time ($X_i$).
-\item Extract the coefficients and the variance-covariance matrix from this
-model.
+\item Removing the two sets of twins and triplets from the sample,
+  estimate a linear model that predicts a runner's final time ($Y_i$)
+  based on her pre-Olympic personal best time ($X_i$).
+\item Extract the coefficients and the variance-covariance matrix from
+  this model.
 \item For each runner, draw $\beta_0$ and $\beta_1$ from a multivariate normal
@@
 mean equal to zero and a standard deviation equal to the standard
-deviation of the model?s residuals ($\sigma$).
+deviation of the model's residuals ($\sigma$).
 \item For each runner, predict the final result by combining the randomly
@@
 \begin{figure}[!ht]
- \caption{The distribution of simulated results}
+  \centering
+  \caption{Distribution of simulated results for the Hahner twins}
   \label{fig:simdiff}
- \begin{subfigure}{.5\textwidth}
- \includegraphics[width=\textwidth, keepaspectratio]{simulated_time.pdf}
+  \begin{subfigure}{.45\textwidth}
+    \includegraphics[width=\textwidth,
+    keepaspectratio]{simulated_time.pdf}
+    \caption{Simulated differences in finishing times}
+    \label{fig:simulatedfinishtimes}
   \end{subfigure}
- \begin{subfigure}{.5\textwidth}
+  \begin{subfigure}{.45\textwidth}
     \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
+    \caption{Simulated differences in places}
+    \label{fig:simulatedranks}
   \end{subfigure}

### 2017-01-29 seconds-behind.R

@@ -2,11 +2,12 @@ library(dplyr)
 
 dta <- read.csv (file = "data/times.csv", header = TRUE)
 
-pdf (file = "plots/seconds-behind.pdf")
+pdf (file = "plots/seconds-behind.pdf", width = 5, height = 5)
 
 plot (0,0, cex = 0, xlim = c(1, 10), ylim = c(0, 2000),
       xlab = "Split", ylab = "Seconds behind",
-      xaxt = "n")
+      xaxt = "n",
+      cex.axis = 0.75)
 
 for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     dta2use <- dta[!is.na(dta[,col2use]), ]
@@ -14,7 +15,7 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     ## set colors and shapes; have to do this for each split since the
     ## rows for the matrix being printed change per split when runners
     ## drop out
-    cex2use <- rep(0.12, times = nrow(dta))
+    cex2use <- rep(0.15, times = nrow(dta))
     cex2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- 1
     cex2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- 1
     cex2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- 1
@@ -37,12 +38,15 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
 }
 
 axis (side = 1, at = c(1:10),
-      labels = c("5K", "10K", "15K", "20K", "Half", "25K", "30K", "35K", "40K", "Final"))
+      labels = c("5K", "10K", "15K", "20K", "Half", "25K", "30K", "35K", "40K", "Final"),
+      cex.axis = 0.75)
 
 legend (x = 1, y = 2000,
         col = c("blue", "orange", "red", "grey50"),
         pch = 19,
-        legend = c("Estonian triplets", "German twins", "North Korean twins", "Others")
+        legend = c("Luik triplets", "Hahner twins", "Kim twins", "Others"),
+        cex = 0.75,
+        bty = "n"
         )
 
 
@@ -51,9 +55,9 @@ dta2use4lisa <- filter(dta, BIB == 749)
 
 
[…]

### 2017-02-01 paper/paper.tex
feb 1 changes

@@
 
+Anna and Lisa finished the marathon at an unusually slow rate, which is what we might expect if the twins intentionally slowed their pace to finish simultaneously.
+However, the observation of a slow finish does not necessarily imply that it was an intentional result .
+It only suggests that Kurschilgen's claim regarding this issue holds up in the data.
+
+Perhaps it would be easier to refute the claim if we could show that the Twins's back-to-back finish was not such an unusual event.
+Given that Anna and Lisa were so close in their personal best times, it may seem reasonable to expect that they would also be close in their finishing times.
+Hence, one way to determine if the twins's finish was intentional is to quantify the likelihood that such a close finish would occur by chance given the differences in their ability.
+
+We can attempt to quantify this likelihood by looking at how differences in the personal best time of two runners translate into differences in Olympic times .
+If Anna and Lisa's finish was unintentional we'd expect that such a close finish would be unlikely given their difference in personal best times.
+Yet, how does the difference in personal best times relate to the difference in finishing times?
+
+One way we can estimate this relationship is to look at all of the differences that we observe in the marathon.
+We can take the difference in personal best time for every combination of runners in the data and look at the difference in their final times.
+We should expect that compared to the difference in their personal best times, the difference in Anna and Lisa's final times should be commensurate with the conditional difference of all other runners.
+%This needs a better explanation.
+
+Therefore, we find all combinations of runners in the marathon and calculate two quantities: 1) the difference in their personal best time and 2) the difference in their final time.
+We plot the relationship between the two quantities in Figure \ref{fig:diffdiffscatter}.
+All 8,778 dyadic combinations of runners are displayed by grey points in the scatter plot.
+The solid black line indicates a one-to-one relationship, where every point would fall if the difference in final times were systematically equivalent to the difference in personal bests.
+
 \begin{figure}[!ht]
  \caption{Relationship between Difference in Personal Best and Difference in Result}
+ \label{fig:diffdiffscatter}
+ \centering
+ \begin{subfigure}{.5\textwidth}
+ \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
+ \end{subfigure}
+\end{figure}
+
+We can see from the yellow dot that the Hahner Twins are just below this one-to-one line.
+Hence, the twins' final times were closer to each other than their personal best times; but not by much.
+In fact, this might be an expected difference if we expected there to be a one-to-one relationship.
+However, although there is a clear relationship between the difference in final time and the difference in personal best, it is not a strong one-to-one relationship.
+We've added a dotted smoother to display a cubic splines estimate of the conditional averages.
+While the Hahner twins may be close to the one-to-one line, they are much further away from the average set of runners.
+Hence, we would expect for any pair of runners, the difference in their final results should be much greater than their difference in their personal best.
+However, the Hahner twins do not fit this expectation.
+Yet, how different from this expectation are they?
+
+One way we can quantify the degree to which the twins are outliers is by comparing their difference in difference to every other pair of runner's difference-in-differences.
+We calculate the difference-in-differences for each pair of runners by simply subtracting their difference in result (y-axis) from their difference in personal best (x-axis).
+Therefore positive difference-in-differences mean that the runners ended up further apart than their difference in personal bests.
+Negative difference-in-differences mean that the runners end up closer together.
+Hence, the Hahner twins have a difference-in-difference of -114 seconds.
+They finished 114 seconds closer together than the distance between their personal bests.
+
+We arrange all 8,778 difference-in-differences from small to large in the first plot in Figure \ref{fig:diffdiff}.
+The Hahner twins are denoted by the yellow point.
+Using this plot we can visualize where the twins fall within the distribution of difference-in-differences.
+It is clear that they are closer to the lower tail of the distribution.
+However, compared to all other difference-in-differences, their result is not that extreme.
+Nearly 21\% of the difference-in-dfferences were less than the that of the twins'.
+Hence, we'd expect at least one in five pairs of runners to have a reduction in difference at least as great as the twins.
+
+However, if we limit the observations to only those runners who had personal best times that were at least as close to each other as the Hahner twins, we get a different persepective.
+Among those runners,  the Hahner twins reduced the distance between them to the greatest extent.
+There was no other pair of runners whose reduced the difference in the final time relative to their difference in personal bests more than them.
+
+\begin{figure}[!ht]
+ \caption{How Hahner Twins Rank in Difference-in-Differences}
  \label{fig:diffdiff}

### 2017-02-01 2-simulate.R

@@ -11,7 +11,7 @@ pdf("plots/scatter_plot.pdf", height = 5,  width = 5)
 ggplot(aes(y=FINAL, x=PB), data = df) + 
   geom_point(size = 1, colour = "gray") + 
   geom_point(data = df[!is.na(df$TWINS),], aes( col = TWINS)) +
-  geom_smooth(method = "lm", colour = "black", alpha = .5, size = .5, linetype = "dashed") + 
+  geom_smooth(method = "lm", colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black") + 
   geom_abline(intercept = 0, slope =1) +
   geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
   geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
@@ -48,7 +48,7 @@ for(i in 1:length(pb)){
 }
 
 
-df2 <- data_frame(name_i, name_j, pb_diff, result_diff, diff_in_diff = abs(result_diff - pb_diff), consecutive)
+df2 <- data_frame(name_i, name_j, pb_diff, result_diff, diff_in_diff = abs(result_diff) - abs(pb_diff), consecutive)
 
 df2$twins <- ifelse(df2$name_i == "Anna Hahner" & df2$name_j == "Lisa Hahner", "Hahner Twins",
                     ifelse(df2$name_i == "Hye-Gyong Kim" & df2$name_j == "Hye-Song Kim", "Kim Twins",
@@ -57,14 +57,32 @@ df2$twins <- ifelse(df2$name_i == "Anna Hahner" & df2$name_j == "Lisa Hahner", "
 
 df2b <- df2 %>% filter(name_i != name_j, !is.na(diff_in_diff)) %>% arrange(diff_in_diff)
 
+pdf("plots/diff_in_diff_scatter_plot.pdf", height = 5,  width = 5)
+ggplot(aes(y=abs(result_diff), x=abs(pb_diff)), data = df2b) + 
+  geom_point(size = .5, colour = "gray", alpha = .5) + 
+  geom_point(data = df2b[!is.na(df2b$twins),], aes( col = twins)) +
+  geom_smooth( colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black") + 
+  geom_abline(intercept = 0, slope =1) +
+  #geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
[…]

### 2017-02-01 paper/paper.tex
a few changes

@@
 marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
-80 countries across the world took off from the starting line en route
-to their destination 42.195 kilometers away. Two hours, twenty-four
-minutes, and four seconds later, Jemima Sumgong of Kenya would be the
-first to cross the finishline and take home gold; Sumgong was just 3.5
-minutes behind her personal best time in the marathon. Approximately
-21 minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
+80 countries across the world left the starting line en route to their
+destination 42.195 kilometers away. Two hours, twenty-four minutes,
+and four seconds later, Jemima Sumgong of Kenya would be the first to
+cross the finishline and take home gold; Sumgong was just 3.5 minutes
+behind her personal best time in the marathon. Approximately 21
+minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
 would cross the finishline together, holding hands and celebrating a
@@
 Kurschilgen averred that the twins slowed down so as to finish
-simultaneously and sought to create a spectacle and ``generate media
+simultaneously and create a spectacle which would ``generate media
 attention.'' Kurschilgen justified his charge with the fact that the
-twins were at least 18 minutes behind their personal best
-times.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
-Not surprisingly, these accusations were denied by the Hahners, who
-claimed that their simultaneous finish was simply an unintended
+twins ran in the Rio marathon at least 18 minutes slower their
+personal best times prior to the
+Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
+Not surprisingly, these accusations were denied by the Hahner twins,
+who claimed that their simultaneous finish was simply an unintended
 coincidence.
@@
 
-What happened in the women's Olympic marathon, and how might we assess
-whether the Hahner finish was coincidental or intentional? These two
-interpretations are clearly at odds. If the former, then the Hahner
-twins are to be celebrated and their finish treated as an expression
-of the spirit behind the Olympic games. If the latter, though, then
-the Hahner twins may have violated this spirt by not trying hard
-enough. It is perhaps too easy for us to write such a glib
-sentence---neither of us cannot fathom being able to complete a
-marathon anywhere in the vicinity of two and a half hours---but we
-nonetheless want to know what the data from the Olympic marathon data
-tell us. Was the women's Olympic marathon a lovely coincidence or
-something else?
+What happened in the women's Olympic marathon, and how might we
+develop a statistical approach that assesses whether the Hahner finish
+was coincidental or intentional?  These two interpretations are
+clearly at odds. If the former, then the Hahner twins are to be
+celebrated and their finish treated as an expression of the spirit
+behind the Olympic games. If the latter, though, then the twins may
+have violated this spirt by not trying hard enough. It is perhaps too
+easy for us to write such a glib sentence---neither of us cannot
+fathom being able to complete a marathon anywhere in the vicinity of
+two and a half hours---but we nonetheless want to know what the data
+from the Olympic marathon data tell us.  Was the Hahner finish in the
+2016 women's Olympic marathon a lovely coincidence or something else?
 
@@
 The marathon also featured twins from North Korea, Kim Hye-song and
-Kim Hye-gyong, who had identical times and finished 10th and 11th,
-respectively. The Kim finish, unlike the Hahner finish, appears devoid
-of post-race controversy. Moreover, three triplets from Estonia
-competed in the Rio marathon, although only two, Lily Luik and Leila
-Luik, finished it, in 97th and 114th place, respectively. The third
-Estonia triplet, Liina Luik, recorded what is known as a DNF---an
-abbreviation that means did not finish, a term that we will use
-throughout this article.
+Kim Hye-gyong, who posted identical times and finished 10th and 11th
+in the race, respectively. The Kim finish, unlike the Hahner finish,
+appears devoid of post-race controversy. Moreover, three triplets from
+Estonia competed in the Rio marathon, although only two, Lily Luik and
+Leila Luik, finished it, in 97th and 114th place, respectively. The
+third Estonia triplet, Liina Luik, recorded what is known as a
+DNF---an abbreviation that means did not finish, a term that we will
+use throughout this article.
 
-\section*{Marathon data}
+\section*{Marathon data and our research design}
 
@@
 observe the effort that an individual put into the race, and we do not
-know why some runners have DNF results; some runners may have injured
+know why some runners have DNF results.  Some runners may have injured
 themselves on the course and accordingly dropped out, and others may
@@
 was thus $\frac{23}{156} \approx 0.15$, and the relatively small
-sample size at our disposal---156 runners---means that a 95\%
-confidence for this rate is fairly wide, namely,
-$\left(0.098, 0.22\right)$.
+sample size at our disposal means that a 95\% confidence for this rate
+is fairly wide, namely, $\left(0.098, 0.22\right)$.  Key in the
+analysis that follows is the deviation between a runner's Rio time and
+her prior personal best time in the marathon.  While there is variance
+across runners in these deviations, they nonetheless may be able to
+provide us with insight as to the Hahner twins' performance in the
+marathon.
 
-Key in the analysis that follows is the deviation between a runner's
-Rio time and her prior personal best time in the marathon. Of course
-there is variance across runners in these deviations, and we want to
-know if patterns in the deviations support an honest effort by the
-Hanher twins or perhaps something else.
+Our research design has two components.  In the first, we present some
+visualizations that present the Hahner twin's results in context; our
+interest here is primarily the two Hanhers although we comment on the
+Kim twins and Luik triplets when appropriate.  Our visualizations
+suggest blah blah bhah.  We then turn to a modeling approach and in
+particular a simulation.  In simulated Olympic marathons, which we
+develop based on relationships (excluding twins and triplets) between
+known personal best times and observed Rio finishing times, we find
+that the Hahner twins finished suspiciously close to each other given
+the disparty between their personal best times.  OR THEY DID NOT.
+
+\section*{Visualizing the Olympic marathon}
+
+SOME PLOTS HERE.  Maybe PB versus FINAL and HALF vs Final.
 
@@
 kilometer split, but the third triplet did not record a half-marathon
-split.  The orange dots in the figure represent the Hahner twins; Lisa
-Hahner was ahead of her sister through 15 kilometers at which point
-Anna surged ahead.  There are two red dots representing the North
-Korean Kim twins although this is not visually apparent; the Kim twins
+split.  The orange dots in the figure represent the Hahner twins; as
+noted in the figure, Lisa Hahner was ahead of her sister through 15
+kilometers at which point Anna surged ahead.  Lastly, Figure
+\ref{fig:secondsbehind} contains two red dots representing the North
+Korean Kim twins; this is not visually apparent because the Kim twins
 had identical split times during the marathon.

### 2017-02-02 2-simulate.R
change model to quadratic

@@ -7,11 +7,14 @@ df <- read.csv("data/times.csv", stringsAsFactors = FALSE) %>% tbl_df()
 # Remove Irvette Van Zyl, since she did not start the race due to injury.  Her bib number is 1172.
 df <- df[df$BIB != 1172,]
 
+# Remove slowest PBs with the assumption that these are different
+# df <- df[df$PB < 10000,]
+
 pdf("plots/scatter_plot.pdf", height = 5,  width = 5)
 ggplot(aes(y=FINAL, x=PB), data = df) + 
   geom_point(size = 1, colour = "gray") + 
   geom_point(data = df[!is.na(df$TWINS),], aes( col = TWINS)) +
-  geom_smooth(method = "lm", colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black") + 
+  geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black") + 
   geom_abline(intercept = 0, slope =1) +
   geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
   geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
@@ -107,11 +110,13 @@ ggplot(df2c, aes(PERCENTILE,diff_in_diff)) +
 dev.off()
 
 # Plot Studentized Residuals
-pdf("plots/studentized_residuals.pdf", height = 5, width = 5)
-fit <- lm(FINAL ~ PB + PB^2, data = df)
+
+fit <- lm(FINAL ~ PB + I(PB^2), data = df)
 dta <- data.frame( df[!is.na(df$FINAL), c("ATHLETE", "TWINS")], STUD_RESID =  rstudent(fit) )
 dta <- arrange(dta, STUD_RESID)
 dta <- dta %>% mutate(PERCENTILE = cume_dist(STUD_RESID)*100 )
+
+pdf("plots/studentized_residuals.pdf", height = 5, width = 5)
 ggplot(dta, aes(PERCENTILE,STUD_RESID)) + 
   geom_point(size = 1, colour = "gray") + 
   geom_point(data = na.omit(dta), aes( col = TWINS)) + 
@@ -130,7 +135,7 @@ dev.off()
 dnf_rate <- sum(is.na(df$FINAL))/length(df$FINAL)
 
[…]

### 2017-02-02 2-simulate.R
minor R updates

@@ -4,7 +4,8 @@ library(lubridate)
 library(stringr)
 df <- read.csv("data/times.csv", stringsAsFactors = FALSE) %>% tbl_df()
 
-# Remove Irvette Van Zyl, since she did not start the race due to injury.  Her bib number is 1172.
+# Remove Irvette Van Zyl, since she did not start the race due to
+# injury.  Her bib number is 1172.
 df <- df[df$BIB != 1172,]
 
 # Remove slowest PBs with the assumption that these are different
@@ -19,13 +20,34 @@ ggplot(aes(y=FINAL, x=PB), data = df) +
   geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
   geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
   coord_fixed(xlim = c(8000,12500), ylim = c(8000,12500)) +
-  ylab("Olympic result in seconds\n") + 
-  xlab("\nPersonal Best in seconds") + 
+  ylab("Olympic result, in seconds\n") + 
+  xlab("\nPersonal Best, in seconds") + 
     ##  ggtitle("Distribution of Y | X") +
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +
   theme(legend.position = c(.8,.2))
 dev.off()
+
+## make scatter plot for half marathon split
+pdf("plots/scatter_plot_half.pdf", height = 5,  width = 5)
+df2use <- filter(df, !is.na(SPLIT_HALF))
+ggplot(aes(y = FINAL, x = SPLIT_HALF), data = df2use) +
+    geom_point(size = 1, colour = "gray") + 
+    geom_point(data = df2use[!is.na(df2use$TWINS),], aes( col = TWINS)) +
+    geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black", se = FALSE) + 
+    ## geom_abline(intercept = 0, slope = 1) +
+    geom_rug(data = df2use[is.na(df2use$FINAL),], aes(x = SPLIT_HALF), colour = "gray") +
+    geom_rug(data = df2use[is.na(df2use$FINAL) & df2use$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
[…]

### 2017-02-02 2-simulate.R
minor R updates

@@ -15,7 +15,7 @@ pdf("plots/scatter_plot.pdf", height = 5,  width = 5)
 ggplot(aes(y=FINAL, x=PB), data = df) + 
   geom_point(size = 1, colour = "gray") + 
   geom_point(data = df[!is.na(df$TWINS),], aes( col = TWINS)) +
-  geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black") + 
+  geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black", se = FALSE) + 
   geom_abline(intercept = 0, slope =1) +
   geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
   geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +

### 2017-02-02 2-simulate.R
minor R updates

@@ -21,7 +21,7 @@ ggplot(aes(y=FINAL, x=PB), data = df) +
   geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
   coord_fixed(xlim = c(8000,12500), ylim = c(8000,12500)) +
   ylab("Olympic result, in seconds\n") + 
-  xlab("\nPersonal Best, in seconds") + 
+  xlab("\nPersonal nest, in seconds") + 
     ##  ggtitle("Distribution of Y | X") +
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +

### 2017-02-02 2-simulate.R
minor R updates

@@ -11,28 +11,35 @@ df <- df[df$BIB != 1172,]
 # Remove slowest PBs with the assumption that these are different
 # df <- df[df$PB < 10000,]
 
+## add indicator for finishing at least 19 minutes slower than PB
+df <- df %>% mutate(slower19 = (FINAL - PB) / 60 >= 19)
+
 pdf("plots/scatter_plot.pdf", height = 5,  width = 5)
 ggplot(aes(y=FINAL, x=PB), data = df) + 
-  geom_point(size = 1, colour = "gray") + 
-  geom_point(data = df[!is.na(df$TWINS),], aes( col = TWINS)) +
-  geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black", se = FALSE) + 
-  geom_abline(intercept = 0, slope =1) +
-  geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
-  geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
-  coord_fixed(xlim = c(8000,12500), ylim = c(8000,12500)) +
-  ylab("Olympic result, in seconds\n") + 
-  xlab("\nPersonal nest, in seconds") + 
+    geom_point(size = 1.25, colour = "gray", aes(shape = slower19)) +
+    geom_point(data = df[!is.na(df$TWINS),], aes(col = TWINS)) +
+    geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black", se = FALSE) + 
+    geom_abline(intercept = 0, slope =1) +
+    geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
+    geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
+    coord_fixed(xlim = c(8000,12500), ylim = c(8000, 12500)) +
+    ylab("Olympic result, in seconds\n") + 
+    xlab("\nPersonal best, in seconds") + 
     ##  ggtitle("Distribution of Y | X") +
-  scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
-  theme_bw() +
-  theme(legend.position = c(.8,.2))
[…]

### 2017-02-02 paper/paper.tex
paper updates

@@
 cross the finishline and take home gold; Sumgong was just 3.5 minutes
-behind her personal best time in the marathon. Approximately 21
+behind her prior personal best time in the marathon. Approximately 21
 minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
@@
 Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
-Not surprisingly, these accusations were denied by the Hahner twins,
-who claimed that their simultaneous finish was simply an unintended
-coincidence.
+Not surprisingly, Kurschilgen's accusations were denied by the Hahner
+twins, who claimed that their simultaneous finish was simply an
+unintended coincidence.
 
@@
 What happened in the women's Olympic marathon, and how might we
-develop a statistical approach that assesses whether the Hahner finish
-was coincidental or intentional?  These two interpretations are
-clearly at odds. If the former, then the Hahner twins are to be
-celebrated and their finish treated as an expression of the spirit
-behind the Olympic games. If the latter, though, then the twins may
-have violated this spirt by not trying hard enough. It is perhaps too
-easy for us to write such a glib sentence---neither of us cannot
-fathom being able to complete a marathon anywhere in the vicinity of
-two and a half hours---but we nonetheless want to know what the data
-from the Olympic marathon data tell us.  Was the Hahner finish in the
-2016 women's Olympic marathon a lovely coincidence or something else?
+develop a statistical approach that assesses whether the Hahner twin's
+finish was coincidental or intentional?  These two interpretations are
+clearly at odds. If the former, then the Hahners are to be celebrated
+and their finish treated as an expression of the spirit behind the
+Olympic games. If the latter, though, then the twins may have violated
+this spirt by not trying hard enough. It is perhaps too easy for us to
+write such a glib sentence---neither of us can fathom being able to
+complete a marathon anywhere in the vicinity of two and a half
+hours---but we nonetheless want to know what the data from the Olympic
+marathon data tell us.  Was the Hahner finish in the 2016 women's
+Olympic marathon a lovely coincidence or something else?
 
 Among female Olympic marathoners, the Hahner twins were not alone in
-their familial ties, and we draw on this in the analysis that follows.
-The marathon also featured twins from North Korea, Kim Hye-song and
-Kim Hye-gyong, who posted identical times and finished 10th and 11th
-in the race, respectively. The Kim finish, unlike the Hahner finish,
-appears devoid of post-race controversy. Moreover, three triplets from
-Estonia competed in the Rio marathon, although only two, Lily Luik and
-Leila Luik, finished it, in 97th and 114th place, respectively. The
-third Estonia triplet, Liina Luik, recorded what is known as a
-DNF---an abbreviation that means did not finish, a term that we will
-use throughout this article.
+their familial ties, and we touch on this in the analysis that
+follows.  The marathon also featured twins from North Korea, Kim
+Hye-song and Kim Hye-gyong, who posted identical times and finished
+10th and 11th in the race, respectively. The Kim finish, unlike the
+Hahner finish, appears devoid of post-race controversy. Moreover,
+three triplets from Estonia competed in the Rio marathon, although
+only two, Lily Luik and Leila Luik, finished it, in 97th and 114th
+place, respectively. The third Estonia triplet, Liina Luik, recorded
+what is known as a DNF---an abbreviation that means did not finish, a
+term that we will use throughout this article.
 
@@
 sample size at our disposal means that a 95\% confidence for this rate
-is fairly wide, namely, $\left(0.098, 0.22\right)$.  Key in the
-analysis that follows is the deviation between a runner's Rio time and
-her prior personal best time in the marathon.  While there is variance
-across runners in these deviations, they nonetheless may be able to
-provide us with insight as to the Hahner twins' performance in the
-marathon.
+is fairly wide, namely, $\left(0.098, 0.22\right)$.
 
-Our research design has two components.  In the first, we present some
-visualizations that present the Hahner twin's results in context; our
-interest here is primarily the two Hanhers although we comment on the
-Kim twins and Luik triplets when appropriate.  Our visualizations
-suggest blah blah bhah.  We then turn to a modeling approach and in
-particular a simulation.  In simulated Olympic marathons, which we
-develop based on relationships (excluding twins and triplets) between
-known personal best times and observed Rio finishing times, we find
-that the Hahner twins finished suspiciously close to each other given
-the disparty between their personal best times.  OR THEY DID NOT.
+The Kurschilgen accusation against the Hahner twins has two
+components, that the twins finished simultaneously, holding hands and
+smiling, \emph{and} that they ran slowly.  We suspect that Kurschilgen
+would not have expressed ire at the Hahners had they finished in 1st
+and 2nd place in Rio, hand-in-hand with wide grins, but of course we
+do not know this.  As such, our investigation of the charges that
+Kurschilgen offered will distinguish between the idea of slow finish
+versus a simultaneous finish.
+
+Our research design has two components.  In the first, we present
+visualizations that describe various features of the Hahner twin's
+results, and an important element of our visualizations is the
+difference between a runner's Rio time and her prior personal best
+time in the marathon.  Overall, our visualizations suggest blah blah
+bhah.  We then turn to modeling and in particular a regression-based
+simulation.  In simulated Olympic marathons, which we develop based on
+relationships (excluding the Hahner and Kim twins and Luik triplets)
+between known personal best times and observed Rio finishing times, we
+find that the Hahner twins finished suspiciously close to each other
+given the disparty between their personal best times.  OR THEY DID
+NOT.  Finally, we return to Kurschilgen claims  about the Hahners and
+offer our thoughts about their validity.
 
@@
 
-SOME PLOTS HERE.  Maybe PB versus FINAL and HALF vs Final.
+Figure \ref{fig:scatter} contains two plots, both of which describe
+how the Rio Olympic women's marathon results varied as a function of
+athletes' personal best times and their half marathon split times.
+The points in each plot are colored by twin/triplet status, and both
+plots contain second-order polynomial regression lines.  We treat an
+athlete's prior personal best marathon time as a measure of the
+athlete's underlying marathon talent, and we interpret an athlete's
+half marathon split as a measure of the athlete's skill at marathon
+running on August 14, 2016, the date of the women's marathon in Rio.
+
+\begin{figure}[!ht]
+  \caption{Olympic results and measures of marathon ability}
+  \label{fig:scatter}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
+    \caption{Finishing and personal best times}
+    \label{fig:45degreeplot}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
+    \caption{Finishing and half marathon split times}
+    \label{fig:45degreeplot_half}
+  \end{subfigure}
+\end{figure}
+
+Considering first Figure \ref{fig:45degreeplot_half}, note that this
+figure includes a solid 45-degree line as the horizontal and vertical
+axes in the figure correspond to full marathons.  The rug marks on the
+bottom of the figure denote the personal best times of runners who
+DNFed.  Given the paucity of points (five of them) below the 45-degree
+line, the vast majority of Olympic marathoners ran slower in Rio
+compared to their personal bests.  Relative to personal best times,
+the Hahner twins were definitely on the slow side, but a number of
+runners had greater differences between their Olympic times and their
+personal bests than the Hahners.  These runners are denoted with
+squares in Figure \ref{fig:45degreeplot}, and there are 13 such
+symbols, highlighting approximately 10\% of the racers who completed
+the marathon.  The figure shows moreover that there were two women who
+had personal best times slightly faster than the Hahner's and yet
+finished after both of the German women.  Although Figure
+\ref{fig:45degreeplot} suggests that the Hahner twins were slower than
+one would have expected given their previous best marathon times, it
+is not consistent with the accusation that they dramatically slowed
+down in the Rio marathon.
+
+% > df <- read.csv("data/times.csv", stringsAsFactors = FALSE) %>% tbl_df()
+% > df <- df[df$BIB != 1172,]
+% > sort(dta$FINAL - dta$PB) / 60
+%   [1] -2.3333333 -2.2166667 -2.0666667 -1.1166667 -0.4000000  0.2166667
+%   [7]  0.6333333  0.9000000  1.0166667  1.3666667  1.5166667  1.7166667
+%  [13]  1.7666667  1.8500000  2.1166667  2.4666667  2.5333333  3.0166667
+%  [19]  3.0500000  3.1666667  3.2666667  3.2833333  3.3666667  3.4500000
+%  [25]  3.6833333  3.9000000  3.9833333  4.2000000  4.2666667  4.3833333
+%  [31]  4.4666667  4.5833333  4.6166667  4.6333333  4.6833333  4.7333333
+%  [37]  4.7333333  4.8333333  5.0666667  5.0833333  5.0833333  5.1000000
+%  [43]  5.1333333  5.3500000  5.3500000  5.4000000  5.7166667  5.7333333
+%  [49]  5.9333333  6.0000000  6.0333333  6.1166667  6.3333333  6.3500000
+%  [55]  6.5000000  6.5333333  6.6000000  6.6666667  6.7166667  6.7833333
+%  [61]  6.8166667  6.9500000  7.1000000  7.1500000  7.2000000  7.3500000
+%  [67]  7.4500000  7.5666667  7.6000000  7.6000000  7.6333333  7.7500000
+%  [73]  7.8833333  7.9666667  7.9833333  8.3000000  8.6000000  9.3000000
+%  [79]  9.5500000  9.5833333  9.6000000  9.7166667  9.9500000  9.9833333
+%  [85] 10.3166667 10.5333333 10.6666667 10.7000000 10.8333333 10.9666667
+%  [91] 11.1500000 11.1833333 11.4666667 11.7166667 12.1666667 12.6000000
+%  [97] 12.9166667 13.2166667 13.2833333 13.3833333 13.7666667 13.8666667
+% [103] 13.9166667 14.0500000 14.0833333 14.5666667 14.6833333 14.7500000
+% [109] 15.0000000 15.0500000 15.5000000 15.9000000 16.2000000 16.9000000
+% [115] 17.4500000 18.1000000 18.6666667 18.7166667 18.8000000 18.9333333
+% [121] 20.1500000 20.6333333 20.7000000 20.8500000 21.3500000 23.4166667
+% [127] 24.9333333 25.6500000 26.5666667 26.7666667 29.2166667 30.7000000
+% [133] 30.9666667
+% >
+
+One limitation of using personal best times as indicators of
+underlying athletic talent is that these times are potentially
+confounded by the marathon courses at which they were set (some*sh
+courses, like the Berlin marathon, are known for fast times) and race
+conditions like weather.  In addition, personal best times may not
+capture raceday idiosyncrasies that might affect individual runners.
+With that in mind, Figure \ref{fig:45degreeplot_half} plots marathon
+finishing times against half marathon splots.  This figure has a
+regression line as before but no 45-degree line.
+
+Figure \ref{fig:45degreeplot_half} shows that there was nothing
+abnormal about the Hahner twins' overall finishing times, conditional
+on their half marathon split times.  This is consistent with the
+previous Figure \ref{fig:45degreeplot} and inconsistent with the
+accusations against the Hahner twins, at least the part of the
+accusation that focuses on their excessively slowing down.
+
+If the Hahner twins did not slow down excessively, might they have
+run somewhat strategically at the end of the race in order to generate
+a simultaneous finish?  We now offer a visualization that speaks to
+this question.
+
+The personal best times of the Hahner twins were 115 seconds apart and
+their official finishing times were separated by one second.  Is such
+a 115 to one compression typical among pairs of runners or was it
+unusual?  We might ask, are there other pairs of runners who had a
+difference between personal best times of 115 seconds apart and, if
+so, how close where their finishing times?
+
+Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
+pairs of runners.  Of these and ignoring the Hahner twins, ten had
+exactly a 115 second gap in personal best times.  Of these ten pairs,
+differences in finishing times in seconds are as follows: 36, 93, 172,
+319, 379, 459, 552, 671, 675, 739.  In other words, of all pairs of
+runners in the Rio marathon who had a personal best difference that
+was equivalent to the Hahner twins' difference, the twins had the
+greatest compression based on finishing time.
+
+% df2b[abs(df2b$pb_diff)==115,]
+% # A tibble: 11 × 8
+%                       name_i              name_j pb_diff result_diff
+%                        <chr>               <chr>   <int>       <int>
+% 1                Anna Hahner         Lisa Hahner    -115          -1
+% 2                  Lily Luik Mariya Korobitskaya     115          36
+% 3        Lisa Jane Weightman      Lanni Marchant    -115          93
+% 4           Irina Smolnikova      Viviana Chávez     115        -172
+% 5              Sonia Samuels    Helalia Johannes     115        -319
+% 6  Jessica Draskau-Petersson       Gladys Tejeda     115         379
+% 7        Maryna Damantsevich       Gladys Tejeda     115         459
+% 8          Panayióta Vlaháki     Matea Matosevic    -115         552
+% 9         Nataliya Lehonkova       Olha Kotovska     115         671
+% 10               Anja Scherl      Desireé Linden     115         675
+% 11              Rose Chelimo    Helalia Johannes    -115        -739
+% # ... with 4 more variables: diff_in_diff <int>, consecutive <lgl>,
+% #   twins <chr>, PERCENTILE <dbl>
+% >
+
+\begin{figure}[!ht]
+ \caption{Differences in personal best times and finishing times}
+ \label{fig:diffdiffscatter}
+ \centering
+ \begin{subfigure}{.5\textwidth}
+ \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
+ \end{subfigure}
+\end{figure}
+
+We can generalize this result by looking at all pairs of runners in
+the marathon.  For the 8,778 pairs of 133 finishers, Figure
+\ref{fig:diffdiff} plots differences in finishing times against
+differences in personal best times, and pairs of twins/triplets are
+identified by the same color scheme we have used earlier.
+
+Consider first the Hahner twins.  They are effectively located on the
+horizontal axis (because their difference in finishing times is one
+second).  However, there are many points about the Hahner's orange
+dot, and this shows that, conditional on a 115 second difference in
+personal best times,
+
+In the
+context of the marathon, that is, did Anna and Lisa Hahner run with an
+eye on each other?
+
+We can see from the yellow dot that the Hahner Twins are just below this one-to-one line.
+Hence, the twins' final times were closer to each other than their personal best times; but not by much.
+In fact, this might be an expected difference if we expected there to be a one-to-one relationship.
+However, although there is a clear relationship between the difference in final time and the difference in personal best, it is not a strong one-to-one relationship.
+We've added a dotted smoother to display a cubic splines estimate of the conditional averages.
+While the Hahner twins may be close to the one-to-one line, they are much further away from the average set of runners.
+Hence, we would expect for any pair of runners, the difference in their final results should be much greater than their difference in their personal best.
+However, the Hahner twins do not fit this expectation.
+Yet, how different from this expectation are they?
+
+One way we can quantify the degree to which the twins are outliers is by comparing their difference in difference to every other pair of runner's difference-in-differences.
+We calculate the difference-in-differences for each pair of runners by simply subtracting their difference in result (y-axis) from their difference in personal best (x-axis).
+Therefore positive difference-in-differences mean that the runners ended up further apart than their difference in personal bests.
+Negative difference-in-differences mean that the runners end up closer together.
+Hence, the Hahner twins have a difference-in-difference of -114 seconds.
+They finished 114 seconds closer together than the distance between their personal bests.
+
+qq
 
@@
     \centering
-    \includegraphics[width=\textwidth,
-    keepaspectratio]{scatter_plot.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
     \caption{Rio finishing times and personal best times}
@@
     \centering
-    % \label{fig:Studentized Residuals}
     \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
@@
 
-\begin{figure}[!ht]
- \caption{Relationship between Difference in Personal Best and Difference in Result}
- \label{fig:diffdiffscatter}
- \centering
- \begin{subfigure}{.5\textwidth}
- \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
- \end{subfigure}
-\end{figure}
+% \begin{figure}[!ht]
+%  \caption{Relationship between Difference in Personal Best and Difference in Result}
+%  \label{fig:diffdiffscatter}
+%  \centering
+%  \begin{subfigure}{.5\textwidth}
+%  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
+%  \end{subfigure}
+% \end{figure}
 

### 2017-02-02 2-simulate.R

@@ -98,8 +98,8 @@ ggplot(aes(y=abs(result_diff), x=abs(pb_diff)), data = df2b) +
   #geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
   #geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
   coord_fixed(xlim = c(0,3500), ylim = c(0,3500)) +
-  ylab("Absolute Difference Olympic result in seconds\n") + 
-  xlab("\nAbsolute Difference in Personal Best in seconds") + 
+  ylab("Absolute difference Olympic result, in seconds\n") + 
+  xlab("\nAbsolute difference in personal best, in seconds") + 
   #ggtitle("Among all dyadic combinations") +
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +

### 2017-02-02 paper/paper.tex
paper updates

@@
 finish as a ``magical moment.''
-\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
+
 The media quickly picked up on the Hanher story as an image of the
@@
 
+%% Honigkuchenpferde reference
+%%
+%% https://www.welt.de/sport/olympia/article157669264/Das-falsche-Laecheln-der-deutschen-Lauf-Zwillinge.html
+
 Not everyone agreed with this rosy interpretation. The twins' happy
-facial expressions at the finish were portrayed as a bit contrived
-(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
-cookie in the shape of a horse), and the sports director of the German
-Athletics Federation, Thomas Kurschilgen, stirred up controversy when
-he suggested that the Hahners' photo-finish was no coincidence.
+facial expressions at the finish were portrayed as a bit
+contrived---smiling like ``Honigkuchenpferde,'' cookies in the shape
+of a horse---and the sports director of the German Athletics
+Federation, Thomas Kurschilgen, stirred up controversy when he
+suggested that the Hahners' photo-finish was no coincidence.
 Kurschilgen averred that the twins slowed down so as to finish
@@
 differences in finishing times in seconds are as follows: 36, 93, 172,
-319, 379, 459, 552, 671, 675, 739.  In other words, of all pairs of
-runners in the Rio marathon who had a personal best difference that
+319, 379, 459, 552, 671, 675, and 739.  In other words, of all pairs
+of runners in the Rio marathon who had a personal best difference that
 was equivalent to the Hahner twins' difference, the twins had the
@@
 differences in personal best times, and pairs of twins/triplets are
-identified by the same color scheme we have used earlier.
+identified by the same color scheme we used earlier.
 
 Consider first the Hahner twins.  They are effectively located on the
-horizontal axis (because their difference in finishing times is one
-second).  However, there are many points about the Hahner's orange
-dot, and this shows that, conditional on a 115 second difference in
-personal best times,
+horizontal axis because their difference in finishing times is one
+second.  However, there are many points about the Hahner's orange dot,
+and this shows that, conditional on a 115 second difference in
+personal best times, most marathoners did not have close finishing
+times like the Hahnes.  Some pairs of runners with 115 second personal
+best differences had finishing time differences of 1000 seconds, i.e.,
+in excess of 15 minutes.  Of course the points in Figure
+\ref{fig:diffdiffscatter} are not independent, but they provide a
+sense of the dispersion in finishing time differences that one can
+expect conditional on differences in personal best times.
 
-In the
-context of the marathon, that is, did Anna and Lisa Hahner run with an
-eye on each other?
+Thinking about the accusations leveled against the Hahner twins,
+Figure \ref{fig:diffdiffscatter} suggests that Anna and Lisa Hahner
+did indeed run with an eye on each other.  In fact, the same can be
+said of the Kim twins, who ran seemingly in lockstep throughout the
+entire Rio marathon.  The North Korean twins had a personal best
+difference of 53 seconds and a finishing time difference of literally
+zero seconds.  Beyond these twins, there were eight pairs of runners
+with a 53 second personal best difference, and resulting finishing
+time differences are as follows: 9, 51, 228, 340, 352, 571, 662, and
+751.  As in the Hahner case, the Kim twins compressed their finishing
+times more than any other pair of runners with a similar personal best
+difference.
 
-We can see from the yellow dot that the Hahner Twins are just below this one-to-one line.
-Hence, the twins' final times were closer to each other than their personal best times; but not by much.
-In fact, this might be an expected difference if we expected there to be a one-to-one relationship.
-However, although there is a clear relationship between the difference in final time and the difference in personal best, it is not a strong one-to-one relationship.
-We've added a dotted smoother to display a cubic splines estimate of the conditional averages.
-While the Hahner twins may be close to the one-to-one line, they are much further away from the average set of runners.
-Hence, we would expect for any pair of runners, the difference in their final results should be much greater than their difference in their personal best.
-However, the Hahner twins do not fit this expectation.
-Yet, how different from this expectation are they?
-
-One way we can quantify the degree to which the twins are outliers is by comparing their difference in difference to every other pair of runner's difference-in-differences.
-We calculate the difference-in-differences for each pair of runners by simply subtracting their difference in result (y-axis) from their difference in personal best (x-axis).
-Therefore positive difference-in-differences mean that the runners ended up further apart than their difference in personal bests.
-Negative difference-in-differences mean that the runners end up closer together.
-Hence, the Hahner twins have a difference-in-difference of -114 seconds.
-They finished 114 seconds closer together than the distance between their personal bests.
-
-qq
-
-As a brief overview of the race, Figure \ref{fig:secondsbehind}
-describes each runner's status at the various split times on the
-marathon course.  Each dot in the figure depicts a recorded split and
-the number of seconds each runner was behind the race leader at the
-time.  There are more dots at earlier splits due to the accumulation
-of DNFs.  Regarding twins and triplets, the Estonian Luik triplets are
-denoted with blue dots; there are three blue dots through the 20
-kilometer split, but the third triplet did not record a half-marathon
-split.  The orange dots in the figure represent the Hahner twins; as
-noted in the figure, Lisa Hahner was ahead of her sister through 15
-kilometers at which point Anna surged ahead.  Lastly, Figure
-\ref{fig:secondsbehind} contains two red dots representing the North
-Korean Kim twins; this is not visually apparent because the Kim twins
-had identical split times during the marathon.
+To get a sense of the lockstep nature of the Kim twins race, Figure
+\ref{fig:secondsbehind} describes each runner's status at the various
+split times on the marathon course.  Each dot in the figure depicts a
+recorded split and the number of seconds each runner was behind the
+race leader at the time.  There are more dots at earlier splits due to
+the accumulation of DNFs.  The usual color scheme applies here, and
+the Estonian DNF prior to the half marathon split is evident.  The
+orange dots in the figure represent the Hahner twins, and as noted in
+the figure Lisa Hahner was ahead of her sister through 15 kilometers
+at which point Anna surged ahead.  Figure \ref{fig:secondsbehind}
+contains two red dots representing the North Korean Kim twins, but
+this is not visually apparent because the Kim twins had identical
+split times during the entire race marathon.
 
@@
 \end{figure}
-
-Figure \ref{fig:secondsbehind} does not directly shed any light on our
-motivating research question, but it does show that the Hahner twins
-finished amidst many other runners.  We will come back to this point
-later.
 
@@
 each other. The challenge, of course, is that this distribution is
-unknown.  % Therefore it needs to be estimated.
+unknown.
 
 \section*{Modeling}
-
-%% THIS SECTION IS A MESS.  I JUST WANTED TO PUT SOMETHING IN WRITING.
-
-%% Ok!  Not sure what all the notation means.
-
-% We know that Anna and Lisa Hahner finished the marathon in Rio
-% simultaneously. We don't know, however, if the finish was intentional;
 
@@
 that they ran independently, as they say they did. Hence, we are
-looking to determine, %% y_a versus Y_a.  Former realized?
+looking to determine,
 
@@
 the expected difference had they run independently. Therefore, under a
-coordinated finish we would expect
-$\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid \psi = I )$.
+coordinated finish we would expect $\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid \psi = I )$.
 
@@
 \begin{enumerate}
-\item Removing the two sets of twins and triplets from the sample,
-  estimate a linear model that predicts a runner's final time ($Y_i$)
-  based on her pre-Olympic personal best time ($X_i$).
-\item Extract the coefficients and the variance-covariance matrix from
-  this model.
-\item For each runner, draw $\beta_0$ and $\beta_1$ from a multivariate normal
-distribution with mean equal to the coefficients and variance equal to the
-variance-covariance matrix.
-\item For each runner, draw an error term ($e$) from a normal distribution with
-mean equal to zero and a standard deviation equal to the standard
+\item Ignoring twins and triplets, estimate with least squares a
+  linear model that predicts a runner's finishing time ($Y_i$) based
+  on her pre-Olympic personal best time ($X_i$) and her best time
+  squared.
+\item Extract coefficient vector and the covariance matrix from this
+  model.
+\item For each simulated rate, draw intercept and slope estimates
+  $\beta_0$, $\beta_1$, and $\beta_2$, respectively, from a
+  multivariate normal distribution with mean equal to the estimated
+  coefficient vector and covariance equal to the estimated covariance
+  matrix.
+\item For each runner, draw an error ($e$) from a normal distribution
+  with mean zero and a standard deviation equal to the standard
   deviation of the model's residuals ($\sigma$).
 \item For each runner, predict the final result by combining the randomly
-  generated beta coefficients and error terms, $\hat{y} = \beta_0 + \beta_1 + e$.
-\item Eliminate each runner from the race with a probability equal to the
-percent of the total runners who did not finish in the actual race.  This
-account for the likelihood of a DNF.
-\item Calculate the difference in time between Anna Hahner and Lisa Hahner.
-\item Calculate the difference in ranking between Anna Hahner and Lisa Hahner.
-\item Repeat steps 3 through 8 ten thousand times.
-\item Plot a histogram of the results in Figure \ref{fig:simdiff}
+  generated beta coefficients and error terms, $\hat{y} = \beta_0 +
+  \beta_1\,X_i + \beta_2\,X_i^2 + e$.
+\item Eliminate each runner from the race with a probability equal to
+  the fraction of runners who did not finish the marathon.
+\item Calculate the difference in time and difference in ranking
+  between Anna Hahner and Lisa Hahner, assuming both finished the
+  race.
+\item Repeat above steps ten thousand times.
+\item Plot a histogram of the results in Figure \ref{fig:simdiff}.
 \end{enumerate}
@@
 
+\section*{Conclusion}
+
+\newpage
+\section*{Further reading}
+
+\begin{itemize}
+
+\item \url{http://www.telegraph.co.uk/olympics/2016/08/17/german-twins-criticised-for-finishing-olympic-marathon-fun-run-h}
+
+\item
+  \url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
+
+\item
+  \url{https://www.welt.de/sport/olympia/article157669264/Das-falsche-Laecheln-der-deutschen-Lauf-Zwillinge.html}
+\end{itemize}
+
 \end{document}

### 2017-02-02 paper/paper.tex
feb 2 edits

@@
 hours---but we nonetheless want to know what the data from the Olympic
-marathon data tell us.  Was the Hahner finish in the 2016 women's
+marathon tell us.  Was the Hahner finish in the 2016 women's
 Olympic marathon a lovely coincidence or something else?
@@
 the marathon.  For the 8,778 pairs of 133 finishers, Figure
-\ref{fig:diffdiff} plots differences in finishing times against
+\ref{fig:diffdiffscatter} plots differences in finishing times against
 differences in personal best times, and pairs of twins/triplets are
@@
 
-However, if we limit the observations to only those runners who had personal best times that were at least as close to each other as the Hahner twins, we get a different persepective.
+However, if we limit the observations to only those runners who had personal best times that were at least as close to each other as the Hahner twins, we get a different perspective.
 Among those runners,  the Hahner twins reduced the distance between them to the greatest extent.
@@
 \newpage
-\subsection*{Steps of the simulation}
+\subsection*{Simulating the Marathon}
+
+The Hahner sisters claimed that they ran the marathon independent of each other.
+Their simultaneous finish was no more than a product of ability and luck.
+One might expect such a finish given that their personal best marathons are within 114 seconds of each other.
+Add a little bit of luck to the occasion and, as the twins have claimed all along, one might easily argue that the simultaneous finish is simply a joyous coincidence rather than a wholly implausible event.
+
+This begs the question: if we take into consideration their similarities as runners as well as some natural variation in finishing a marathon, what are the odds that the twins unintentionally finished at the same time?
+To answer this question we would want to know the counterfactual distribution of potential finishing times that would have occurred had the twins independently and repeatedly run the marathon over again.
+Holding the twins' similarities constant, this would establish the set of potential outcomes that would have occurred as result of natural variation in the finishing the race.
+Then we could use this distribution of outcomes to determine the likelihood that a simultaneous finish (or at least a near-simultaneous finish) occurred by chance alone.
+If the twins rarely finished the race together in this counterfactual, then one should be skeptical that such a finish occurred without coordination.
+
+Unfortunately, it is not possible to rerun the women's marathon in Rio to establish this counterfactual distribution.
+However, we can attempt to simulate it, by estimating the distribution of every other runner's results -  conditioned on their prior abilities - and drawing from that distribution in order to observe the likelihood that Lisa and Anna finish simultaneously.
+
+To estimate the conditional distribution of every runner's results, we must assume that each runner's time is distributed with a mean that is a quadratic function of $X_i$  and $\beta_j$, such that
+
+$$Y_i \mid X_i \sim N(\beta_0 + \beta_1X_{i} + \beta_2X^2_{i}  + e_i, \sigma).$$
+
+This means that we are assuming that each runner's time is drawn from the normal distribution with a mean that is a quadratic function of their personal best and variance that is constant across all runners.
+We simply estimate $\beta_j$ and $\sigma$ using ordinary least squares.
+We then draw every runner's time from this estimated distribution, conditioned on their personal best outcome.
+Once a race is simulated, we locate Anna and Lisa and record the time between their finishes and the difference in their rank.
+We then simulate a new race - drawing a new set of finishing times - and record the same quantities.
+After repeating this procedure 10,000 times, we plot the distribution in Figure \ref{fig:simdiff}.
+The steps of the simulation are detailed below.
 
@@
 
+We can observe from the results of the simulations in Figure \ref{fig:simdiff} that if Anna and Lisa had run independently of each other, they would have rarely finished simultaneously.
+We see that Anna and Lisa finished within 30 seconds of each other in less than 300 of the 10,000 simulated races.
+And they finished in consecutive rank in less than 175 of the 10,000 races.
+Therefore, the close finish that we observed would be very unlikely if we were to believe the twins raced independently of each other.
+
 \begin{figure}[!ht]
@@
     \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
+    \caption{Simulated differences in places}
+    \label{fig:simulatedranks}
+  \end{subfigure}
+\end{figure}
+
+\begin{figure}[!ht]
+  \centering
+  \caption{Distribution of simulated results for the Hahner twins (using halfway split as predictor)}
+  \label{fig:simdiffhalf} 
+  \begin{subfigure}{.45\textwidth}
+    \includegraphics[width=\textwidth,
+    keepaspectratio]{simulated_time_half.pdf}
+    \caption{Simulated differences in finishing times}
+    \label{fig:simulatedfinishtimes}
+  \end{subfigure}
+  \begin{subfigure}{.45\textwidth}
+    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half.pdf}
     \caption{Simulated differences in places}

### 2017-02-02 2-simulate.R

@@ -225,11 +225,15 @@ pct_consecutive_hahner <- sum(hahner_rank)/length(hahner_rank)
 
 hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
 
+max_t <- max(hahner_df$time_diff, na.rm = T)
+max_r <- max(hahner_df$rank_diff, na.rm = T)
+
 pdf("plots/simulated_time.pdf", height = 5, width = 5)
 ggplot(hahner_df, aes(time_diff)) + 
   geom_histogram(binwidth = 30, colour = "black", fill = NA) +
   xlab("\nDifference in seconds") +
   ylab("Count\n") +
+  xlim(0, max_t) + 
 #  ggtitle("\nSimulated final time") +
   theme_bw()
 dev.off()
@@ -239,10 +243,13 @@ ggplot(hahner_df, aes(rank_diff)) +
   geom_histogram(binwidth = 1, colour = "black", fill = NA) +
   xlab("\nDifference in rank") +
   ylab("Count\n") +
+  xlim(0, max_r) + 
 #  ggtitle("\nSimulated final rank") +
   theme_bw()
 dev.off()
 
+
+source("simhalf.r")
 
 # Compare Gyong to Song
 #

### 2017-02-02 seconds-behind.R

@@ -4,10 +4,12 @@ dta <- read.csv (file = "data/times.csv", header = TRUE)
 
 pdf (file = "plots/seconds-behind.pdf", width = 5, height = 5)
 
-plot (0,0, cex = 0, xlim = c(1, 10), ylim = c(0, 2000),
+plot (0,0, cex = 0, xlim = c(5, 42.195), ylim = c(0, 2000),
       xlab = "Split", ylab = "Seconds behind",
       xaxt = "n",
       cex.axis = 0.75)
+
+dists <- c(5, 10, 15, 20, 42.195/2, 25, 30, 35, 40, 42.195)
 
 for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     dta2use <- dta[!is.na(dta[,col2use]), ]
@@ -16,9 +18,9 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     ## rows for the matrix being printed change per split when runners
     ## drop out
     cex2use <- rep(0.15, times = nrow(dta))
-    cex2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- 1
-    cex2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- 1
-    cex2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- 1
+    cex2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- .7
+    cex2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- .7
+    cex2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- .7
     
     pch2use <- rep(19, times = nrow(dta2use))
     pch2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- 19
@@ -30,18 +32,26 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- "red"
     color2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- "blue"
 
-    points (x = rep (match(col2use, cols2use), times = nrow(dta2use)), y = dta2use[, col2use] - min(dta2use[, col2use]),
[…]

### 2017-02-02 simhalf.R

@@ -1,0 +1,88 @@
+# Calculate the percent of runners who did not finish.
+# (includes women who did not start)
+df_half <- df[!is.na(df$SPLIT_HALF),] 
+dnf_rate <- sum(is.na(df_half$FINAL))/length(df_half$FINAL)
+
+# Estimate linear relationship between personal best and marathon result, excluding twins.
+m <- lm(FINAL ~ SPLIT_HALF + I(SPLIT_HALF^2), data = df_half[is.na(df_half$TWINS),])
+
+# Store the total number of runners, N
+N <- length(df_half$SPLIT_HALF)
+
+# Save the vector of personal bests
+x <- df_half$SPLIT_HALF
+
+# Create data frame to capture simulation output
+sim_df <- data_frame(ATHLETE = df_half$ATHLETE)
+
+# Set seed
+set.seed(14252345)
+
+# Set total number of sims
+total_sims <- 10000
+
+# Begin simulation
+for (i in 1:total_sims){
+  
+  # Determine whether each runner finishes given the dnf rate
+  dnf <- sample(x = c(F,T), size = N, replace = T, prob = c(1-dnf_rate, dnf_rate) )
+  
+  # Draw each runner's result from a multivariate normal distribution with 
+  # mean Beta and sigma equal to the variance-covariance of Beta
+  # Draw an error term from a normal distibution with standard deviation equal 
+  # to the standard deviation of the model residuals.
+  B <- mvrnorm(n = 1, mu = coef(m), Sigma =  vcov(m))
+  e <- rnorm(n = N, mean = 0, sd = summary(m)$sigma)
+  
+  # Remove the results of those that did not finish
+  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] +  e )
+  
+  
+  # Collect the results
+  sim_df[[paste0(i)]] <- y_hat
+  
+}
+
+sim_df <- sim_df %>% gather(SIM, YHAT, -ATHLETE, convert = T) %>% group_by(SIM) %>% mutate(RANK = min_rank(YHAT))
+
+
+# Compare Anna to Lisa
+#
+anna <- df %>% filter(ATHLETE == "Anna Hahner")
+anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
+
[…]

### 2017-02-03 paper/paper.tex
description of simulation histograms

@@
 
-We can observe from the results of the simulations in Figure \ref{fig:simdiff} that if Anna and Lisa had run independently of each other, they would have rarely finished simultaneously.
-We see that Anna and Lisa finished within 30 seconds of each other in less than 300 of the 10,000 simulated races.
-And they finished in consecutive rank in less than 175 of the 10,000 races.
-Therefore, the close finish that we observed would be very unlikely if we were to believe the twins raced independently of each other.
-
 \begin{figure}[!ht]
@@
 \end{figure}
+
+A histogram of the differences between Anna and Lisa's simulated finishes can be found in Figure \ref{fig:simdiff}.
+The first plot shows the distribution of the absolute differences in time.
+Differences are grouped in 30 second bins along the x-axis and the count for each bin is denoted by the vertical length of bar along the y-axis.
+The second plot shows the distribution of the absolute differences in ranking.
+Differences are grouped as single units ranging from no runner between the twins to nearly 120 runners between them.
+
+The results plotted in the histograms question the credibility of Anna and Lisa's story.
+The results suggest that  a simultaneous finish would be very rare if Anna and Lisa had run independently of each other.
+For example, in less than 300 of the 10,000 simulated races did Anna and Lisa finished within 30 seconds of each other .
+And they finished in consecutive rank in less than 175 of the 10,000 races.
+Therefore, the close finish that we observed, where Anna and Lisa crossed the finish line one-after-the-other, would be very unlikely if we were to believe the twins raced independently of each other.
+
+The results holds even if you replace their personal best time with their time halfway through the race.
+While this reduces the variation in the predicted outcome of each runner and, therefore, reduces the expected distance between Anna and Lisa, it is still quite rare for Anna and Lisa to finish simultaneously or consecutively.
 

### 2017-02-03 paper/paper.tex
paper updates on friday

@@
 
-On August 14, 2016, at 9:30 in the morning, the Women's Olympic
-marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
-80 countries across the world left the starting line en route to their
-destination 42.195 kilometers away. Two hours, twenty-four minutes,
-and four seconds later, Jemima Sumgong of Kenya would be the first to
-cross the finishline and take home gold; Sumgong was just 3.5 minutes
-behind her prior personal best time in the marathon. Approximately 21
-minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
-would cross the finishline together, holding hands and celebrating a
-personal victory. Although the Hahners would finish 81st and 82nd,
-respectively, with times slightly more than 18 minutes slower than
-corresponding personal bests, Anna Hahner would describe their joint
-finish as a ``magical moment.''
-
-The media quickly picked up on the Hanher story as an image of the
-beaming twins finishing hand-in-hand captured a public audience. Many
-believed the moment was a reflection of the Olympic spirit.
+At 9:30am on August 14, 2016, the Women's Olympic marathon kicked off
+in Rio de Janeiro, Brazil, when 156 runners from 80 countries across
+the world left the starting line en route to their destination 42.195
+kilometers away. Two hours, twenty-four minutes, and four seconds
+later, Jemima Sumgong of Kenya would be the first to cross the
+finishline and take home gold; Sumgong was just three and one-half
+minutes slower than her prior personal best time in the
+marathon. Approximately 21 minutes later, twin marathoners from
+Germany, Anna and Lisa Hahner, would cross the finishline together,
+holding hands and celebrating a personal victory. Although the Hahners
+would finish 81st and 82nd, respectively, well behind the winners of
+the marathon, Anna Hahner would describe their joint finish as a
+``magical moment.''
 
@@
 
-Not everyone agreed with this rosy interpretation. The twins' happy
-facial expressions at the finish were portrayed as a bit
-contrived---smiling like ``Honigkuchenpferde,'' cookies in the shape
-of a horse---and the sports director of the German Athletics
-Federation, Thomas Kurschilgen, stirred up controversy when he
-suggested that the Hahners' photo-finish was no coincidence.
-Kurschilgen averred that the twins slowed down so as to finish
-simultaneously and create a spectacle which would ``generate media
-attention.'' Kurschilgen justified his charge with the fact that the
-twins ran in the Rio marathon at least 18 minutes slower their
-personal best times prior to the
+The media quickly picked up on the Hanher story as an image of the
+beaming twins finishing hand-in-hand captured a public audience. While
+many believed the moment was a reflection of the Olympic spirit, not
+everyone agreed with this rosy interpretation. The twins' happy facial
+expressions at the finish were portrayed as a bit contrived---smiling
+like ``Honigkuchenpferde,'' cookies in the shape of a horse---and the
+sports director of the German Athletics Federation, Thomas
+Kurschilgen, stirred up controversy when he suggested that the
+Hahners' photo-finish was no coincidence.  Kurschilgen averred that
+the twins slowed down so as to finish simultaneously and create a
+spectacle which would ``generate media attention.'' Kurschilgen
+justified his charge with the fact that the twins ran in the Rio
+marathon at least 18 minutes slower their personal best times prior to
+the
 Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
@@
 Among female Olympic marathoners, the Hahner twins were not alone in
-their familial ties, and we touch on this in the analysis that
-follows.  The marathon also featured twins from North Korea, Kim
-Hye-song and Kim Hye-gyong, who posted identical times and finished
-10th and 11th in the race, respectively. The Kim finish, unlike the
-Hahner finish, appears devoid of post-race controversy. Moreover,
-three triplets from Estonia competed in the Rio marathon, although
-only two, Lily Luik and Leila Luik, finished it, in 97th and 114th
-place, respectively. The third Estonia triplet, Liina Luik, recorded
-what is known as a DNF---an abbreviation that means did not finish, a
-term that we will use throughout this article.
+their familial ties.  The marathon also featured twins from North
+Korea, Kim Hye-song and Kim Hye-gyong, who posted identical times and
+finished 10th and 11th in the race, respectively. The Kim finish,
+unlike the Hahner finish, appears devoid of post-race
+controversy. Moreover, three triplets from Estonia competed in the Rio
+marathon, although only two, Lily Luik and Leila Luik, finished it, in
+97th and 114th place, respectively. The third Estonia triplet, Liina
+Luik, recorded what is known as a DNF---an abbreviation that means did
+not finish.  Although our focus here is the Hahner twins, we touch on
+the Kim twins and Luik triplets in the course of what follows.
 
@@
 The Kurschilgen accusation against the Hahner twins has two
-components, that the twins finished simultaneously, holding hands and
-smiling, \emph{and} that they ran slowly.  We suspect that Kurschilgen
-would not have expressed ire at the Hahners had they finished in 1st
-and 2nd place in Rio, hand-in-hand with wide grins, but of course we
-do not know this.  As such, our investigation of the charges that
-Kurschilgen offered will distinguish between the idea of slow finish
-versus a simultaneous finish.
+components, that these two women ran slowly \emph{and} that they
+finished simultaneously.  We suspect that Kurschilgen would not have
+expressed ire at the Hahners had they finished in 1st and 2nd place in
+Rio, hand-in-hand with wide grins, but of course we do not know this.
+As such, our investigation of the charges that Kurschilgen offered
+will distinguish between the idea of slow finish versus a simultaneous
+finish.
 
-Our research design has two components.  In the first, we present
-visualizations that describe various features of the Hahner twin's
-results, and an important element of our visualizations is the
-difference between a runner's Rio time and her prior personal best
-time in the marathon.  Overall, our visualizations suggest blah blah
-bhah.  We then turn to modeling and in particular a regression-based
-simulation.  In simulated Olympic marathons, which we develop based on
-relationships (excluding the Hahner and Kim twins and Luik triplets)
-between known personal best times and observed Rio finishing times, we
-find that the Hahner twins finished suspiciously close to each other
-given the disparty between their personal best times.  OR THEY DID
-NOT.  Finally, we return to Kurschilgen claims  about the Hahners and
-offer our thoughts about their validity.
+Our research design is twofold.  First, we present visualizations that
+describe various features of the Hahner twin's results, and an
+important element in our visualizations is the difference between a
+runner's Rio time and her prior personal best time in the marathon.
+Overall, our visualizations suggest that the Hahner twins' pace in the
+marathon was slow, but not excessively so, but that their simultaneous
+finish was quite unusual given the twins' differences in abilities
+(and similarly for the Kim twins).  We then turn to a regression-based
+simulation of the marathon, which we develop based on relationships
+(excluding the Hahner and Kim twins and Luik triplets) between known
+personal best times and observed Rio finishing times.  Our simulations
+show that the Hahner twins finished suspiciously close to each other
+given the disparty between their personal best times.  Finally, we
+return to Kurschilgen claims about the Hahners and offer our thoughts
+about their validity.
 
@@
 Figure \ref{fig:scatter} contains two plots, both of which describe
-how the Rio Olympic women's marathon results varied as a function of
-athletes' personal best times and their half marathon split times.
-The points in each plot are colored by twin/triplet status, and both
-plots contain second-order polynomial regression lines.  We treat an
-athlete's prior personal best marathon time as a measure of the
-athlete's underlying marathon talent, and we interpret an athlete's
-half marathon split as a measure of the athlete's skill at marathon
-running on August 14, 2016, the date of the women's marathon in Rio.
+how finishing times from the Rio Olympic women's marathon varied as a
+function of athletes' personal best times and their Rio half marathon
+split times.  The points in the plots are colored by twin/triplet
+status, and both plots contain second-order polynomial regression
+lines.  We treat an athlete's prior personal best marathon time as a
+measure of the athlete's underlying marathon talent, and we interpret
+an athlete's half marathon split as a measure of the athlete's skill
+at marathon running on August 14, 2016, the date of the women's
+marathon in Rio.
 
@@
 
-Considering first Figure \ref{fig:45degreeplot_half}, note that this
-figure includes a solid 45-degree line as the horizontal and vertical
-axes in the figure correspond to full marathons.  The rug marks on the
-bottom of the figure denote the personal best times of runners who
-DNFed.  Given the paucity of points (five of them) below the 45-degree
-line, the vast majority of Olympic marathoners ran slower in Rio
-compared to their personal bests.  Relative to personal best times,
-the Hahner twins were definitely on the slow side, but a number of
-runners had greater differences between their Olympic times and their
-personal bests than the Hahners.  These runners are denoted with
+Figure \ref{fig:45degreeplot_half} includes a solid 45-degree line as
+the axes in the figure correspond to full marathons; in addition, the
+figure's rug marks denote personal best times of runners who DNFed.
+Given the paucity of points (five of them) below the pictured
+45-degree line, the vast majority of Olympic marathoners ran slower in
+Rio compared to their personal bests.  Relative to personal best
+times, the Hahner twins were definitely on the slow side, but a number
+of runners had greater differences between their Olympic times and
+their personal bests than the Hahners.  These runners are denoted with
 squares in Figure \ref{fig:45degreeplot}, and there are 13 such
@@
 had personal best times slightly faster than the Hahner's and yet
-finished after both of the German women.  Although Figure
+finished after the German twins.  Although Figure
 \ref{fig:45degreeplot} suggests that the Hahner twins were slower than
@@
 underlying athletic talent is that these times are potentially
-confounded by the marathon courses at which they were set (some*sh
+confounded by the marathon courses at which they were set (some
 courses, like the Berlin marathon, are known for fast times) and race
@@
 With that in mind, Figure \ref{fig:45degreeplot_half} plots marathon
-finishing times against half marathon splots.  This figure has a
+finishing times against half marathon splits.  This figure has a
 regression line as before but no 45-degree line.
@@
 on their half marathon split times.  This is consistent with the
-previous Figure \ref{fig:45degreeplot} and inconsistent with the
-accusations against the Hahner twins, at least the part of the
-accusation that focuses on their excessively slowing down.
+previous Figure \ref{fig:45degreeplot} and inconsistent with
+accusations made against the Hahner twins, at least the part of the
+accusation that focused on their  overall pace.
 
@@
 their official finishing times were separated by one second.  Is such
-a 115 to one compression typical among pairs of runners or was it
-unusual?  We might ask, are there other pairs of runners who had a
-difference between personal best times of 115 seconds apart and, if
-so, how close where their finishing times?
+a 115 to one compression typical among pairs of runners or, in
+contrast, was it unusual?  Are there other pairs of marathoners who
+had a difference between personal best times of 115 seconds apart and,
+if so, how close where their finishing times?
 
@@
 exactly a 115 second gap in personal best times.  Of these ten pairs,
-differences in finishing times in seconds are as follows: 36, 93, 172,
+differences in finishing times, in seconds, are as follows: 36, 93, 172,
 319, 379, 459, 552, 671, 675, and 739.  In other words, of all pairs
@@
 We can generalize this result by looking at all pairs of runners in
-the marathon.  For the 8,778 pairs of 133 finishers, Figure
+the marathon.  For all 8,778 pairs of 133 finishers, Figure
 \ref{fig:diffdiffscatter} plots differences in finishing times against
@@
 second.  However, there are many points about the Hahner's orange dot,
-and this shows that, conditional on a 115 second difference in
-personal best times, most marathoners did not have close finishing
-times like the Hahnes.  Some pairs of runners with 115 second personal
-best differences had finishing time differences of 1000 seconds, i.e.,
-in excess of 15 minutes.  Of course the points in Figure
-\ref{fig:diffdiffscatter} are not independent, but they provide a
-sense of the dispersion in finishing time differences that one can
+and this shows that, conditional on an approximate 115 second
+difference in personal best times, most marathoners did not have close
+finishing times like the Hahnes.  Some pairs of runners with around
+115 second personal best differences had finishing time differences of
+1000 seconds, i.e., in excess of 15 minutes.  Of course the points in
+Figure \ref{fig:diffdiffscatter} are not independent, but they provide
+a sense of the dispersion in finishing time differences that one can
 expect conditional on differences in personal best times.
@@
 difference of 53 seconds and a finishing time difference of literally
-zero seconds.  Beyond these twins, there were eight pairs of runners
-with a 53 second personal best difference, and resulting finishing
-time differences are as follows: 9, 51, 228, 340, 352, 571, 662, and
-751.  As in the Hahner case, the Kim twins compressed their finishing
-times more than any other pair of runners with a similar personal best
-difference.
+zero seconds.  Beyond these twins, there were eight pairs of Rio
+runners with a 53 second personal best difference, and resulting
+finishing time differences are as follows: 9, 51, 228, 340, 352, 571,
+662, and 751.  As in the Hahner case, the Kim twins compressed their
+finishing times more than any other pair of runners with a similar
+personal best difference.
 
@@
 the figure Lisa Hahner was ahead of her sister through 15 kilometers
-at which point Anna surged ahead.  Figure \ref{fig:secondsbehind}
-contains two red dots representing the North Korean Kim twins, but
-this is not visually apparent because the Kim twins had identical
-split times during the entire race marathon.
+at which point Anna surged.  Figure \ref{fig:secondsbehind} contains
+two red dots representing the North Korean Kim twins, but this is not
+visually apparent because the Kim twins had identical split times
+during the entire race marathon.
 

### 2017-02-03 paper/paper.tex
paper updates on friday

@@
 
-\newcommand{\yi}{\ensuremath{Y_i}}
+\newcommand{\Yi}{\ensuremath{Y_i}}
 
@@
 
-\section*{The probability of an unintentional simultaneous finish}
+\section*{Simulating the Marathon}
 
-Thomas Kurschilgen's accusations of an intentional finish stems from
-two observations. First, the Hahner twins finished slower than he
-expected given respective personal bests. Second, the twins finished
-together at the exactly same time. Kurschilgen clearly believed that
-neither of these events would have occurred had both Hahner run the
-marathon independently, absent coordination. By his logic, the German
-twins should have run faster and not have finished simultaneously.
+% The Hahner sisters claimed that they ran the marathon independent of
+% each other.  Their simultaneous finish was no more than a product of
+% ability and luck.  One might expect such a finish given that their
+% personal best marathons are within 114 seconds of each other.  Add a
+% little bit of luck to the occasion and, as the twins have claimed all
+% along, one might easily argue that the simultaneous finish is simply a
+% joyous coincidence rather than a wholly implausible event.
 
-On the other hand, only a handful of runners completed the Rio
-marathon with times that were faster than their recorded best times.
-Thus, 18 minutes behind a personal record may not be the outlier that
-Kurschilgen claimed it to be. Moreover, there is good reason to
-suspect that an unintended simultaneous finish was more likely for the
-Hahner twin than it would be for any other set of runners. After all,
-these two women are twins with presumably similar abilities. Not only
-do they train together, but their pre-Rio best times are less than two
-minutes apart. While Anna may be slightly faster than Lisa measured by
-personal bests, we might expect the difference in their Rio result to
-be just as close as the differences in their recorded bests. And given
-random variation in finishing times, a simultaneous finish might not
-be out of the ordinary.
+We now consider the following: if we take into consideration the
+Hahner twin's similarities as runners as well as some natural
+variation in finishing a marathon, what is the probability that the
+twins finished at roughly the same time and/or sequentially?  To
+answer this question we need to know the counterfactual distribution
+of potential marathon finishing times that would have occurred had the
+Hahner twins independently and repeatedly run the Rio marathon over
+again.  Holding the twin's abilities constant, this would establish
+the set of potential outcomes that would have occurred as result of
+natural variation in the finishing the marathon.  We could use this
+distribution of outcomes to determine the likelihood that a
+simultaneous finish by the Hanhers, or at least a near-simultaneous
+finish, occurred by chance alone.  If the twins rarely finish the
+marathon together in such a counterfactual world, then one might be
+skeptical that their observed finish occurred without
+coordination---as they in fact claimed.
 
-Even though they seem similar, finishing at very similar times and
-finishing together are different phenomena. If, for example, both
-Hahners were of similar ability to each other and also to many other
-runners, then we might expect similar finishing times yet not
-necessarily similar placements. The latter will be a function of the
-extent to which all runners on the Rio course have similar talent
-levels.  This point is an important one and will be evident in the
-results that follow.
+Unfortunately for us but nonetheless fortunately for the race's
+participants, it is not possible to rerun the women's Olympic marathon
+in Rio to establish a counterfactual distribution of potential race
+outcomes for the Hahner twins.  However, we can attempt to simulate
+this distribution by estimating the distribution of every other
+runner's finishing time, conditioned on prior abilities, and drawing
+from that distribution in order to observe the likelihood that Lisa
+and Anna Hahner finish simultaneously.
 
-To test the assertion that the Hahner twins paced themselves to finish
-at the same time and with back-to-back placements, we need to compute
-the probability that such a result would have occurred
-unintentionally. Hence we need to know the probability distribution of
-Anna and Lisa's finishing times if their runs had been independent of
-each other. The challenge, of course, is that this distribution is
-unknown.
+To estimate the conditional distribution of Rio runner's result, we
+assume that each runner's marathon time \Yi\ is distributed normally
+with a mean that is a quadratic function of her personal best time
+$X_i$:
 
-\section*{Modeling}
-
-The Hahner twins either acted independently $I$ or coordinated their
-times $C$; these are the two possible states of the race
-$\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
-times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
-that they ran independently, as they say they did. Hence, we are
-looking to determine,
-
-$$P(\psi = I \mid y_A  \cap y_L )$$
-
-However, to determine this, we need to have some understanding of a
-likelihood function that specifies Rio finishing times. We want to
-know the likelihood of Anna's and Lisa's final times given
-independence $P(y_A \cap y_L \mid \psi = I )$. We can estimate this
-function with a few assumptions. First, we assume that under
-independence, any given runner's final time $Y_i$ is conditional on
-his/her running ability plus noise. Specifically, we assume that the
-$Y_i$ is a linear function of the runner's ability $X_i$ plus a
-normally distributed error term $e_i \sim N(0, \sigma)$. Second, we
-assume that every runner shares the same linear relationship - meaning
-the slope and intercept remain constant across runners. Third, we
-assume that the error term is drawn from a common distribution across
-runners. Hence, luck and misfortune are drawn from the same
-distribution. Therefore,
-
-$$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e_i, \sigma)$$
-
-We also assume that a runner's ability $X_i$ can be measured precisely
-by her best marathon performance leading up to the Olympics.
-
-%% Any measurement error must therefore be negligible.
-
-If Anna and Lisa intentionally slowed down as a result of coordination
-then we would likely observe $y_A > E(Y_A \mid \psi = I) $ and
-$y_L > E(Y_L\mid \psi = I )$. In other words, the final times that
-Anna and Lisa recorded in the race would be greater than we would
-expect if they had run independently.
-
-Moreover, if Anna and Lisa coordinated to finish simultaneously, then
-the difference between the two sisters' final times would be less that
-the expected difference had they run independently. Therefore, under a
-coordinated finish we would expect $\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid \psi = I )$.
-
-\section*{Did Anna and Lisa intentionally slow down?}
-
-According to Kurschilgen, Anna and Lisa underperformed in the Rio marathon.
-He claimed that because their goal was to finish simultaneously rather than finish at their fastest pace, their times were slower than they otherwise would have been.
-He claimed that the twins were simply trading speed for a photo-finish.
-
-If this were the case, the function generating Anna and Lisa's final times would deviate from the function that generated everyone else's final times.
-Given everyone else would draw their times from the independent distribution  $Y_i \sim N(\beta_0 + X_{i}\beta_1 + e, \sigma)$,  Anna and Lisa would draw from a distribution of times that are slower in expectation.
-Hence, they would lie well-above the line that links a runner's performance in Rio to their previous best performance.
-
-\begin{figure}[!ht]
-  \caption{Relationship between Personal Best and Result  -- caption?}
-  \label{fig:scatter}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
-    \caption{Rio finishing times and personal best times}
-    \label{fig:45degreeplot}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
-    \caption{Studentized residuals}
-    \label{fig:studentizedresiduals}
-  \end{subfigure}
-\end{figure}
-
-Figure \ref{fig:45degreeplot} displays the relationship between a
-runner's personal best (horizontal axis) and her final result
-(vertical axis).  The 133 finishers are plotted as grey points, and
-the 23 DNFs are depicted below along the horizontal axis with
-tick-marks.  The dashed black line with grey confidence intervals
-represent ordinarily least squares estimates of the linear
-relationship between racer's finishing and personal best times.
-
-One can see in Figure \ref{fig:scatter} that there is a clear
-relationship between a runner's personal best time and her performance
-in the Olympics.  While there is a significant amount of noise in the
-relationship, the Hahner twins' dots are well above the regression's
-fitted line notwithstanding this point.  They finished the Rio
-marathon much slower than they should have, per their pre-Olympic
-personal best times.  In fact, if we sort studentized residual, we can
-see in Figure \ref{fig:studentizedresiduals} that the Hahner twins are
-in the tail-end of the residual distribution.  Anna Hahner's residual
-deviation is greater than 91.7\% of the runners who completed the
-marathon while Lisa Hahner's deviation is greater than 86.5\%.  Hence,
-the twins appear to have finished at a much slower pace than expected,
-which is what we might expect if they were coordinating their runs.
-The North Korean Kim twins are also in the tail of the residual
-distribution---but the left tail.  These twins finished much faster
-than expected.
-
-\section*{Did Anna and Lisa intentionally finish together?}
-
-Anna and Lisa finished the marathon at an unusually slow rate, which is what we might expect if the twins intentionally slowed their pace to finish simultaneously.
-However, the observation of a slow finish does not necessarily imply that it was an intentional result .
-It only suggests that Kurschilgen's claim regarding this issue holds up in the data.
-
-Perhaps it would be easier to refute the claim if we could show that the Twins's back-to-back finish was not such an unusual event.
-Given that Anna and Lisa were so close in their personal best times, it may seem reasonable to expect that they would also be close in their finishing times.
-Hence, one way to determine if the twins's finish was intentional is to quantify the likelihood that such a close finish would occur by chance given the differences in their ability.
-
-We can attempt to quantify this likelihood by looking at how differences in the personal best time of two runners translate into differences in Olympic times .
-If Anna and Lisa's finish was unintentional we'd expect that such a close finish would be unlikely given their difference in personal best times.
-Yet, how does the difference in personal best times relate to the difference in finishing times?
-
-One way we can estimate this relationship is to look at all of the differences that we observe in the marathon.
-We can take the difference in personal best time for every combination of runners in the data and look at the difference in their final times.
-We should expect that compared to the difference in their personal best times, the difference in Anna and Lisa's final times should be commensurate with the conditional difference of all other runners.
-%This needs a better explanation.
-
-Therefore, we find all combinations of runners in the marathon and calculate two quantities: 1) the difference in their personal best time and 2) the difference in their final time.
-We plot the relationship between the two quantities in Figure \ref{fig:diffdiffscatter}.
-All 8,778 dyadic combinations of runners are displayed by grey points in the scatter plot.
-The solid black line indicates a one-to-one relationship, where every point would fall if the difference in final times were systematically equivalent to the difference in personal bests.
-
-% \begin{figure}[!ht]
-%  \caption{Relationship between Difference in Personal Best and Difference in Result}
-%  \label{fig:diffdiffscatter}
-%  \centering
-%  \begin{subfigure}{.5\textwidth}
-%  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
-%  \end{subfigure}
-% \end{figure}
-
-We can see from the yellow dot that the Hahner Twins are just below this one-to-one line.
-Hence, the twins' final times were closer to each other than their personal best times; but not by much.
-In fact, this might be an expected difference if we expected there to be a one-to-one relationship.
-However, although there is a clear relationship between the difference in final time and the difference in personal best, it is not a strong one-to-one relationship.
-We've added a dotted smoother to display a cubic splines estimate of the conditional averages.
-While the Hahner twins may be close to the one-to-one line, they are much further away from the average set of runners.
-Hence, we would expect for any pair of runners, the difference in their final results should be much greater than their difference in their personal best.
-However, the Hahner twins do not fit this expectation.
-Yet, how different from this expectation are they?
-
-One way we can quantify the degree to which the twins are outliers is by comparing their difference in difference to every other pair of runner's difference-in-differences.
-We calculate the difference-in-differences for each pair of runners by simply subtracting their difference in result (y-axis) from their difference in personal best (x-axis).
-Therefore positive difference-in-differences mean that the runners ended up further apart than their difference in personal bests.
-Negative difference-in-differences mean that the runners end up closer together.
-Hence, the Hahner twins have a difference-in-difference of -114 seconds.
-They finished 114 seconds closer together than the distance between their personal bests.
-
-We arrange all 8,778 difference-in-differences from small to large in the first plot in Figure \ref{fig:diffdiff}.
-The Hahner twins are denoted by the yellow point.
-Using this plot we can visualize where the twins fall within the distribution of difference-in-differences.
-It is clear that they are closer to the lower tail of the distribution.
-However, compared to all other difference-in-differences, their result is not that extreme.
-Nearly 21\% of the difference-in-dfferences were less than the that of the twins'.
-Hence, we'd expect at least one in five pairs of runners to have a reduction in difference at least as great as the twins.
-
-However, if we limit the observations to only those runners who had personal best times that were at least as close to each other as the Hahner twins, we get a different perspective.
-Among those runners,  the Hahner twins reduced the distance between them to the greatest extent.
-There was no other pair of runners whose reduced the difference in the final time relative to their difference in personal bests more than them.
-
-\begin{figure}[!ht]
- \caption{How Hahner Twins Rank in Difference-in-Differences}
- \label{fig:diffdiff}
- \begin{subfigure}{.5\textwidth}
- \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_1.pdf}
- \end{subfigure}
- \begin{subfigure}{.5\textwidth}
- \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_2.pdf}
- \end{subfigure}
-\end{figure}
-
-\newpage
-\subsection*{Simulating the Marathon}
-
-The Hahner sisters claimed that they ran the marathon independent of each other.
-Their simultaneous finish was no more than a product of ability and luck.
-One might expect such a finish given that their personal best marathons are within 114 seconds of each other.
-Add a little bit of luck to the occasion and, as the twins have claimed all along, one might easily argue that the simultaneous finish is simply a joyous coincidence rather than a wholly implausible event.
-
-This begs the question: if we take into consideration their similarities as runners as well as some natural variation in finishing a marathon, what are the odds that the twins unintentionally finished at the same time?
-To answer this question we would want to know the counterfactual distribution of potential finishing times that would have occurred had the twins independently and repeatedly run the marathon over again.
-Holding the twins' similarities constant, this would establish the set of potential outcomes that would have occurred as result of natural variation in the finishing the race.
-Then we could use this distribution of outcomes to determine the likelihood that a simultaneous finish (or at least a near-simultaneous finish) occurred by chance alone.
-If the twins rarely finished the race together in this counterfactual, then one should be skeptical that such a finish occurred without coordination.
-
-Unfortunately, it is not possible to rerun the women's marathon in Rio to establish this counterfactual distribution.
-However, we can attempt to simulate it, by estimating the distribution of every other runner's results -  conditioned on their prior abilities - and drawing from that distribution in order to observe the likelihood that Lisa and Anna finish simultaneously.
-
-To estimate the conditional distribution of every runner's results, we must assume that each runner's time is distributed with a mean that is a quadratic function of $X_i$  and $\beta_j$, such that
-
-$$Y_i \mid X_i \sim N(\beta_0 + \beta_1X_{i} + \beta_2X^2_{i}  + e_i, \sigma).$$
-
-This means that we are assuming that each runner's time is drawn from the normal distribution with a mean that is a quadratic function of their personal best and variance that is constant across all runners.
-We simply estimate $\beta_j$ and $\sigma$ using ordinary least squares.
-We then draw every runner's time from this estimated distribution, conditioned on their personal best outcome.
-Once a race is simulated, we locate Anna and Lisa and record the time between their finishes and the difference in their rank.
-We then simulate a new race - drawing a new set of finishing times - and record the same quantities.
-After repeating this procedure 10,000 times, we plot the distribution in Figure \ref{fig:simdiff}.
-The steps of the simulation are detailed below.
+$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right).$$
+%%
+where $N\left(\cdot,\cdot\right)$ denotes a normal distribution. We
+estimate $\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least
+squares and our set of 133 finishing runners.  For a simulated
+marathon, we then draw a runner's time from an estimated distribution,
+conditioned on the runner's personal best outcome.  Once a race is
+simulated, we locate Anna and Lisa Hahner and record both the time
+between their finishes and the difference in their ranks.  We then
+simulate a new race---drawing a new set of finishing times---and
+record the same quantities.  In particular, the steps of the
+simulation are as follows.
 
@@
   squared.
-\item Extract coefficient vector and the covariance matrix from this
-  model.
+\item Extract coefficient vector, estimated covariance matrix, and
+  estimated regression variance from this model.
 \item For each simulated rate, draw intercept and slope estimates
-  $\beta_0$, $\beta_1$, and $\beta_2$, respectively, from a
-  multivariate normal distribution with mean equal to the estimated
-  coefficient vector and covariance equal to the estimated covariance
-  matrix.
-\item For each runner, draw an error ($e$) from a normal distribution
-  with mean zero and a standard deviation equal to the standard
-  deviation of the model's residuals ($\sigma$).
+  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
+  respectively, from a multivariate normal distribution with mean
+  equal to the estimated coefficient vector and covariance equal to
+  the estimated covariance matrix.
+\item For each runner, draw an error $\tilde{e}$ from a normal
+  distribution with mean zero and a standard deviation equal to the
+  standard deviation of the original regression model's residuals.
 \item For each runner, predict the final result by combining the randomly
-  generated beta coefficients and error terms, $\hat{y} = \beta_0 +
-  \beta_1\,X_i + \beta_2\,X_i^2 + e$.
+  generated beta coefficients and error terms, $\tilde{y} = \tilde{\beta}_0 +
+  \tilde{\beta}_1\,X_i + \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
 \item Eliminate each runner from the race with a probability equal to
-  the fraction of runners who did not finish the marathon.
+  the observed fraction of marathoners who did not finish the
+  marathon.
 \item Calculate the difference in time and difference in ranking
@@
   race.
-\item Repeat above steps ten thousand times.
-\item Plot a histogram of the results in Figure \ref{fig:simdiff}.
+\item Repeat above steps 10,000 times.
 \end{enumerate}
+
+Before considering results, DNF biz...
+
+Figure \ref{fig:simdiff} contains two histograms of differences
+between Anna and Lisa's finishes.  Figure
+\ref{fig:simulatedfinishtimes} shows the distribution of absolute
+differences in time where differences are grouped in 30 second bins
+along the horizontal axis of the histogram.  The count for each bin is
+denoted by the vertical length of the bars.  Figure
+\ref{fig:simulatedranks} shows the distribution of the absolute
+differences in Hahner twin rankings.  Differences are grouped as
+single units ranging from no runner between the twins to nearly 120
+runners between them.
 
@@
 
-A histogram of the differences between Anna and Lisa's simulated finishes can be found in Figure \ref{fig:simdiff}.
-The first plot shows the distribution of the absolute differences in time.
-Differences are grouped in 30 second bins along the x-axis and the count for each bin is denoted by the vertical length of bar along the y-axis.
-The second plot shows the distribution of the absolute differences in ranking.
-Differences are grouped as single units ranging from no runner between the twins to nearly 120 runners between them.
+The results plotted in the histograms question the credibility of Anna
+and Lisa's story.  The results suggest that a simultaneous finish
+would be very rare if Anna and Lisa had run independently of each
+other.  For example, in less than 300 of the 10,000 simulated races
+did Anna and Lisa finished within 30 seconds of each other .  And they
+finished in consecutive rank in less than 175 of the 10,000 races.
+Therefore, the close finish that we observed, where Anna and Lisa
+crossed the finish line one-after-the-other, would be very unlikely if
+we were to believe the twins raced independently of each other.
 
-The results plotted in the histograms question the credibility of Anna and Lisa's story.
-The results suggest that  a simultaneous finish would be very rare if Anna and Lisa had run independently of each other.
-For example, in less than 300 of the 10,000 simulated races did Anna and Lisa finished within 30 seconds of each other .
-And they finished in consecutive rank in less than 175 of the 10,000 races.
-Therefore, the close finish that we observed, where Anna and Lisa crossed the finish line one-after-the-other, would be very unlikely if we were to believe the twins raced independently of each other.
-
-The results holds even if you replace their personal best time with their time halfway through the race.
-While this reduces the variation in the predicted outcome of each runner and, therefore, reduces the expected distance between Anna and Lisa, it is still quite rare for Anna and Lisa to finish simultaneously or consecutively.
+The results holds even if you replace their personal best time with
+their time halfway through the race.  While this reduces the variation
+in the predicted outcome of each runner and, therefore, reduces the
+expected distance between Anna and Lisa, it is still quite rare for
+Anna and Lisa to finish simultaneously or consecutively.
 
@@
 
+% \section*{The probability of an unintentional simultaneous finish}
+
+% Thomas Kurschilgen's accusations of an intentional finish stems from
+% two observations. First, the Hahner twins finished slower than he
+% expected given respective personal bests. Second, the twins finished
+% together at the exactly same time. Kurschilgen clearly believed that
+% neither of these events would have occurred had both Hahner run the
+% marathon independently, absent coordination. By his logic, the German
+% twins should have run faster and not have finished simultaneously.
+
+% On the other hand, only a handful of runners completed the Rio
+% marathon with times that were faster than their recorded best times.
+% Thus, 18 minutes behind a personal record may not be the outlier that
+% Kurschilgen claimed it to be. Moreover, there is good reason to
+% suspect that an unintended simultaneous finish was more likely for the
+% Hahner twin than it would be for any other set of runners. After all,
+% these two women are twins with presumably similar abilities. Not only
+% do they train together, but their pre-Rio best times are less than two
+% minutes apart. While Anna may be slightly faster than Lisa measured by
+% personal bests, we might expect the difference in their Rio result to
+% be just as close as the differences in their recorded bests. And given
+% random variation in finishing times, a simultaneous finish might not
+% be out of the ordinary.
+
+% Even though they seem similar, finishing at very similar times and
+% finishing together are different phenomena. If, for example, both
+% Hahners were of similar ability to each other and also to many other
+% runners, then we might expect similar finishing times yet not
+% necessarily similar placements. The latter will be a function of the
+% extent to which all runners on the Rio course have similar talent
+% levels.  This point is an important one and will be evident in the
+% results that follow.
+
+% To test the assertion that the Hahner twins paced themselves to finish
+% at the same time and with back-to-back placements, we need to compute
+% the probability that such a result would have occurred
+% unintentionally. Hence we need to know the probability distribution of
+% Anna and Lisa's finishing times if their runs had been independent of
+% each other. The challenge, of course, is that this distribution is
+% unknown.
+
+% \section*{Modeling}
+
+% The Hahner twins either acted independently $I$ or coordinated their
+% times $C$; these are the two possible states of the race
+% $\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
+% times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
+% that they ran independently, as they say they did. Hence, we are
+% looking to determine,
+
+% $$P(\psi = I \mid y_A  \cap y_L )$$
+
+% However, to determine this, we need to have some understanding of a
+% likelihood function that specifies Rio finishing times. We want to
+% know the likelihood of Anna's and Lisa's final times given
+% independence $P(y_A \cap y_L \mid \psi = I )$. We can estimate this
+% function with a few assumptions. First, we assume that under
+% independence, any given runner's final time $Y_i$ is conditional on
+% his/her running ability plus noise. Specifically, we assume that the
+% $Y_i$ is a linear function of the runner's ability $X_i$ plus a
+% normally distributed error term $e_i \sim N(0, \sigma)$. Second, we
+% assume that every runner shares the same linear relationship - meaning
+% the slope and intercept remain constant across runners. Third, we
+% assume that the error term is drawn from a common distribution across
+% runners. Hence, luck and misfortune are drawn from the same
+% distribution. Therefore,
+
+% $$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e_i, \sigma)$$
+
+% We also assume that a runner's ability $X_i$ can be measured precisely
+% by her best marathon performance leading up to the Olympics.
+
+% %% Any measurement error must therefore be negligible.
+
+% If Anna and Lisa intentionally slowed down as a result of coordination
+% then we would likely observe $y_A > E(Y_A \mid \psi = I) $ and
+% $y_L > E(Y_L\mid \psi = I )$. In other words, the final times that
+% Anna and Lisa recorded in the race would be greater than we would
+% expect if they had run independently.
+
+% Moreover, if Anna and Lisa coordinated to finish simultaneously, then
+% the difference between the two sisters' final times would be less that
+% the expected difference had they run independently. Therefore, under a
+% coordinated finish we would expect $\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid \psi = I )$.
+
+% \section*{Did Anna and Lisa intentionally slow down?}
+
+% According to Kurschilgen, Anna and Lisa underperformed in the Rio marathon.
+He claimed that because their goal was to finish simultaneously rather than finish at their fastest pace, their times were slower than they otherwise would have been.
+He claimed that the twins were simply trading speed for a photo-finish.
+
+% If this were the case, the function generating Anna and Lisa's final times would deviate from the function that generated everyone else's final times.
+Given everyone else would draw their times from the independent distribution  $Y_i \sim N(\beta_0 + X_{i}\beta_1 + e, \sigma)$,  Anna and Lisa would draw from a distribution of times that are slower in expectation.
+Hence, they would lie well-above the line that links a runner's performance in Rio to their previous best performance.
+
+% \begin{figure}[!ht]
+%   \caption{Relationship between Personal Best and Result  -- caption?}
+%   \label{fig:scatter}
+%   \begin{subfigure}{.5\textwidth}
+%     \centering
+%     \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
+%     \caption{Rio finishing times and personal best times}
+%     \label{fig:45degreeplot}
+%   \end{subfigure}
+%   \begin{subfigure}{.5\textwidth}
+%     \centering
+%     \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
+%     \caption{Studentized residuals}
+%     \label{fig:studentizedresiduals}
+%   \end{subfigure}
+% \end{figure}
+
+% Figure \ref{fig:45degreeplot} displays the relationship between a
+% runner's personal best (horizontal axis) and her final result
+% (vertical axis).  The 133 finishers are plotted as grey points, and
+% the 23 DNFs are depicted below along the horizontal axis with
+% tick-marks.  The dashed black line with grey confidence intervals
+% represent ordinarily least squares estimates of the linear
+% relationship between racer's finishing and personal best times.
+
+% One can see in Figure \ref{fig:scatter} that there is a clear
+% relationship between a runner's personal best time and her performance
+% in the Olympics.  While there is a significant amount of noise in the
+% relationship, the Hahner twins' dots are well above the regression's
+% fitted line notwithstanding this point.  They finished the Rio
+% marathon much slower than they should have, per their pre-Olympic
+% personal best times.  In fact, if we sort studentized residual, we can
+% see in Figure \ref{fig:studentizedresiduals} that the Hahner twins are
+% in the tail-end of the residual distribution.  Anna Hahner's residual
+% deviation is greater than 91.7\% of the runners who completed the
+% marathon while Lisa Hahner's deviation is greater than 86.5\%.  Hence,
+% the twins appear to have finished at a much slower pace than expected,
+% which is what we might expect if they were coordinating their runs.
+% The North Korean Kim twins are also in the tail of the residual
+% distribution---but the left tail.  These twins finished much faster
+% than expected.
+
+% \section*{Did Anna and Lisa intentionally finish together?}
+
+% Anna and Lisa finished the marathon at an unusually slow rate, which is what we might expect if the twins intentionally slowed their pace to finish simultaneously.
+However, the observation of a slow finish does not necessarily imply that it was an intentional result .
+It only suggests that Kurschilgen's claim regarding this issue holds up in the data.
+
+% Perhaps it would be easier to refute the claim if we could show that the Twins's back-to-back finish was not such an unusual event.
+Given that Anna and Lisa were so close in their personal best times, it may seem reasonable to expect that they would also be close in their finishing times.
+Hence, one way to determine if the twins's finish was intentional is to quantify the likelihood that such a close finish would occur by chance given the differences in their ability.
+
+% We can attempt to quantify this likelihood by looking at how differences in the personal best time of two runners translate into differences in Olympic times .
+If Anna and Lisa's finish was unintentional we'd expect that such a close finish would be unlikely given their difference in personal best times.
+Yet, how does the difference in personal best times relate to the difference in finishing times?
+
+% One way we can estimate this relationship is to look at all of the differences that we observe in the marathon.
+We can take the difference in personal best time for every combination of runners in the data and look at the difference in their final times.
+We should expect that compared to the difference in their personal best times, the difference in Anna and Lisa's final times should be commensurate with the conditional difference of all other runners.
+%This needs a better explanation.
+
+% Therefore, we find all combinations of runners in the marathon and calculate two quantities: 1) the difference in their personal best time and 2) the difference in their final time.
+We plot the relationship between the two quantities in Figure \ref{fig:diffdiffscatter}.
+All 8,778 dyadic combinations of runners are displayed by grey points in the scatter plot.
+The solid black line indicates a one-to-one relationship, where every point would fall if the difference in final times were systematically equivalent to the difference in personal bests.
+
+% % \begin{figure}[!ht]
+% %  \caption{Relationship between Difference in Personal Best and Difference in Result}
+% %  \label{fig:diffdiffscatter}
+% %  \centering
+% %  \begin{subfigure}{.5\textwidth}
+% %  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
+% %  \end{subfigure}
+% % \end{figure}
+
+% We can see from the yellow dot that the Hahner Twins are just below this one-to-one line.
+Hence, the twins' final times were closer to each other than their personal best times; but not by much.
+In fact, this might be an expected difference if we expected there to be a one-to-one relationship.
+However, although there is a clear relationship between the difference in final time and the difference in personal best, it is not a strong one-to-one relationship.
+We've added a dotted smoother to display a cubic splines estimate of the conditional averages.
+While the Hahner twins may be close to the one-to-one line, they are much further away from the average set of runners.
+Hence, we would expect for any pair of runners, the difference in their final results should be much greater than their difference in their personal best.
+However, the Hahner twins do not fit this expectation.
+Yet, how different from this expectation are they?
+
+% One way we can quantify the degree to which the twins are outliers is by comparing their difference in difference to every other pair of runner's difference-in-differences.
+We calculate the difference-in-differences for each pair of runners by simply subtracting their difference in result (y-axis) from their difference in personal best (x-axis).
+Therefore positive difference-in-differences mean that the runners ended up further apart than their difference in personal bests.
+Negative difference-in-differences mean that the runners end up closer together.
+Hence, the Hahner twins have a difference-in-difference of -114 seconds.
+They finished 114 seconds closer together than the distance between their personal bests.
+
+% We arrange all 8,778 difference-in-differences from small to large in the first plot in Figure \ref{fig:diffdiff}.
+The Hahner twins are denoted by the yellow point.
+Using this plot we can visualize where the twins fall within the distribution of difference-in-differences.
+It is clear that they are closer to the lower tail of the distribution.
+However, compared to all other difference-in-differences, their result is not that extreme.
+Nearly 21\% of the difference-in-dfferences were less than the that of the twins'.
+Hence, we'd expect at least one in five pairs of runners to have a reduction in difference at least as great as the twins.
+
+% However, if we limit the observations to only those runners who had personal best times that were at least as close to each other as the Hahner twins, we get a different perspective.
+Among those runners,  the Hahner twins reduced the distance between them to the greatest extent.
+There was no other pair of runners whose reduced the difference in the final time relative to their difference in personal bests more than them.
+
+% \begin{figure}[!ht]
+%  \caption{How Hahner Twins Rank in Difference-in-Differences}
+%  \label{fig:diffdiff}
+%  \begin{subfigure}{.5\textwidth}
+%  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_1.pdf}
+%  \end{subfigure}
+%  \begin{subfigure}{.5\textwidth}
+%  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_2.pdf}
+%  \end{subfigure}
+% \end{figure}
+
 \section*{Conclusion}

### 2017-02-03 2-simulate.R

@@ -230,18 +230,18 @@ max_r <- max(hahner_df$rank_diff, na.rm = T)
 
 pdf("plots/simulated_time.pdf", height = 5, width = 5)
 ggplot(hahner_df, aes(time_diff)) + 
-  geom_histogram(binwidth = 30, colour = "black", fill = NA) +
-  xlab("\nDifference in seconds") +
-  ylab("Count\n") +
-  xlim(0, max_t) + 
-#  ggtitle("\nSimulated final time") +
-  theme_bw()
+    geom_histogram(binwidth = 30, colour = "black", fill = NA) +
+    xlab("\nFinishing time difference, in seconds") +
+    ylab("Count\n") +
+    xlim(0, max_t) + 
+    ##  ggtitle("\nSimulated final time") +
+    theme_bw()
 dev.off()
 
 pdf("plots/simulated_rank.pdf", height = 5, width = 5)
 ggplot(hahner_df, aes(rank_diff)) + 
   geom_histogram(binwidth = 1, colour = "black", fill = NA) +
-  xlab("\nDifference in rank") +
+  xlab("\nRank difference") +
   ylab("Count\n") +
   xlim(0, max_r) + 
 #  ggtitle("\nSimulated final rank") +

### 2017-02-03 paper/paper.tex
paper updates on friday later

@@
 
-We now consider the following: if we take into consideration the
-Hahner twin's similarities as runners as well as some natural
-variation in finishing a marathon, what is the probability that the
-twins finished at roughly the same time and/or sequentially?  To
-answer this question we need to know the counterfactual distribution
-of potential marathon finishing times that would have occurred had the
-Hahner twins independently and repeatedly run the Rio marathon over
-again.  Holding the twin's abilities constant, this would establish
-the set of potential outcomes that would have occurred as result of
-natural variation in the finishing the marathon.  We could use this
-distribution of outcomes to determine the likelihood that a
-simultaneous finish by the Hanhers, or at least a near-simultaneous
-finish, occurred by chance alone.  If the twins rarely finish the
+We now consider the following question: if we take into consideration
+the Hahner twin's similarities as runners as well as natural variation
+in marathon finishing times, what is the probability that the twins
+would finish the Rio race at roughly the same time and/or
+sequentially?  To answer this question we need to know the
+counterfactual distribution of potential marathon finishing times that
+would have occurred had the Hahner twins independently (in particular,
+of each other) and repeatedly run the Rio marathon, holding constant
+marathon conditions, the abilities of other runners, and so forth.
+Access to such a distribution would establish the set of potential
+outcomes that could have occurred on August 14, 2016, and we could in
+principle use this distribution to determine the likelihood that a
+simultaneous finish by the Hahners, or at least a near-simultaneous
+finish, occurred by chance alone.  If these twins rarely finish the
 marathon together in such a counterfactual world, then one might be
@@
 participants, it is not possible to rerun the women's Olympic marathon
-in Rio to establish a counterfactual distribution of potential race
-outcomes for the Hahner twins.  However, we can attempt to simulate
-this distribution by estimating the distribution of every other
-runner's finishing time, conditioned on prior abilities, and drawing
-from that distribution in order to observe the likelihood that Lisa
-and Anna Hahner finish simultaneously.
+in Rio to establish a distribution of potential race outcomes for the
+Hahner twins.  However, we can attempt to simulate this distribution
+by estimating the distribution of every other runner's finishing time,
+conditional on prior abilities, and drawing from that distribution in
+order to observe the likelihood that, for example, Lisa and Anna
+Hahner finished simultaneously.
 
-To estimate the conditional distribution of Rio runner's result, we
+To estimate the conditional distribution of each Rio final results, we
 assume that each runner's marathon time \Yi\ is distributed normally
@@
 %%
-where $N\left(\cdot,\cdot\right)$ denotes a normal distribution. We
-estimate $\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least
-squares and our set of 133 finishing runners.  For a simulated
-marathon, we then draw a runner's time from an estimated distribution,
-conditioned on the runner's personal best outcome.  Once a race is
-simulated, we locate Anna and Lisa Hahner and record both the time
-between their finishes and the difference in their ranks.  We then
-simulate a new race---drawing a new set of finishing times---and
-record the same quantities.  In particular, the steps of the
-simulation are as follows.
+where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.
+Personal best times are all prior to the Rio marathon.  We estimate
+$\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least squares and
+all finishing runners except for the twins and triplets.  For a
+simulated marathon, we then draw a runner's time from an estimated
+distribution, conditioned on the runner's personal best marathon time.
+Once a race is simulated for all runners, twins and triplets included,
+we locate Anna and Lisa Hahner in our set of simulated results and
+record both the time between their simulated finishes and the
+difference in their simulated ranks.  We then simulate a new
+race---drawing a new set of finishing times---and record the same
+quantities.  In particular, the steps of the simulation are as
+follows.
 
@@
 \item Ignoring twins and triplets, estimate with least squares a
-  linear model that predicts a runner's finishing time ($Y_i$) based
-  on her pre-Olympic personal best time ($X_i$) and her best time
-  squared.
-\item Extract coefficient vector, estimated covariance matrix, and
-  estimated regression variance from this model.
+  linear model that predicts a runner's finishing time $Y_i$ based on
+  her pre-Olympic personal best time $X_i$ and her best time squared.
+\item Extract the resulting coefficient vector, estimated covariance
+  matrix, and estimated regression variance from this model.
 \item For each simulated rate, draw intercept and slope estimates
@@
   respectively, from a multivariate normal distribution with mean
-  equal to the estimated coefficient vector and covariance equal to
-  the estimated covariance matrix.
-\item For each runner, draw an error $\tilde{e}$ from a normal
-  distribution with mean zero and a standard deviation equal to the
-  standard deviation of the original regression model's residuals.
-\item For each runner, predict the final result by combining the randomly
-  generated beta coefficients and error terms, $\tilde{y} = \tilde{\beta}_0 +
-  \tilde{\beta}_1\,X_i + \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
-\item Eliminate each runner from the race with a probability equal to
-  the observed fraction of marathoners who did not finish the
+  equal to the previously estimated coefficient vector and covariance
+  equal to the previously estimated covariance matrix.
+\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
+  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
+  normal distribution with mean zero and a standard deviation equal to
+  the standard deviation of the original regression model's residuals.
+\item For each runner, predict her final result by combining the
+  randomly generated beta coefficients and error terms,
+  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
+  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
+\item Eliminate each runner from the simulated race with a probability
+  equal to the observed fraction of marathoners who did not finish the
   marathon.
@@
 
-Before considering results, DNF biz...
+%% Before considering results, DNF biz...
 
-Figure \ref{fig:simdiff} contains two histograms of differences
-between Anna and Lisa's finishes.  Figure
-\ref{fig:simulatedfinishtimes} shows the distribution of absolute
-differences in time where differences are grouped in 30 second bins
-along the horizontal axis of the histogram.  The count for each bin is
-denoted by the vertical length of the bars.  Figure
-\ref{fig:simulatedranks} shows the distribution of the absolute
-differences in Hahner twin rankings.  Differences are grouped as
-single units ranging from no runner between the twins to nearly 120
-runners between them.
+From our simulations we can generate intervals which describe the
+extent of the variability in racer finishing times.  In 95\% of the
+simulations in which Anna Hahner completed the marathon, her finishing
+time was between 9,480 seconds and 10,703 seconds.  This is consistent
+with Anna's observed finishing time, which was 9,932 seconds.  In
+fact, our simulations estimate that Anna should have finished the Rio
+marathon in 10,092 seconds, which is slightly more than two and
+one-half minutes slower than her actual time.  Lisa Hahner's
+corresponding interval is 9,427 to 10,619 with an observed finishing
+time of 9,933 seconds.
+
+The bottom line here is that our simulated marathon finishes are
+consistent with the Hahner twins' finishing times.  And note that the
+regression model underlying the simulations was estimated without the
+Hahner twins (and same for the Kim twins and Luik triplets).
+According to the our simulations, then, neither Hahner twin ran
+appreciably slowly, conditional on personal best times prior to the
+Rio Olympics.
+
+% 95% simulated interval for Anna Hahner finishing time in seconds: 9480 10703 
+% 95% simulated interval for Lisa Hahner finishing time in seconds: 9427 10619
+
+% > df$FINAL[df$ATHLETE == "Anna Hahner"]
+% [1] 9932
+% > df$FINAL[df$ATHLETE == "Lisa Hahner"]
+% [1] 9933
+
+With an eye on the matter of simultaneous finishing, Figure
+\ref{fig:simdiff} contains two histograms based on our simulated race
+results. Figure \ref{fig:simulatedfinishtimes} is a histogram which
+shows the distribution of absolute differences in Hahner twin
+finishing time where differences are grouped in 30 second bins; counts
+for the various bins are denoted by the vertical lengths of the bars.
+Figure \ref{fig:simulatedranks} is similar but depicts the
+distribution of the absolute differences in Hahner twin rankings.
+Differences are grouped as single units ranging from no runner between
+the twins to nearly 120 runners between them.
 
@@
 
-The results plotted in the histograms question the credibility of Anna
-and Lisa's story.  The results suggest that a simultaneous finish
-would be very rare if Anna and Lisa had run independently of each
-other.  For example, in less than 300 of the 10,000 simulated races
-did Anna and Lisa finished within 30 seconds of each other .  And they
-finished in consecutive rank in less than 175 of the 10,000 races.
-Therefore, the close finish that we observed, where Anna and Lisa
-crossed the finish line one-after-the-other, would be very unlikely if
-we were to believe the twins raced independently of each other.
+The results plotted in the histograms raise questions about the
+credibility of Anna and Lisa Hahner's story.  In particular, the
+histogram suggest that a simultaneous finish would be very rare if
+Anna and Lisa had run independently of each other.  For example, in
+fewer than 300 of the 10,000 simulated races did Anna and Lisa
+finished within 30 seconds of each other.  Moreover, they finished in
+consecutive rank in fewer than 175 of the 10,000 races.  Therefore,
+the close finish that was observed, where Anna and Lisa crossed the
+finish line one-after-the-other, would be highly unlikely if the two
+German twins had raced independently of each other.
 
-The results holds even if you replace their personal best time with
-their time halfway through the race.  While this reduces the variation
-in the predicted outcome of each runner and, therefore, reduces the
-expected distance between Anna and Lisa, it is still quite rare for
+Earlier we noted that personal best times might arguably be poor
+measures of runner abilities.  With this in mind, we repeated our
+simulations but replaced personal best times with half marathon
+splits.  See Figure \ref{fig:simdiffhalf}.  While this change reduces
+the variation in the predicted outcome of each runner and therefore,
+reduces the expected distance between Anna and Lisa Hahner, even with
+half marathon split as a measure of ability it is still quite rare for
 Anna and Lisa to finish simultaneously or consecutively.

### 2017-02-03 2-simulate.R

@@ -91,19 +91,19 @@ df2b <- df2 %>% filter(name_i != name_j, !is.na(diff_in_diff)) %>% arrange(diff_
 
 pdf("plots/diff_in_diff_scatter_plot.pdf", height = 5,  width = 5)
 ggplot(aes(y=abs(result_diff), x=abs(pb_diff)), data = df2b) + 
-  geom_point(size = .5, colour = "gray", alpha = .5) + 
-  geom_point(data = df2b[!is.na(df2b$twins),], aes( col = twins)) +
-  geom_smooth( colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black") + 
-  geom_abline(intercept = 0, slope =1) +
-  #geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
-  #geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
-  coord_fixed(xlim = c(0,3500), ylim = c(0,3500)) +
-  ylab("Absolute difference Olympic result, in seconds\n") + 
-  xlab("\nAbsolute difference in personal best, in seconds") + 
-  #ggtitle("Among all dyadic combinations") +
-  scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
-  theme_bw() +
-  theme(legend.position = c(.8,.2))
+    geom_point(size = .5, colour = "gray", alpha = .5) + 
+    geom_point(data = df2b[!is.na(df2b$twins),], aes( col = twins)) +
+    ## geom_smooth( colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black") +  # removed smoother because points are not independent
+    geom_abline(intercept = 0, slope =1) +
+    ## geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
+    ## geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
+    coord_fixed(xlim = c(0,3500), ylim = c(0,3500)) +
+    ylab("Absolute difference Olympic result, in seconds\n") + 
+    xlab("\nAbsolute difference in personal best, in seconds") + 
+    ## ggtitle("Among all dyadic combinations") +
+    scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
+    theme_bw() +
+    theme(legend.position = c(.8,.2))
[…]

### 2017-02-03 simhalf.R

@@ -68,7 +68,7 @@ hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_dif
 pdf("plots/simulated_time_half.pdf", height = 5, width = 5)
 p <- ggplot(hahner_df, aes(time_diff)) + 
   geom_histogram(binwidth = 30, colour = "black", fill = NA) +
-  xlab("\nDifference in seconds") +
+  xlab("\nFinishing time difference, in seconds") +
   ylab("Count\n") +
   xlim(0, max_t) + 
   #  ggtitle("\nSimulated final time") +

### 2017-02-03 paper/paper.tex
paper stuff

@@
   \centering
-  \caption{Distribution of simulated results for the Hahner twins (using halfway split as predictor)}
+  \caption{Distribution of simulated results for the Hahner twins
+    based on half marathon split}
   \label{fig:simdiffhalf} 
@@
 \end{figure}
+
+An important point regarding our simulations is the way that we
+handled DNFs.  As our description above indicates, we assumed that DNF
+probabilities are the same for all runners and that the likelihood of
+a DNF is a not function of a runner's anticipated marathon finishing
+time.  The rug marks in Figure \ref{fig:45degreeplot} suggest that
+runners with better personal best times may be more likely to DNF than
+other runners, \emph{ceteris paribus}.  We suspect that this occurs
+because some better runners may expend excessive energy trying to
+achieve a good result in the marathon and in so doing injure or
+exhaust themselves; lesser runners, in contrast, may be content to
+finish respectably.  Regardless of the validity of this conjecture,
+Figure \ref{fig:45degreeplot} shows that the Hahner twins are
+representative of sort of runners who DNFed in Rio.  Since our
+simulated Hahner statistics are conditioned on both Hahner twins
+finishing, it follows that they are conservative.  The fact that both
+women finished the marathon was notable in and of itself, and by
+discounting the possibility of a Hahner DNF we are giving the benefit
+of the doubt to the Hahner twins.
+
+\section*{Conclusion}
+
+Turning back to the motivation of this article, we have offered two
+perspectives on Anna and Lisa Hahner'
 
@@
 
-\section*{Conclusion}
-
-\newpage
 \section*{Further reading}

### 2017-02-03 paper/paper.tex
for david

@@
 
-%% Before considering results, DNF biz...
+David---comments here?    Or above?  This is not from the simulation
+--- so perhaps move to top of section??
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[keepaspectratio]{studentized_residuals.pdf}
+    \caption{Studentized residuals}
+    \label{fig:studentizedresiduals}
+  \end{figure}
 
@@
 
-Turning back to the motivation of this article, we have offered two
-perspectives on Anna and Lisa Hahner'
+Looking back at the original motivation of this article, we have
+offered two perspectives on Anna and Lisa Hahner's near-simultaneous
+finish in the 2016 women's Olympic marathon in Rio.  The German twins
+were accused of deliberately slowing down and finishing next to each
+other so as to generate media attention.  They denied this, not
+entirely surprisingly,
+
+We have offered two perspectives on the marathon, one based on
+visualizations and simple calculations and a second that draws on a
+simulation.  Both perspectives have the same implications, and they
+are as follows.  In a global sense, the Hahner twins did not slow down
+appreciably during the marathon.  Their times were not fast, but they
+were within reason for runners of the Hahners' abilities.  Locally,
+though, we find that the Hahner's finish was in fact contrived.  Their
+finish---one second between the two women---was an extremely low
+probability even.  Compared to their differences as measured by
+personal best times prior to Rio, the Hahner twins difference in
+finishing times was unusually compressed.  This is evidence that their
+finishing had elements of intentionality---perhaps at the last minute
+but intentionality nonetheless.
+
+Now what... David?
 

### 2017-02-03 2-simulate.R

@@ -116,7 +116,7 @@ ggplot(df2b, aes(PERCENTILE,diff_in_diff)) +
   geom_text(data = df2b[df2b$twins == "Kim Twins",], aes(label = paste0(formatC(round(PERCENTILE,1), format = "f", flag = "0", digits = 1 ), "%")), nudge_y = 200, colour = "red") +
   ggtitle("Distribution of Difference in Differences\n(All Dyads)") +
   ylab("Difference in Difference (sec)\n") +
-  xlab("\nPrecentile") + 
+  xlab("\nPercentile") + 
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +
   theme(legend.position = c(.85,.15))
@@ -132,7 +132,7 @@ ggplot(df2c, aes(PERCENTILE,diff_in_diff)) +
   geom_text(data = df2c[df2c$twins == "Kim Twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_y = 100, colour = "red") +
   ggtitle("Distribution of Difference in Differences\n(Diads where difference in PB < Hahner's)") +
   ylab("Difference in Difference (sec)\n") +
-  xlab("\nPrecentile") + 
+  xlab("\nPercentile") + 
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +
   theme(legend.position = c(.85,.15))
@@ -153,7 +153,7 @@ ggplot(dta, aes(PERCENTILE,STUD_RESID)) +
   geom_text(data = dta[dta$TWINS == "Kim Twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_x = 10, colour = "red") +
 ##  ggtitle("Distribution of Studentized Residuals") +
   ylab("Studentized residual\n") +
-  xlab("\nPrecentile") + 
+  xlab("\nPercentile") + 
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +
   theme(legend.position = c(.8,.2))

### 2017-02-03 paper/paper.tex
feb 3 edits

@@
 
-Figure \ref{fig:scatter} contains two plots, both of which describe
+Figure \ref{fig:scatter} and Figure \ref{fig:scatterhalf} contain two plots, both of which describe
 how finishing times from the Rio Olympic women's marathon varied as a
-function of athletes' personal best times and their Rio half marathon
+function of athletes' personal best times and and as function of their Rio half marathon
 split times.  The points in the plots are colored by twin/triplet
@@
 lines.  We treat an athlete's prior personal best marathon time as a
-measure of the athlete's underlying marathon talent, and we interpret
+measure of the athlete's underlying marathon talent (seen in Figure \ref{fig:scatter}), and we interpret
 an athlete's half marathon split as a measure of the athlete's skill
 at marathon running on August 14, 2016, the date of the women's
-marathon in Rio.
+marathon in Rio (seen in Figure \ref{fig:scatterhalf}).
 
 \begin{figure}[!ht]
-  \caption{Olympic results and measures of marathon ability}
+  \caption{Olympic results as a function of ability measured by personal best}
   \label{fig:scatter}
@@
     \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
-    \caption{Finishing and personal best times}
+    \caption{} 
     \label{fig:45degreeplot}
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
-    \caption{Finishing and half marathon split times}
-    \label{fig:45degreeplot_half}
+    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals.pdf}
+    \caption{}
+    \label{fig:studentizedresiduals}
   \end{subfigure}
@@
 
-Figure \ref{fig:45degreeplot_half} includes a solid 45-degree line as
+\begin{figure}[!ht]
+  \caption{Olympic results as a function of ability measured by half marathon split}
+  \label{fig:scatterhalf}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
+    \caption{}
+    \label{fig:45degreeplot_half}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half.pdf}
+    \caption{}
+    \label{fig:studentizedresidualshalf}
+  \end{subfigure}
+\end{figure}
+
+Figure \ref{fig:45degreeplot} includes a solid 45-degree line as
 the axes in the figure correspond to full marathons; in addition, the
@@
  \caption{Differences in personal best times and finishing times}
+ \centering
+ \begin{subfigure}{.49\textwidth}
+ \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
   \label{fig:diffdiffscatter}
- \centering
- \begin{subfigure}{.5\textwidth}
- \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
+ \end{subfigure}
+  \begin{subfigure}{.49\textwidth}
+ \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot_half.pdf}
+  \label{fig:diffdiffscatterhalf}
  \end{subfigure}
@@
 difference in personal best times, most marathoners did not have close
-finishing times like the Hahnes.  Some pairs of runners with around
+finishing times like the Hahners.  Some pairs of runners with around
 115 second personal best differences had finishing time differences of
@@
 --- so perhaps move to top of section??
-
-\begin{figure}[!ht]
-    \centering
-    \includegraphics[keepaspectratio]{studentized_residuals.pdf}
-    \caption{Studentized residuals}
-    \label{fig:studentizedresiduals}
-  \end{figure}
 

### 2017-02-03 1-create.R

@@ -54,9 +54,9 @@ splits <- full_join(splits_dist,splits_rank, by = "ATHLETE")
 
 df <- start %>% full_join(splits, by = "ATHLETE") %>% full_join(result, by = "ATHLETE")
 
-df <- df %>% mutate(TWINS = ifelse(BIB %in% c(748, 749), "Hahner Twins",
-                                   ifelse(BIB %in% c(1137, 1138), "Kim Twins",
-                                          ifelse(BIB %in% c(633:635), "Luik Triplets", NA))))
+df <- df %>% mutate(TWINS = ifelse(BIB %in% c(748, 749), "Hahner twins",
+                                   ifelse(BIB %in% c(1137, 1138), "Kim twins",
+                                          ifelse(BIB %in% c(633:635), "Luik triplets", NA))))
 
 nms <- c("LANE","BIB","ATHLETE","COUNTRY","TWINS","SB","PB","SPLIT_5K", "SPLIT_10K","SPLIT_15K", "SPLIT_20K", "SPLIT_HALF", "SPLIT_25K", "SPLIT_30K", "SPLIT_35K", "SPLIT_40K", "FINAL", "RK_PB", "RK_5K", "RK_10K","RK_15K","RK_20K","RK_HALF","RK_25K","RK_30K","RK_35K","RK_40K","RK_FINAL") 
 

### 2017-02-03 2-simulate.R

@@ -21,8 +21,9 @@ ggplot(aes(y=FINAL, x=PB), data = df) +
     geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black", se = FALSE) + 
     geom_abline(intercept = 0, slope =1) +
     geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
-    geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
-    coord_fixed(xlim = c(8000,12500), ylim = c(8000, 12500)) +
+    geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik triplets",], aes(x = PB), colour = "blue") +
+    xlim(8000,12500) +
+    ylim(8000, 12500) +
     ylab("Olympic result, in seconds\n") + 
     xlab("\nPersonal best, in seconds") + 
     ##  ggtitle("Distribution of Y | X") +
@@ -31,8 +32,6 @@ ggplot(aes(y=FINAL, x=PB), data = df) +
     theme_bw() +
     theme(legend.position = c(.8,.2))
 dev.off()
-
-
 
 
 ## make scatter plot for half marathon split
@@ -44,7 +43,7 @@ ggplot(aes(y = FINAL, x = SPLIT_HALF), data = df2use) +
     geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black", se = FALSE) + 
     ## geom_abline(intercept = 0, slope = 1) +
     geom_rug(data = df2use[is.na(df2use$FINAL),], aes(x = SPLIT_HALF), colour = "gray") +
-    geom_rug(data = df2use[is.na(df2use$FINAL) & df2use$TWINS == "Luik Triplets",], aes(x = PB), colour = "blue") +
+    geom_rug(data = df2use[is.na(df2use$FINAL) & df2use$TWINS == "Luik triplets",], aes(x = PB), colour = "blue") +
     ##    coord_fixed(xlim = c(4000, 6000), ylim = c(8500, 12500)) +
     ylab("Olympic result, in seconds\n") + 
     xlab("\nHalf marathon split, in seconds") + 
[…]

### 2017-02-03 2-simulate.R
axis

@@ -93,8 +93,8 @@ ggplot(dta, aes(STUD_RESID, PERCENTILE)) +
   geom_text(data = dta[dta$TWINS == "Luik triplets" & !is.na(dta$TWINS),][1,], aes(label = print_pct(PERCENTILE)), nudge_x = -.5, colour = "blue") +
   geom_text(data = dta[dta$TWINS == "Luik triplets" & !is.na(dta$TWINS),][2,], aes(label = print_pct(PERCENTILE)), nudge_x = .5, colour = "blue") +
   ##  ggtitle("Distribution of Studentized Residuals") +
-  ylab("Studentized residual\n") +
-  xlab("\nPercentile") + 
+  ylab("Percentile\n") +
+  xlab("\nStudentized residual") + 
   scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
   theme_bw() +
   theme(legend.position = c(.8,.2))

### 2017-02-03 paper/paper.tex
correct figure

@@
 
-\title{\Large{All in the family:\\Anna and Lisa Hahner's finishing
-    times in\\the 2016 women's Olympic
-  marathon}}\author{David Cottrell\thanks{Postdoctoral Research
-  Fellow, Program in Quantitative Social Science, Dartmouth College,
-    6108 Silsby Hall, Hanover, NH
-    03755 (\texttt{david.cottrell@dartmouth.edu}).} \and Michael C.\
-  Herron\thanks{Visiting Scholar, Hertie School of Governance, Berlin,
-    Germany, and Professor of Government, Dartmouth College, 6108
-    Silsby Hall, Hanover, NH 03755
+\title{\Large{All in the family:\\German twin finishing times in\\the
+    2016 women's Olympic marathon}}\author{David
+  Cottrell\thanks{Postdoctoral Research Fellow, Program in
+    Quantitative Social Science, Dartmouth College, 6108 Silsby Hall,
+    Hanover, NH 03755 (\texttt{david.cottrell@dartmouth.edu}).} \and
+  Michael C.\ Herron\thanks{Visiting Scholar, Hertie School of
+    Governance, Berlin, Germany, and Professor of Government,
+    Dartmouth College, 6108 Silsby Hall, Hanover, NH 03755
     (\texttt{michael.c.herron@dartmouth.edu}).}}
@@
 As such, our investigation of the charges that Kurschilgen offered
-will distinguish between the idea of slow finish versus a simultaneous
-finish.
+will distinguish between a slow finish versus a simultaneous finish.
 
@@
 Overall, our visualizations suggest that the Hahner twins' pace in the
-marathon was slow, but not excessively so, but that their simultaneous
-finish was quite unusual given the twins' differences in abilities
-(and similarly for the Kim twins).  We then turn to a regression-based
-simulation of the marathon, which we develop based on relationships
-(excluding the Hahner and Kim twins and Luik triplets) between known
-personal best times and observed Rio finishing times.  Our simulations
-show that the Hahner twins finished suspiciously close to each other
-given the disparty between their personal best times.  Finally, we
-return to Kurschilgen claims about the Hahners and offer our thoughts
-about their validity.
+marathon was slow, albeit not excessively so, but that their
+simultaneous finish was quite unusual given the twins' differences in
+abilities (and similarly for the Kim twins).  We then turn to a
+regression-based simulation of the marathon, which we develop based on
+relationships across the Olympic marathoners between personal best
+times and observed Rio finishing times.  Our simulations show that the
+Hahner twins did not run appreciably slowly yet finished suspiciously
+close to each other given the disparty between their personal best
+times.  We return in our conclusion to Kurschilgen's claims about the
+Hahners and offer our thoughts about their validity.
 
@@
  \caption{Differences in personal best times and finishing times}
+  \label{fig:diffdiff}
  \centering
@@
  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
+ \caption{}
   \label{fig:diffdiffscatter}
@@
  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot_half.pdf}
+ \caption{}
   \label{fig:diffdiffscatterhalf}

### 2017-02-03 paper/paper.tex
ready for david final editing

@@
 
-At 9:30am on August 14, 2016, the Women's Olympic marathon kicked off
+At 9:30am on August 14, 2016, the women's Olympic marathon kicked off
 in Rio de Janeiro, Brazil, when 156 runners from 80 countries across
@@
 
-The media quickly picked up on the Hanher story as an image of the
+The media quickly picked up the Hanher story as an image of the
 beaming twins finishing hand-in-hand captured a public audience. While
@@
 expressions at the finish were portrayed as a bit contrived---smiling
-like ``Honigkuchenpferde,'' cookies in the shape of a horse---and the
-sports director of the German Athletics Federation, Thomas
-Kurschilgen, stirred up controversy when he suggested that the
-Hahners' photo-finish was no coincidence.  Kurschilgen averred that
-the twins slowed down so as to finish simultaneously and create a
-spectacle which would ``generate media attention.''  Kurschilgen
-justified his charge with the fact that the twins ran in the Rio
-marathon at least 18 minutes slower their personal best times prior to
-the
-Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
-Not surprisingly, Kurschilgen's accusations were denied by the Hahner
-twins, who claimed that their simultaneous finish was simply an
-unintended coincidence.
+like ``Honigkuchenpferde,'' cookies in the shape of a horse, was the
+description offered by one editorialist---and the sports director of
+the German Athletics Federation, Thomas Kurschilgen, stirred up
+controversy when he suggested that the Hahners' photo-finish was no
+coincidence.  Kurschilgen averred that the twins slowed down so as to
+finish simultaneously and create a spectacle which would ``generate
+media attention.''  Kurschilgen justified his charge with the fact
+that the twins ran the Rio marathon over 18 minutes slower their
+personal best times prior to the Olympics.  Not surprisingly,
+Kurschilgen's accusations were denied by the Hahner twins, who claimed
+that their simultaneous finish was simply an unintended coincidence.
 
@@
 hours---but we nonetheless want to know what the data from the Olympic
-marathon tell us.  Was the Hahner finish in the 2016 women's
-Olympic marathon a lovely coincidence or something else?
+marathon tell us.
+
+%% Was the Hahner finish in the 2016 women's Olympic marathon a lovely coincidence or something else?
 
 Among female Olympic marathoners, the Hahner twins were not alone in
-their familial ties.  The marathon also featured twins from North
-Korea, Kim Hye-song and Kim Hye-gyong, who posted identical times and
-finished 10th and 11th in the race, respectively. The Kim finish,
-unlike the Hahner finish, appears devoid of post-race
+sharing familial ties.  The Rio marathon also featured twins from
+North Korea, Kim Hye-song and Kim Hye-gyong, who posted identical
+times and finished 10th and 11th in the race, respectively. The Kim
+finish, unlike the Hahner finish, appears devoid of post-race
 controversy. Moreover, three triplets from Estonia competed in the Rio
@@
 expressed ire at the Hahners had they finished in 1st and 2nd place in
-Rio, hand-in-hand with wide grins, but of course we do not know this.
-As such, our investigation of the charges that Kurschilgen offered
-will distinguish between a slow finish versus a simultaneous finish.
+Rio, hand-in-hand with wide grins, but of course we cannot be certain
+of this.  Thus, our investigation of the charges that Kurschilgen
+offered will distinguish between a slow finish versus a simultaneous
+finish.
 
@@
 describe various features of the Hahner twin's results, and an
-important element in our visualizations is the difference between a
-runner's Rio time and her prior personal best time in the marathon.
-Overall, our visualizations suggest that the Hahner twins' pace in the
-marathon was slow, albeit not excessively so, but that their
-simultaneous finish was quite unusual given the twins' differences in
-abilities (and similarly for the Kim twins).  We then turn to a
-regression-based simulation of the marathon, which we develop based on
-relationships across the Olympic marathoners between personal best
-times and observed Rio finishing times.  Our simulations show that the
-Hahner twins did not run appreciably slowly yet finished suspiciously
-close to each other given the disparty between their personal best
-times.  We return in our conclusion to Kurschilgen's claims about the
-Hahners and offer our thoughts about their validity.
+important element in our visualizations is the difference between
+runners' Rio times and their prior personal best marathon times.  Our
+visualizations suggest that the Hahner twin's pace in the marathon was
+slow, albeit not excessively so, but that their simultaneous finish
+was quite unusual given the twins' differences in abilities (and
+similarly for the Kim twins).  We then turn to a regression-based
+simulation of the marathon, which we develop based on relationships
+across the Olympic marathoners between personal best times and
+observed Rio finishing times.  Our simulations reinforce what we
+observed in visualization, namely, that the Hahner twins did not run
+appreciably slowly yet finished suspiciously close to each other.  We
+return in our conclusion to Kurschilgen's claims about the Hahners and
+offer thoughts about their validity.
 
@@
 
-Figure \ref{fig:scatter} and Figure \ref{fig:scatterhalf} contain two plots, both of which describe
-how finishing times from the Rio Olympic women's marathon varied as a
-function of athletes' personal best times and and as function of their Rio half marathon
-split times.  The points in the plots are colored by twin/triplet
-status, and both plots contain second-order polynomial regression
-lines.  We treat an athlete's prior personal best marathon time as a
-measure of the athlete's underlying marathon talent (seen in Figure \ref{fig:scatter}), and we interpret
-an athlete's half marathon split as a measure of the athlete's skill
-at marathon running on August 14, 2016, the date of the women's
-marathon in Rio (seen in Figure \ref{fig:scatterhalf}).
+One way that we might assess whether an athlete's Rio finish was
+unusual---or, say, whether two athletes' finishes were jointly
+unusual---is by comparing observed finishing times on August 14 to a
+measure of underlying marathon talent.  For the latter we use prior
+personal best marathon times.  This is a natural measure of marathon
+talent, we believe, but it is not entirely free of complications.  One
+limitation of using personal best times as indicators of underlying
+marathon skill is that these times are potentially confounded by the
+marathon courses at which they were set; some courses, like the Berlin
+marathon, are known for relatively fast times.  In addition, personal
+best times may be confounded by conditions, like weather, that varied
+across races.  Finally, personal best times may not capture Olympic
+raceday idiosyncrasies that could affect individual runners.  For
+example, a runner could have woken up with a minor cold on the morning
+of August 14, 2016.  With these concerns in mind, we use half marathon
+split times from the Olympic marathon as a secondary measure of
+athletic skill in marathoning.  Our two measures of marathon talent
+are featured in the visualizations below.
+
+Figure \ref{fig:scatter} contains two plots, both of which describe
+how finishing times from the Rio women's marathon varied as a function
+of athletes' personal best times and as function of their half
+marathon split times.  The points in the plots are colored by
+twin/triplet status, and both plots contain second-order polynomial
+regression lines.  Rug marks in both plot indicate times (either
+personal best or half marathon split) of runners who earned a DNF
+result.
 
 \begin{figure}[!ht]
-  \caption{Olympic results as a function of ability measured by personal best}
+  \centering
+  \caption{Olympic finishing times as a function of underlying marathon talent}
   \label{fig:scatter}
-  \begin{subfigure}{.5\textwidth}
+  \begin{subfigure}{.495\textwidth}
     \centering
     \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
-    \caption{} 
+    \caption{Based on personal best} 
     \label{fig:45degreeplot}
   \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
+  \begin{subfigure}{.495\textwidth}
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals.pdf}
-    \caption{}
-    \label{fig:studentizedresiduals}
+    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
+    \caption{Based on half marathon split} 
+    \label{fig:45degreeplot_half}
   \end{subfigure}
@@
 
-\begin{figure}[!ht]
-  \caption{Olympic results as a function of ability measured by half marathon split}
-  \label{fig:scatterhalf}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
-    \caption{}
-    \label{fig:45degreeplot_half}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half.pdf}
-    \caption{}
-    \label{fig:studentizedresidualshalf}
-  \end{subfigure}
-\end{figure}
-
-Figure \ref{fig:45degreeplot} includes a solid 45-degree line as
-the axes in the figure correspond to full marathons; in addition, the
-figure's rug marks denote personal best times of runners who DNFed.
-Given the paucity of points (five of them) below the pictured
-45-degree line, the vast majority of Olympic marathoners ran slower in
-Rio compared to their personal bests.  Relative to personal best
-times, the Hahner twins were definitely on the slow side, but a number
-of runners had greater differences between their Olympic times and
-their personal bests than the Hahners.  These runners are denoted with
-squares in Figure \ref{fig:45degreeplot}, and there are 13 such
-symbols, highlighting approximately 10\% of the racers who completed
-the marathon.  The figure shows moreover that there were two women who
-had personal best times slightly faster than the Hahner's and yet
-finished after the German twins.  Although Figure
-\ref{fig:45degreeplot} suggests that the Hahner twins were slower than
-one would have expected given their previous best marathon times, it
-is not consistent with the accusation that they dramatically slowed
-down in the Rio marathon.
+Considering first the relationship between Olympic finishing and
+personal best times, Figure \ref{fig:45degreeplot}'s solid 45-degree
+line is informative.  Given the paucity of points (only five of them)
+below this line, it follows that the vast majority of Olympic
+marathoners ran slower in Rio compared to prior personal bests.
+Relative to personal best times, the Hahner twins were definitely on
+the slow side, but a number of runners had greater differences between
+their Olympic times and their personal bests than the Hahners.  These
+runners are denoted with squares in Figure \ref{fig:45degreeplot}, and
+there are 13 such symbols, highlighting approximately 10\% of the
+racers who completed the Rio marathon.  Moreover, the figure shows
+that there were two women who had personal best times slightly faster
+than the Hahner's and yet finished after the German women.  Although
+Figure \ref{fig:45degreeplot} suggests that the Hahner twins were
+slower than one would have expected given their previous best marathon
+times, it is not consistent with the accusation that they dramatically
+slowed down in the Rio marathon.
 
@@
 
-One limitation of using personal best times as indicators of
-underlying athletic talent is that these times are potentially
-confounded by the marathon courses at which they were set (some
-courses, like the Berlin marathon, are known for fast times) and race
-conditions like weather.  In addition, personal best times may not
-capture raceday idiosyncrasies that might affect individual runners.
-With that in mind, Figure \ref{fig:45degreeplot_half} plots marathon
-finishing times against half marathon splits.  This figure has a
-regression line as before but no 45-degree line.
+With respect to our second measure of marathon skill, Figure
+\ref{fig:45degreeplot_half} shows that there was nothing abnormal
+about the Hahner twins' overall finishing times, conditional on their
+half marathon split times.  This is consistent with the previous
+Figure \ref{fig:45degreeplot} and inconsistent with accusations made
+against the Hahner twins, at least the part of the accusation that
+focused on their overall pace.
 
-Figure \ref{fig:45degreeplot_half} shows that there was nothing
-abnormal about the Hahner twins' overall finishing times, conditional
-on their half marathon split times.  This is consistent with the
-previous Figure \ref{fig:45degreeplot} and inconsistent with
-accusations made against the Hahner twins, at least the part of the
-accusation that focused on their  overall pace.
+Figure \ref{fig:residuals} is a visualization of how unusual, or not,
+the Hahner and Kim twins' finishing times were in the Olympic
+marathon, conditional on our two measures of marathon talent.  The
+left and right panels of the figure are based on the polynomial
+regressions in Figure \ref{fig:scatter}'s panels, respectively.  We
+calculate residuals from Figures \ref{fig:45degreeplot} and
+\ref{fig:45degreeplot_half}, we Studentize the residuals, and then we
+plot the ordered residuals in Figure \ref{fig:residuals}.
 
-If the Hahner twins did not slow down excessively, might they have
-run somewhat strategically at the end of the race in order to generate
-a simultaneous finish?  We now offer a visualization that speaks to
-this question.
+\begin{figure}[!ht]
+  \centering
+  \caption{Studentized residuals from Olympic finishing time regressions}
+  \label{fig:residuals}
+  \begin{subfigure}{.495\textwidth}
+    \centering
+    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals.pdf}
+    \caption{Based on personal best} 
+    \label{fig:studentizedresiduals}
+  \end{subfigure}
+  \begin{subfigure}{.495\textwidth}
+    \centering
+    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half.pdf}
+    \caption{Based on half marathon split}
+    \label{fig:studentizedresiduals_half}
+  \end{subfigure}
+\end{figure}
+
+The cumulative residuals depicted in Figure \ref{fig:residuals} are
+consistent with our interpretation of the Hahner twin's Olympic
+finishing times as not particularly remarkable.  With personal best
+time as a measure of marathon talent, the two Hahner residuals are in
+the right tail of the residual distribution; however, their locations
+are not extreme.  One residual is located at the 88th percentile and
+the other at the 93rd.  Kim twin residuals are similarly not
+remarkable although they are in the left tail of the residual
+distribution, indicating that the Kims ran faster than one would have
+expected.  If one uses half marathon split times as a measure of
+marathon talent, similar results obtain.  Neither Hahner twin had a
+finishing time that was particularly unusual given her half marathon
+split, and this applies to the Kim twins as well.
+
+If the Hahner twins did not slow down excessively, might they have run
+somewhat strategically at the end of the Olympic marathon in order to
+generate a simultaneous finish?  We now offer a visualization that
+speaks to this question.
 
@@
 their official finishing times were separated by one second.  Is such
-a 115 to one compression typical among pairs of runners or, in
-contrast, was it unusual?  Are there other pairs of marathoners who
-had a difference between personal best times of 115 seconds apart and,
-if so, how close where their finishing times?
+a 115 to one compression typical among pairs of runners?  Moreover,
+are there other pairs of Olympic marathoners who had a difference
+between personal best times of 115 seconds apart and, if so, how close
+where their finishing times?
 
@@
 \begin{figure}[!ht]
- \caption{Differences in personal best times and finishing times}
+ \caption{Pair-wise differences in Olympic finishing times and
+   differences in marathon talent}
   \label{fig:diffdiff}
@@
  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
- \caption{}
+ \caption{Based on differences in personal best times}
   \label{fig:diffdiffscatter}
@@
  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot_half.pdf}
- \caption{}
+ \caption{Based on differences in half marathon splits}
   \label{fig:diffdiffscatterhalf}
@@
 horizontal axis because their difference in finishing times is one
-second.  However, there are many points about the Hahner's orange dot,
+second.  However, there are many points above the Hahner's orange dot,
 and this shows that, conditional on an approximate 115 second
@@
 
-To get a sense of the lockstep nature of the Kim twins race, Figure
-\ref{fig:secondsbehind} describes each runner's status at the various
-split times on the marathon course.  Each dot in the figure depicts a
-recorded split and the number of seconds each runner was behind the
-race leader at the time.  There are more dots at earlier splits due to
-the accumulation of DNFs.  The usual color scheme applies here, and
-the Estonian DNF prior to the half marathon split is evident.  The
-orange dots in the figure represent the Hahner twins, and as noted in
-the figure Lisa Hahner was ahead of her sister through 15 kilometers
-at which point Anna surged.  Figure \ref{fig:secondsbehind} contains
-two red dots representing the North Korean Kim twins, but this is not
-visually apparent because the Kim twins had identical split times
-during the entire race marathon.
+Similar conclusions follow from Figure \ref{fig:diffdiffscatterhalf},
+which plots pair-wise differences between Olympic finishing times and
+half marathon split times.  Namely, many pairs of runners had similar
+differences in half marathon times as the Hahner and Kim twins, but
+the vast majority of these pairs did not have close finishing times.
+
+Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
+of the Kim twins, and it describes each Olympic runner's status at
+various split times on the marathon course.  Each dot in the figure
+depicts a recorded split and the number of seconds each runner was
+behind the race leader at the time.  There are more dots at earlier
+splits due to the accumulation of DNFs.  The usual color scheme
+applies here, and the Estonian triplet DNF prior to the half marathon
+split is evident. As noted in the figure Lisa Hahner was ahead of her
+sister through 15 kilometers at which point Anna took the lead.
+Figure \ref{fig:secondsbehind} contains two red dots representing the
+North Korean Kim twins, but this is not visually apparent because the
+Kim twins had identical split times during the entire marathon.
 
@@
   \includegraphics[scale = 1]{seconds-behind.pdf}
-  \begin{flushleft}
-    \emph{Note: each dot represents one runner at a split.  DNFs are
-      not pictured, and splits are not to scale.}
-  \end{flushleft}
+  % \begin{flushleft}
+  %   \emph{Note: each dot represents one runner at a split.  DNFs are
+  %     not pictured, and splits are not to scale.}
+  % \end{flushleft}
 \end{figure}
@@
 
-We now consider the following question: if we take into consideration
-the Hahner twin's similarities as runners as well as natural variation
-in marathon finishing times, what is the probability that the twins
-would finish the Rio race at roughly the same time and/or
-sequentially?  To answer this question we need to know the
-counterfactual distribution of potential marathon finishing times that
-would have occurred had the Hahner twins independently (in particular,
-of each other) and repeatedly run the Rio marathon, holding constant
-marathon conditions, the abilities of other runners, and so forth.
-Access to such a distribution would establish the set of potential
-outcomes that could have occurred on August 14, 2016, and we could in
-principle use this distribution to determine the likelihood that a
-simultaneous finish by the Hahners, or at least a near-simultaneous
-finish, occurred by chance alone.  If these twins rarely finish the
-marathon together in such a counterfactual world, then one might be
-skeptical that their observed finish occurred without
-coordination---as they in fact claimed.
+Our visualizations are useful and they shed a fair bit of light on the
+Hahner twin's performance in Rio.  In the interest of increasing
+precision, we now pose the following question: if we take into
+consideration the Hahner twin's similarities in marathon talent as
+well as natural variation in marathon finishing times, what is the
+probability that the twins would finish the Rio race at roughly the
+same time and/or sequentially?  To answer this question, we need to
+know the counterfactual distribution of potential marathon finishing
+times that would have occurred had the Hahner twins independently (in
+particular, of each other) and repeatedly run the Rio marathon,
+holding constant marathon conditions, the abilities of other runners,
+and so forth.  Access to such a distribution would establish the set
+of potential outcomes that could have occurred on August 14, 2016, and
+we could in principle use this distribution to determine the
+likelihood that a simultaneous finish by the Hahners, or at least a
+near-simultaneous finish, occurred by chance alone.  If these twins
+rarely finish the marathon together in such a counterfactual world,
+then one might be skeptical that their observed finish occurred
+without coordination---as they in fact claimed.
 
@@
 participants, it is not possible to rerun the women's Olympic marathon
-in Rio to establish a distribution of potential race outcomes for the
-Hahner twins.  However, we can attempt to simulate this distribution
-by estimating the distribution of every other runner's finishing time,
-conditional on prior abilities, and drawing from that distribution in
-order to observe the likelihood that, for example, Lisa and Anna
-Hahner finished simultaneously.
+to establish a distribution of potential race outcomes for the Hahner
+twins.  However, we can attempt to simulate this distribution by
+estimating the distribution of every other runner's finishing time,
+conditional on marathon talent, and then drawing from this
+distribution in order to calculate the likelihood that, for example,
+Lisa and Anna Hahner finished simultaneously.
 
@@
 assume that each runner's marathon time \Yi\ is distributed normally
-with a mean that is a quadratic function of her personal best time
-$X_i$:
-
-$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right).$$
+with a mean that is a quadratic function of marathon skill $X_i$. Here
+we start by assuming that marathon skill can be measured by personal
+best marathon time:
+$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right),$$
 %%
-where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.
-Personal best times are all prior to the Rio marathon.  We estimate
-$\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least squares and
-all finishing runners except for the twins and triplets.  For a
-simulated marathon, we then draw a runner's time from an estimated
-distribution, conditioned on the runner's personal best marathon time.
-Once a race is simulated for all runners, twins and triplets included,
-we locate Anna and Lisa Hahner in our set of simulated results and
-record both the time between their simulated finishes and the
-difference in their simulated ranks.  We then simulate a new
-race---drawing a new set of finishing times---and record the same
-quantities.  In particular, the steps of the simulation are as
-follows.
-
+where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.  We
+estimate $\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least
+squares and all finishing Rio marathoners except for the Hahner/Kim
+twins and Luik triplets.  For a simulated marathon, we draw a runner's
+time from an estimated distribution and condition on the runner's
+personal best marathon time.  Once a race is simulated for all
+runners, twins and triplets included, we record both the time between
+Anne and Lisa Hahner's simulated finishes and the difference in their
+simulated ranks.  We then simulate a new race---drawing a new set of
+finishing times---and record the same quantities. The steps of the
+simulation are as follows.
 \begin{enumerate}
@@
   marathon.
-\item Calculate the difference in time and difference in ranking
-  between Anna Hahner and Lisa Hahner, assuming both finished the
-  race.
+% \item Calculate the difference in time and difference in ranking
+%   between Anna Hahner and Lisa Hahner, assuming both finished the
+%   race.
 \item Repeat above steps 10,000 times.
@@
 
-David---comments here?    Or above?  This is not from the simulation
---- so perhaps move to top of section??
-
 From our simulations we can generate intervals which describe the
-extent of the variability in racer finishing times.  In 95\% of the
-simulations in which Anna Hahner completed the marathon, her finishing
-time was between 9,480 seconds and 10,703 seconds.  This is consistent
-with Anna's observed finishing time, which was 9,932 seconds.  In
-fact, our simulations estimate that Anna should have finished the Rio
-marathon in 10,092 seconds, which is slightly more than two and
-one-half minutes slower than her actual time.  Lisa Hahner's
-corresponding interval is 9,427 to 10,619 with an observed finishing
-time of 9,933 seconds.
+extent of the variability in marathon finishing times. For example, in
+95\% of the simulations in which Anna Hahner completed the marathon,
+her finishing time was between 9,480 seconds and 10,703 seconds. This
+is consistent with Anna's observed finishing time in Rio, which was
+9,932 seconds. In fact, our simulations estimate that Anna should have
+finished the Rio marathon in 10,092 seconds, which is slightly more
+than two and one-half minutes slower than her actual time. Lisa
+Hahner's corresponding 95\% interval is 9,427 to 10,619 with an
+observed finishing time of 9,933 seconds.
 
@@
 regression model underlying the simulations was estimated without the
-Hahner twins (and same for the Kim twins and Luik triplets).
-According to the our simulations, then, neither Hahner twin ran
-appreciably slowly, conditional on personal best times prior to the
-Rio Olympics.
+Hahner twins (and same for the Kim twins and Luik triplets). According
+to our simulations, then, neither Hahner twin ran appreciably slowly,
+conditional on personal best times prior to the Rio Olympics.
 
@@
 
-The results plotted in the histograms raise questions about the
+The histograms in Figure \ref{fig:simdiff} raise questions about the
 credibility of Anna and Lisa Hahner's story. In particular, the
-histogram suggest that a simultaneous finish would be very rare if
-Anna and Lisa had run independently of each other.  For example, in
-fewer than 300 of the 10,000 simulated races did Anna and Lisa
-finished within 30 seconds of each other.  Moreover, they finished in
-consecutive rank in fewer than 175 of the 10,000 races.  Therefore,
-the close finish that was observed, where Anna and Lisa crossed the
-finish line one-after-the-other, would be highly unlikely if the two
-German twins had raced independently of each other.
+histogram suggest that a simultaneous finish in the Rio marathon would
+be very rare if Anna and Lisa had run independently of each other. For
+example, in fewer than 300 of the 10,000 simulated races did Anna and
+Lisa finish within 30 seconds of each other. Moreover, they finished
+in consecutive rank in fewer than 175 of the 10,000 simulated races.
+The close finish that was observed, where Anna and Lisa crossed the
+finish line one after the other, would have been highly unlikely if
+the two German twins had raced independently of each other.
 
-Earlier we noted that personal best times might arguably be poor
-measures of runner abilities.  With this in mind, we repeated our
-simulations but replaced personal best times with half marathon
-splits.  See Figure \ref{fig:simdiffhalf}.  While this change reduces
-the variation in the predicted outcome of each runner and therefore,
-reduces the expected distance between Anna and Lisa Hahner, even with
-half marathon split as a measure of ability it is still quite rare for
-Anna and Lisa to finish simultaneously or consecutively.
+Parallel with our prior analyses, we repeated our simulations but this
+time used half marathon splits as the predicting variable in our
+simulations. The results of this are in Figure \ref{fig:simdiffhalf}.
+While this use of marathon splits reduces the variation in the
+predicted outcome of each runner and therefore reduces the expected
+distance between Anna and Lisa Hahner, even with half marathon split
+as a measure of ability it is still quite rare for Anna and Lisa to
+finish simultaneously or consecutively.
 
@@
 Figure \ref{fig:45degreeplot} shows that the Hahner twins are
-representative of sort of runners who DNFed in Rio.  Since our
+representative of the sort of runners who DNFed in Rio. Since our
 simulated Hahner statistics are conditioned on both Hahner twins
@@
 
-Looking back at the original motivation of this article, we have
-offered two perspectives on Anna and Lisa Hahner's near-simultaneous
-finish in the 2016 women's Olympic marathon in Rio.  The German twins
-were accused of deliberately slowing down and finishing next to each
-other so as to generate media attention.  They denied this, not
-entirely surprisingly,
+We have studied Anna and Lisa Hahner's near-simultaneous finish in the
+2016 women's Olympic marathon in Rio. This finish elicited a
+controversy insofar as the German twins were accused of deliberately
+slowing down and finishing next to each other so as to generate media
+attention. They denied this, not entirely surprisingly,
 
@@
 visualizations and simple calculations and a second that draws on a
-simulation.  Both perspectives have the same implications, and they
-are as follows.  In a global sense, the Hahner twins did not slow down
-appreciably during the marathon.  Their times were not fast, but they
-were within reason for runners of the Hahners' abilities.  Locally,
-though, we find that the Hahner's finish was in fact contrived.  Their
-finish---one second between the two women---was an extremely low
-probability even.  Compared to their differences as measured by
-personal best times prior to Rio, the Hahner twins difference in
-finishing times was unusually compressed.  This is evidence that their
-finishing had elements of intentionality---perhaps at the last minute
-but intentionality nonetheless.
+simulation. Both perspectives have the same implications, and they are
+as follows. In a global sense, the Hahner twins did not slow down
+appreciably during the Rio marathon. Their times were not fast, but
+they were within reason for runners of the Hahners' abilities.
+Locally, though, we find that the Hahner's finish was in fact
+contrived. Their finish---one second between the two women---was an
+extremely low probability event. Compared to their differences in
+talent, the Hahner twins difference in finishing times was unusually
+compressed. This is evidence that their finishing had elements of
+intentionality---perhaps at the last minute but intentionality
+nonetheless.
 
-Now what... David?
+Our goal is not to speak to the question of whether the Hahner twins
+should or should not have enjoyed a somewhat contrived moment of
+Olympic glory. Neither was in contention for a podium finish in Rio,
+and compared to the doping allegations that presently surround
+endurance sports in general, what the Hahners did seems relatively
+tame. Still, it might have behooved them to be a bit more open about
+their end-of-race tactics, and we have shown here how a simple data
+analysis can shed light on claims about racing results.  Should the
+2020 Summer Olympics feature twins and triplets again, we look forward
+to a comparative analysis using the techniques illustrated here.
 

### 2017-02-03 2-simulate.R

@@ -24,8 +24,8 @@ ggplot(aes(y=FINAL, x=PB), data = df) +
     geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik triplets",], aes(x = PB), colour = "blue") +
     xlim(8000,12500) +
     ylim(8000, 12500) +
-    ylab("Olympic result, in seconds\n") + 
-    xlab("\nPersonal best, in seconds") + 
+    ylab("Olympic finishing time, in seconds\n") + 
+    xlab("\nSeconds") + 
     ##  ggtitle("Distribution of Y | X") +
     scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
     scale_shape_manual(values = c(19, 15), guide = FALSE) +
@@ -45,8 +45,8 @@ ggplot(aes(y = FINAL, x = SPLIT_HALF), data = df2use) +
     geom_rug(data = df2use[is.na(df2use$FINAL),], aes(x = SPLIT_HALF), colour = "gray") +
     geom_rug(data = df2use[is.na(df2use$FINAL) & df2use$TWINS == "Luik triplets",], aes(x = PB), colour = "blue") +
     ##    coord_fixed(xlim = c(4000, 6000), ylim = c(8500, 12500)) +
-    ylab("Olympic result, in seconds\n") + 
-    xlab("\nHalf marathon split, in seconds") + 
+    ylab("Olympic finishing time, in seconds\n") + 
+    xlab("\nSeconds") + 
     ##  ggtitle("Distribution of Y | X") +
     scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
     theme_bw() +
@@ -146,8 +146,8 @@ ggplot(aes(y=abs(result_diff), x=abs(pb_diff)), data = df2b) +
     ## geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
     ## geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik triplets",], aes(x = PB), colour = "blue") +
     coord_fixed(xlim = c(0,3500), ylim = c(0,3500)) +
-    ylab("Absolute difference Olympic result, in seconds\n") + 
-    xlab("\nAbsolute difference in personal best, in seconds") + 
+    ylab("Absolute difference in Olympic finishing times, in seconds\n") + 
+    xlab("\nAbsolute differences, in seconds") + 
     ## ggtitle("Among all dyadic combinations") +
[…]

### 2017-02-03 seconds-behind.R
changed seconds behind plot

@@ -23,9 +23,9 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     cex2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- .7
     
     pch2use <- rep(19, times = nrow(dta2use))
-    pch2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- 19
-    pch2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- 19
-    pch2use[dta2use$BIB == 633  | dta2use$BIB == 634 | dta2use$BIB == 635] <- 19
+    pch2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- 1
+    pch2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- 1
+    pch2use[dta2use$BIB == 633  | dta2use$BIB == 634 | dta2use$BIB == 635] <- 1
 
     color2use <- rep("grey50", times = nrow(dta2use))
     color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- "orange"
@@ -40,21 +40,20 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
 }
 
 axis (side = 1, at = dists,
-      labels = c("5K", "10K", "15K", "20K", "", "25K", "30K", "35K", "40K", ""),
+      labels = c("5K", "10K", "15K", "", "", "25K", "30K", "35K", "", ""),
       cex.axis = 0.75,
-      padj = -1.25)
+      padj = 0)
 
-axis (side = 1, at = dists[c(5,10)],
-      labels = c("(Half)", "(Final)"),
-      cex.axis = 0.7,
-      padj = 1,
-      tick = F,
-      tck = -.1)
+axis (side = 1, at = dists[c(4,9)] + c(dists[c(5,10)] - dists[c(4,9)])/2 ,
+      labels = c("20K Half", "40K Final"),
+      cex.axis = 0.75,
+      padj = 0,
+      tick = F)
 
 legend (x = 5, y = 2000,
-        col = c("blue", "orange", "red", "grey50"),
-        pch = 19,
-        legend = c("Luik triplets", "Hahner twins", "Kim twins", "Others"),
+        col = c("blue", "orange", "red"),
+        pch = 1,
+        legend = c("Luik triplets", "Hahner twins", "Kim twins"),
         cex = 0.75,
         bty = "n"
         )

### 2017-02-04 paper/paper.tex
edits

@@
 twin/triplet status, and both plots contain second-order polynomial
-regression lines.  Rug marks in both plot indicate times (either
+regression lines.  The ticks along the x-axis in both plots indicate times (either
 personal best or half marathon split) of runners who earned a DNF
@@
 below this line, it follows that the vast majority of Olympic
-marathoners ran slower in Rio compared to prior personal bests.
-Relative to personal best times, the Hahner twins were definitely on
-the slow side, but a number of runners had greater differences between
+marathoners ran slower in Rio compared to their prior personal bests.
+The Hahner twins were definitely on the slow side, well above the 45-degree line,
+but a number of runners had even greater differences between
 their Olympic times and their personal bests than the Hahners.  These
@@
 about the Hahner twins' overall finishing times, conditional on their
-half marathon split times.  This is consistent with the previous
+half marathon split times. As one might expect, a runner's time halfway through the race is a
+fairly strong predictor of their finishing time at the end of the race.  Here we see that the relationship
+between the twins' halfway time and their finishing time is similar to that
+of the other runners.  The two points representing the Hahners are
+in the middle of the distribution of points and therefore imply that the twins do not seem to have slowed down in any unusual way. 
+ This is consistent with the previous
 Figure \ref{fig:45degreeplot} and inconsistent with accusations made
@@
 
-Figure \ref{fig:residuals} is a visualization of how unusual, or not,
+Figure \ref{fig:residuals} gives a visualization of how unusual, or not,
 the Hahner and Kim twins' finishing times were in the Olympic
 marathon, conditional on our two measures of marathon talent.  The
-left and right panels of the figure are based on the polynomial
-regressions in Figure \ref{fig:scatter}'s panels, respectively.  We
-calculate residuals from Figures \ref{fig:45degreeplot} and
-\ref{fig:45degreeplot_half}, we Studentize the residuals, and then we
-plot the ordered residuals in Figure \ref{fig:residuals}.
+left and right panels of the figure display the studentized residuals derived from the polynomial
+regressions in Figure \ref{fig:scatter}'s panels, respectively.  Hence, we
+compute the residuals from Figures \ref{fig:45degreeplot} and
+\ref{fig:45degreeplot_half}, Studentize them, and arrange the result from least to most
+along the x-axis in Figure \ref{fig:residuals} to display their cumulative distribution.
 
@@
 are not extreme.  One residual is located at the 88th percentile and
-the other at the 93rd.  Kim twin residuals are similarly not
-remarkable although they are in the left tail of the residual
+the other at the 93rd.  While they are in the tail end of the distribution, they are not the
+major outliers that would lead us to reject the twins' claim.  
+Moreover, the residuals for the Kim twins are similarly not
+remarkable, although they are in the left tail of the residual
 distribution, indicating that the Kims ran faster than one would have
@@
 between personal best times of 115 seconds apart and, if so, how close
-where their finishing times?
+were their finishing times?
 
@@
 662, and 751.  As in the Hahner case, the Kim twins compressed their
-finishing times more than any other pair of runners with a similar
+finishing times - meaning that they finished with less time between  
+them than the difference in their personal bests - more than any other pair of runners with a similar
 personal best difference.

### 2017-02-04 paper/paper.tex
first saaturdy edits

@@
 similarly for the Kim twins).  We then turn to a regression-based
-simulation of the marathon, which we develop based on relationships
-across the Olympic marathoners between personal best times and
-observed Rio finishing times.  Our simulations reinforce what we
-observed in visualization, namely, that the Hahner twins did not run
-appreciably slowly yet finished suspiciously close to each other.  We
-return in our conclusion to Kurschilgen's claims about the Hahners and
-offer thoughts about their validity.
+simulation of the marathon, and our simulations reinforce what we
+observed in prior visualizations, namely, that the Hahner twins did
+not run appreciably slowly yet finished suspiciously close to each
+other.  We return in our conclusion to Kurschilgen's claims about the
+Hahners and offer thoughts about their validity.
 
@@
 unusual---or, say, whether two athletes' finishes were jointly
-unusual---is by comparing observed finishing times on August 14 to a
-measure of underlying marathon talent.  For the latter we use prior
-personal best marathon times.  This is a natural measure of marathon
-talent, we believe, but it is not entirely free of complications.  One
-limitation of using personal best times as indicators of underlying
-marathon skill is that these times are potentially confounded by the
-marathon courses at which they were set; some courses, like the Berlin
-marathon, are known for relatively fast times.  In addition, personal
-best times may be confounded by conditions, like weather, that varied
-across races.  Finally, personal best times may not capture Olympic
-raceday idiosyncrasies that could affect individual runners.  For
-example, a runner could have woken up with a minor cold on the morning
-of August 14, 2016.  With these concerns in mind, we use half marathon
-split times from the Olympic marathon as a secondary measure of
-athletic skill in marathoning.  Our two measures of marathon talent
+unusual---is by comparing a runner's observed Olympic women's marathon
+finishing time on August 14, 2016, to a measure of her underlying
+marathon talent.  For the latter we use prior personal best marathon
+times.  This is a natural measure of marathon talent, we believe, but
+it is not entirely free of complications.  For example, personal best
+times are potentially confounded by the marathon courses on which they
+were set; some courses, like the Berlin marathon, are known for
+relatively fast times.  In addition, personal best times may be
+confounded by conditions, like weather, that varied across races.
+Finally, personal best times may not capture Olympic raceday
+idiosyncrasies that could affect individual runners.  For example, a
+runner could have woken up with a minor cold on the morning of August
+14, 2016.  With these concerns in mind, we use a runner's half
+marathon split time from the Olympic marathon as a secondary measure
+of athletic skill in marathoning.  Our two measures of marathon talent
 are featured in the visualizations below.
@@
 how finishing times from the Rio women's marathon varied as a function
-of athletes' personal best times and as function of their half
-marathon split times.  The points in the plots are colored by
-twin/triplet status, and both plots contain second-order polynomial
-regression lines.  The ticks along the x-axis in both plots indicate times (either
-personal best or half marathon split) of runners who earned a DNF
-result.
+of athletes' personal best times and half marathon split times.  The
+points in the plots are colored by twin/triplet status, and both plots
+contain second-order polynomial regression lines.  The ticks along the
+x-axis in both plots indicate times (either personal best or half
+marathon split) of runners who earned  DNF results.
 
@@
 about the Hahner twins' overall finishing times, conditional on their
-half marathon split times. As one might expect, a runner's time halfway through the race is a
-fairly strong predictor of their finishing time at the end of the race.  Here we see that the relationship
-between the twins' halfway time and their finishing time is similar to that
-of the other runners.  The two points representing the Hahners are
-in the middle of the distribution of points and therefore imply that the twins do not seem to have slowed down in any unusual way. 
- This is consistent with the previous
-Figure \ref{fig:45degreeplot} and inconsistent with accusations made
-against the Hahner twins, at least the part of the accusation that
-focused on their overall pace.
+half marathon splits. As one might expect, a runner's time halfway
+through the course is a fairly strong predictor of her finishing time.
+Here we see that the relationship between the Hehner twin's half
+marathon split times and their finishing times is similar to that of
+other Rio runners.  The two points representing the Hahners are in the
+middle of the distribution of points and therefore imply that the
+twins do not seem to have slowed down in any unusual way.  This is
+consistent with the previous Figure \ref{fig:45degreeplot} and
+inconsistent with accusations made against the Hahner twins, at least
+the part of the accusation that focused on their overall pace.
 
-Figure \ref{fig:residuals} gives a visualization of how unusual, or not,
-the Hahner and Kim twins' finishing times were in the Olympic
-marathon, conditional on our two measures of marathon talent.  The
-left and right panels of the figure display the studentized residuals derived from the polynomial
-regressions in Figure \ref{fig:scatter}'s panels, respectively.  Hence, we
-compute the residuals from Figures \ref{fig:45degreeplot} and
-\ref{fig:45degreeplot_half}, Studentize them, and arrange the result from least to most
-along the x-axis in Figure \ref{fig:residuals} to display their cumulative distribution.
+Another perspective on the extent to which the Hahners' overall
+marathon finishing times were not unusual can be gleaned from Figure
+\ref{fig:residuals}.  This figure presents cumulative distributions of
+the polynomial regression residuals from the regression models
+displayed in the two panels of Figure \ref{fig:scatter}.  In
+particular, we calculated the residuals from Figures
+\ref{fig:45degreeplot} and \ref{fig:45degreeplot_half}, Studentized
+them, and then arranged resulting Studentized residuals from least to
+most along the horizontal axes in Figures
+\ref{fig:studentizedresiduals} and \ref{fig:studentizedresiduals_half}.
 
@@
 
-The cumulative residuals depicted in Figure \ref{fig:residuals} are
-consistent with our interpretation of the Hahner twin's Olympic
-finishing times as not particularly remarkable.  With personal best
-time as a measure of marathon talent, the two Hahner residuals are in
-the right tail of the residual distribution; however, their locations
-are not extreme.  One residual is located at the 88th percentile and
-the other at the 93rd.  While they are in the tail end of the distribution, they are not the
-major outliers that would lead us to reject the twins' claim.  
-Moreover, the residuals for the Kim twins are similarly not
-remarkable, although they are in the left tail of the residual
-distribution, indicating that the Kims ran faster than one would have
-expected.  If one uses half marathon split times as a measure of
-marathon talent, similar results obtain.  Neither Hahner twin had a
-finishing time that was particularly unusual given her half marathon
-split, and this applies to the Kim twins as well.
+The cumulative residuals depicted in the two panels of Figure
+\ref{fig:residuals} are consistent with our interpretation of the
+Hahner twin's Olympic finishing times as not particularly remarkable.
+With personal best time as a measure of marathon talent as in Figure
+\ref{fig:studentizedresiduals}, the two Hahner residuals are in the
+right tail of the residual distribution; however, their locations are
+not extreme.  One residual is located at the 88th percentile and the
+other at the 93rd.  While these two residuals are in the tail end of
+the residual distribution, they are not the major outliers that would
+lead us to think that the Hahner twin's marathon finishing times were
+extremely slow.  Moreover, the residuals for the Kim twins are
+similarly not remarkable, although they are in the left tail of the
+residual distribution, indicating that the Kims ran faster than one
+would have expected.  If one uses half marathon split times as a
+measure of marathon talent, as in Figure
+\ref{fig:studentizedresiduals_half}, similar results obtain.  Neither
+Hahner twin had a finishing time that was particularly unusual given
+her half marathon split, and this applies to the Kim twins as well.
 
@@
 pairs of runners.  Of these and ignoring the Hahner twins, ten had
-exactly a 115 second gap in personal best times.  Of these ten pairs,
-differences in finishing times, in seconds, are as follows: 36, 93, 172,
-319, 379, 459, 552, 671, 675, and 739.  In other words, of all pairs
-of runners in the Rio marathon who had a personal best difference that
-was equivalent to the Hahner twins' difference, the twins had the
-greatest compression based on finishing time.
+exactly a 115 second gap in personal best times.  Differences in
+finishing times of these ten pairs, in seconds, are as follows: 36,
+93, 172, 319, 379, 459, 552, 671, 675, and 739.  In other words, of
+all pairs of runners in the Rio marathon who had a personal best
+difference that was equivalent to the Hahner twins' difference, the
+twins had the greatest compression based on finishing time.
 
@@
 where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.  We
-estimate $\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least
+estimate $\beta_1$, $\beta_2$, and $\sigma$ using ordinary least
 squares and all finishing Rio marathoners except for the Hahner/Kim
@@
   \centering
-  \caption{Distribution of simulated results for the Hahner twins}
+  \caption{Distribution of simulated results for the Hahner twins
+    based on personal best times}
   \label{fig:simdiff}
@@
   \caption{Distribution of simulated results for the Hahner twins
-    based on half marathon split}
+    based on half marathon splits}
   \label{fig:simdiffhalf} 

### 2017-02-04 paper/paper.tex
version sent to scott

@@
 that the twins ran the Rio marathon over 18 minutes slower their
-personal best times prior to the Olympics.  Not surprisingly,
-Kurschilgen's accusations were denied by the Hahner twins, who claimed
+personal best times prior to the Olympics.  Kurschilgen's accusations
+were denied by the Hahner twins, perhaps not surprisingly, who claimed
 that their simultaneous finish was simply an unintended coincidence.
@@
 
-What happened in the women's Olympic marathon, and how might we
+What happened in the women's Olympic marathon in Rio, and how might we
 develop a statistical approach that assesses whether the Hahner twin's
-finish was coincidental or intentional?  These two interpretations are
-clearly at odds. If the former, then the Hahners are to be celebrated
-and their finish treated as an expression of the spirit behind the
-Olympic games. If the latter, though, then the twins may have violated
-this spirt by not trying hard enough. It is perhaps too easy for us to
-write such a glib sentence---neither of us can fathom being able to
-complete a marathon anywhere in the vicinity of two and a half
-hours---but we nonetheless want to know what the data from the Olympic
-marathon tell us.
+finish in the race was coincidental or intentional?  These two
+interpretations are clearly at odds. If the former, then the Hahners
+are to be celebrated and their finish treated as an expression of the
+spirit behind the Olympic games. If the latter, though, then the twins
+may have violated this spirt by not trying hard enough. It is perhaps
+too easy for us to write such a glib sentence---neither of us can
+fathom being able to complete a marathon anywhere in the vicinity of
+two and a half hours---but we nonetheless want to know what the data
+from the Olympic marathon tell us.
 
@@
 
-The Kurschilgen accusation against the Hahner twins has two
-components, that these two women ran slowly \emph{and} that they
-finished simultaneously.  We suspect that Kurschilgen would not have
-expressed ire at the Hahners had they finished in 1st and 2nd place in
-Rio, hand-in-hand with wide grins, but of course we cannot be certain
-of this.  Thus, our investigation of the charges that Kurschilgen
-offered will distinguish between a slow finish versus a simultaneous
-finish.
+Kurschilgen's accusation against the Hahner twins has two components,
+that these two women ran slowly \emph{and} that they finished
+simultaneously.  We suspect that Kurschilgen would not have expressed
+ire at the Hahners had they finished in 1st and 2nd place in Rio,
+hand-in-hand with wide grins.  Thus, our investigation of the charges
+that Kurschilgen offered will distinguish between a slow finish versus
+a simultaneous finish.
 
 Our research design is twofold.  First, we present visualizations that
-describe various features of the Hahner twin's results, and an
+describe various features of the 2016 women's Olympic marathon, and an
 important element in our visualizations is the difference between
@@
 unusual---is by comparing a runner's observed Olympic women's marathon
-finishing time on August 14, 2016, to a measure of her underlying
+finishing time on August 14, 2016, with a measure of her underlying
 marathon talent.  For the latter we use prior personal best marathon
@@
 contain second-order polynomial regression lines.  The ticks along the
-x-axis in both plots indicate times (either personal best or half
-marathon split) of runners who earned  DNF results.
+horizontal axes in both plots indicate times, either personal bests or
+half marathon splits, of runners who earned DNF results.
 
@@
 differences in personal best times, and pairs of twins/triplets are
-identified by the same color scheme we used earlier.
-
-Consider first the Hahner twins.  They are effectively located on the
+identified by the same color scheme we used earlier.  Consider first
+the Hahner twins.  The two German women are effectively located on the
 horizontal axis because their difference in finishing times is one
@@
 115 second personal best differences had finishing time differences of
-1000 seconds, i.e., in excess of 15 minutes.  Of course the points in
+1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
 Figure \ref{fig:diffdiffscatter} are not independent, but they provide
@@
 662, and 751.  As in the Hahner case, the Kim twins compressed their
-finishing times - meaning that they finished with less time between  
-them than the difference in their personal bests - more than any other pair of runners with a similar
-personal best difference.
+finishing times---meaning that they finished with less time between
+them than the difference in their personal bests---more than any other
+pair of runners with similar personal best differences.
 
@@
 applies here, and the Estonian triplet DNF prior to the half marathon
-split is evident. As noted in the figure Lisa Hahner was ahead of her
+split is evident. As noted in the figure, Lisa Hahner was ahead of her
 sister through 15 kilometers at which point Anna took the lead.
@@
 
-Our visualizations are useful and they shed a fair bit of light on the
-Hahner twin's performance in Rio.  In the interest of increasing
-precision, we now pose the following question: if we take into
-consideration the Hahner twin's similarities in marathon talent as
-well as natural variation in marathon finishing times, what is the
-probability that the twins would finish the Rio race at roughly the
-same time and/or sequentially?  To answer this question, we need to
-know the counterfactual distribution of potential marathon finishing
-times that would have occurred had the Hahner twins independently (in
-particular, of each other) and repeatedly run the Rio marathon,
-holding constant marathon conditions, the abilities of other runners,
-and so forth.  Access to such a distribution would establish the set
-of potential outcomes that could have occurred on August 14, 2016, and
-we could in principle use this distribution to determine the
-likelihood that a simultaneous finish by the Hahners, or at least a
-near-simultaneous finish, occurred by chance alone.  If these twins
-rarely finish the marathon together in such a counterfactual world,
-then one might be skeptical that their observed finish occurred
-without coordination---as they in fact claimed.
+Our visualizations are useful and shed a fair bit of light on the
+Hahner twin's performance in the 2016 women's Olympic marathon.  In
+the interest of increasing precision, we now pose the following
+question: if we take into consideration the Hahner twin's similarities
+in marathon talent as well as natural variation in marathon finishing
+times, what is the probability that the twins would finish the Rio
+race at roughly the same time and/or sequentially?  To answer this
+question, we need to know the counterfactual distribution of potential
+marathon finishing times that would have occurred had the Hahner twins
+independently (in particular, of each other) and repeatedly run the
+Rio marathon, holding constant marathon conditions, the abilities of
+other runners, and so forth.  Access to such a distribution would
+establish the set of potential outcomes that could have occurred on
+August 14, 2016, and we could in principle use this distribution to
+determine the likelihood that a simultaneous finish by the Hahners, or
+at least a near-simultaneous finish, occurred by chance alone.  If
+these twins rarely finish the marathon together in such a
+counterfactual world, then one might be skeptical that their observed
+finish occurred without coordination.
 
@@
 The histograms in Figure \ref{fig:simdiff} raise questions about the
-credibility of Anna and Lisa Hahner's story. In particular, the
-histogram suggest that a simultaneous finish in the Rio marathon would
-be very rare if Anna and Lisa had run independently of each other. For
-example, in fewer than 300 of the 10,000 simulated races did Anna and
-Lisa finish within 30 seconds of each other. Moreover, they finished
-in consecutive rank in fewer than 175 of the 10,000 simulated races.
-The close finish that was observed, where Anna and Lisa crossed the
-finish line one after the other, would have been highly unlikely if
-the two German twins had raced independently of each other.
+credibility of Anna and Lisa Hahner's story and in particular suggest
+that a simultaneous finish in the Rio marathon would be very rare if
+Anna and Lisa had run independently of each other. For example, in
+fewer than 300 of the 10,000 simulated races did Anna and Lisa Hahner
+finish within 30 seconds of one another. Moreover, they finished in
+consecutive rank in fewer than 175 of the 10,000 simulated races.  The
+close finish that was observed, where Anna and Lisa crossed the finish
+line one after the other, would have been highly unlikely if the two
+German twins had raced independently of each other.
 
@@
 distance between Anna and Lisa Hahner, even with half marathon split
-as a measure of ability it is still quite rare for Anna and Lisa to
-finish simultaneously or consecutively.
+as a measure of ability it is still quite rare for the twins to finish
+simultaneously or consecutively.
 
@@
 tame. Still, it might have behooved them to be a bit more open about
-their end-of-race tactics, and we have shown here how a simple data
-analysis can shed light on claims about racing results.  Should the
-2020 Summer Olympics feature twins and triplets again, we look forward
-to a comparative analysis using the techniques illustrated here.
+their end-of-race tactics, and to this end we have shown here how a
+simple data analysis can shed light on claims about racing results.
+Should the 2020 Summer Olympics feature twins and triplets again, we
+look forward to a comparative analysis using the techniques
+illustrated here.
 

### 2017-02-04 paper/paper.tex
real version sent to scott

@@
 The media quickly picked up the Hanher story as an image of the
-beaming twins finishing hand-in-hand captured a public audience. While
-many believed the moment was a reflection of the Olympic spirit, not
-everyone agreed with this rosy interpretation. The twins' happy facial
-expressions at the finish were portrayed as a bit contrived---smiling
-like ``Honigkuchenpferde,'' cookies in the shape of a horse, was the
-description offered by one editorialist---and the sports director of
-the German Athletics Federation, Thomas Kurschilgen, stirred up
-controversy when he suggested that the Hahners' photo-finish was no
-coincidence.  Kurschilgen averred that the twins slowed down so as to
-finish simultaneously and create a spectacle which would ``generate
-media attention.''  Kurschilgen justified his charge with the fact
-that the twins ran the Rio marathon over 18 minutes slower their
-personal best times prior to the Olympics.  Kurschilgen's accusations
-were denied by the Hahner twins, perhaps not surprisingly, who claimed
-that their simultaneous finish was simply an unintended coincidence.
+beaming twins, finishing hand-in-hand, captured a public
+audience. While many believed the twin's near-simultaneous finish was
+a reflection of Olympic spirit, not everyone agreed with this rosy
+interpretation. The twins' happy facial expressions at the finish were
+portrayed as a bit contrived---smiling like ``Honigkuchenpferde,''
+cookies in the shape of a horse, was the description offered by one
+editorialist---and the sports director of the German Athletics
+Federation, Thomas Kurschilgen, stirred up controversy when he
+suggested that the Hahners' photo-finish was no coincidence.
+Kurschilgen averred that the twins slowed down so as to finish
+simultaneously and create a spectacle which would ``generate media
+attention.''  Kurschilgen justified his charge with the fact that the
+twins ran the Rio marathon over 18 minutes slower their personal best
+times prior to the Olympics.  Kurschilgen's accusations were denied by
+the Hahner twins, perhaps not surprisingly, who claimed that their
+joint finish was simply a fortuitous coincidence.
 

### 2017-02-17 paper/chance_memo.tex
ref memo outline

\documentclass[12pt]{article}
\sloppy
\usepackage{times}
\usepackage{url}
\usepackage{arydshln}

%% uncomment to use a watermark
%\usepackage{draftwatermark}
%\SetWatermarkScale{1}

\begin{document}
\thispagestyle{empty}

\noindent {\Huge Memo}\\[0.5in]

\noindent
\begin{tabular}{ll}
  \textbf{To:} &  Scott Evans, editor, \emph{Chance}\\
  \textbf{From:} & David Cottrell and Michael C.\ Herron, Dartmouth College\\
  \textbf{Date:} & \today \\
  \textbf{Re:} &  Revisions to women's marathon submission
\end{tabular}

\vspace{.1in}
\hrule
\vspace{.2in}

\bigskip

This memorandum is a response to your February 16, 2017, email
regarding our manuscript, ``All in the family: German twin finishing
times in the 2016 women's Olympic marathon.'' Thank you very much for
inviting us to submit a revision of our manuscript, and this
memorandum explains how we have updated it in light of comments from
you and the Chance Editorial Board.

blah blah blah

 \end{document}

### 2017-02-17 paper/chance_memo.tex
ref memo start

@@
 \begin{tabular}{ll}
-  \textbf{To:} &  Scott Evans, editor, \emph{Chance}\\
+  \textbf{To:} &  Scott Evans, editor, \emph{CHANCE}\\
   \textbf{From:} & David Cottrell and Michael C.\ Herron, Dartmouth College\\
@@
 times in the 2016 women's Olympic marathon.'' Thank you very much for
-inviting us to submit a revision of our manuscript, and this
-memorandum explains how we have updated it in light of comments from
-you and the Chance Editorial Board.
+inviting us to submit a revision of our manuscript to \emph{CHANCE},
+and this memorandum explains how we have updated it in light of
+comments from you and the Editorial Board.
 
-blah blah blah
+\begin{itemize}
+\item ``Am not convinced of quadratic model. Two points in upper right
+  may have influenced this. What is p-value for linear vs quadratic?
+  What happens if those 2 points are removed? Could also simply plot
+  differences from personal best as frequency dist'n. Could also plot
+  projected from half vs actual.''
+
+\item ``I printed this out in black and white. Would help see them if
+  Hahners were darkest color.''
+
+\item ``a similar conclusion is reached?''
+
+  We changed the language in the paper so it uses this phrasing.
+
+\item ``pairs are not completely independent as same runner can be in
+  different pairs. Not as though runners are drawn with replacement.''
+
+\item ``due to subsequent DNF's?''
+
+  We changed the language in the paper so it uses this phrasing.
+
+\item ``I don't get this part. If Yi is function of Xi, personal best,
+  won't estimates be same for each simulation?  What I would try for
+  an analysis would be look at distribution of finishing times,
+  expecting it to be normal, as most human performance is fairly
+  normal (most are average, some above, some below). I'd use the
+  observed mean and variance of the finishers plus a random error, and
+  see what this theoretical dist'n looks like. How often are finishers
+  within x seconds of each other? Could also use historical marathon
+  data from other races.''
+
+\item ``5a'' and ``5b''
+
+\item ``Would help to expand left side bins, as these are the finishes
+  of main interest.''
+
+\item ``what are interval widths in seconds? Hard to see.''
+
+\item ``Suddenly Latin?''
+
+  We have changed the Latin to, all things equal.
+
+\item ``likely''
+
+\end{itemize}
 

### 2017-02-17 paper/paper.tex
ref memo start

@@
 measure of marathon talent, as in Figure
-\ref{fig:studentizedresiduals_half}, similar results obtain.  Neither
-Hahner twin had a finishing time that was particularly unusual given
-her half marathon split, and this applies to the Kim twins as well.
+\ref{fig:studentizedresiduals_half}, a similar conclusion is reached.
+Neither Hahner twin had a finishing time that was particularly unusual
+given her half marathon split, and this applies to the Kim twins as
+well.
 
@@
 behind the race leader at the time.  There are more dots at earlier
-splits due to the accumulation of DNFs.  The usual color scheme
-applies here, and the Estonian triplet DNF prior to the half marathon
-split is evident. As noted in the figure, Lisa Hahner was ahead of her
-sister through 15 kilometers at which point Anna took the lead.
-Figure \ref{fig:secondsbehind} contains two red dots representing the
-North Korean Kim twins, but this is not visually apparent because the
-Kim twins had identical split times during the entire marathon.
+splits due to subsequent DNFs.  The usual color scheme applies here,
+and the Estonian triplet DNF prior to the half marathon split is
+evident. As noted in the figure, Lisa Hahner was ahead of her sister
+through 15 kilometers at which point Anna took the lead.  Figure
+\ref{fig:secondsbehind} contains two red dots representing the North
+Korean Kim twins, but this is not visually apparent because the Kim
+twins had identical split times during the entire marathon.
 
@@
 runners with better personal best times may be more likely to DNF than
-other runners, \emph{ceteris paribus}. We suspect that this occurs
-because some better runners may expend excessive energy trying to
-achieve a good result in the marathon and in so doing injure or
-exhaust themselves; lesser runners, in contrast, may be content to
-finish respectably. Regardless of the validity of this conjecture,
-Figure \ref{fig:45degreeplot} shows that the Hahner twins are
-representative of the sort of runners who DNFed in Rio. Since our
-simulated Hahner statistics are conditioned on both Hahner twins
-finishing, it follows that they are conservative. The fact that both
-women finished the marathon was notable in and of itself, and by
-discounting the possibility of a Hahner DNF we are giving the benefit
-of the doubt to the Hahner twins.
+other runners, all things equal. We suspect that this occurs because
+some better runners may expend excessive energy trying to achieve a
+good result in the marathon and in so doing injure or exhaust
+themselves; lesser runners, in contrast, may be content to finish
+respectably. Regardless of the validity of this conjecture, Figure
+\ref{fig:45degreeplot} shows that the Hahner twins are representative
+of the sort of runners who DNFed in Rio. Since our simulated Hahner
+statistics are conditioned on both Hahner twins finishing, it follows
+that they are conservative. The fact that both women finished the
+marathon was notable in and of itself, and by discounting the
+possibility of a Hahner DNF we are giving the benefit of the doubt to
+the Hahner twins.
 

### 2017-02-17 paper/chance_memo.tex
ref memo start
memo update

@@
 
+\item ``Marathon finish may be linearly related to the fastest
+  previous marathon with perhaps age in a model. An interesting though
+  experiment may be, if they were blindfolded, would they have
+  finished together?''
+
 \item ``I printed this out in black and white. Would help see them if
@@
 
-\item ``5a'' and ``5b''
+\item There are errors in figure labels, ``5a'' and ``5b''
+
+  We apologize for this oversight and have fixed the associated
+  labeling errors.
 

### 2017-02-17 1-create.R
updated R code to merge in birthday data

@@ -66,6 +66,11 @@ df <- df %>%
   select(-(RK_PB:RK_FINAL)) %>%
   mutate_each(funs(replace(., . %in% c("0:0:00", "DNS", "DNF"),  NA)), SB:FINAL) %>%
   mutate_each(funs(period_to_seconds(hms(.))), SB:FINAL)
-  
+
+## merge in birthdate data
+bdays <- read.csv (file = "data/birthday-data.txt", header = TRUE, sep = "\t")
+bdays <- bdays %>% mutate (years2marathon = as.numeric((as.Date("8/14/2016", format = "%m/%d/%Y") - as.Date(paste(birthmonth, "/", birthday, "/", birthyear, sep = ""), format = "%m/%d/%Y")) / 365))
+df <- left_join (df, bdays, by = c("BIB" = "bib"))
+
 write.csv(df, "data/times.csv", row.names = FALSE)
 

### 2017-02-22 paper/paper.tex
ref memo start
simulations with age and change Hahner color

@@
 through the course is a fairly strong predictor of her finishing time.
-Here we see that the relationship between the Hehner twin's half
+Here we see that the relationship between the Hahner twin's half
 marathon split times and their finishing times is similar to that of
@@
 horizontal axis because their difference in finishing times is one
-second.  However, there are many points above the Hahner's orange dot,
+second.  However, there are many points above the Hahner's red dot,
 and this shows that, conditional on an approximate 115 second
@@
 through 15 kilometers at which point Anna took the lead.  Figure
-\ref{fig:secondsbehind} contains two red dots representing the North
+\ref{fig:secondsbehind} contains two orange dots representing the North
 Korean Kim twins, but this is not visually apparent because the Kim
@@
 
-To estimate the conditional distribution of each Rio final results, we
+To estimate the conditional distribution of each Rio final result, we
 assume that each runner's marathon time \Yi\ is distributed normally
-with a mean that is a quadratic function of marathon skill $X_i$. Here
-we start by assuming that marathon skill can be measured by personal
-best marathon time:
-$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right),$$
+with a mean that is a quadratic function of the runner's marathon skill $X_i$ and her age $Z_i$. Hence:
+$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + \beta_3\,Z_{i} + e_i, \sigma\right),$$
 %%
-where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.  We
-estimate $\beta_1$, $\beta_2$, and $\sigma$ using ordinary least
-squares and all finishing Rio marathoners except for the Hahner/Kim
-twins and Luik triplets.  For a simulated marathon, we draw a runner's
+where $N\left(\cdot,\cdot\right)$ denotes a normal distribution. Taking each runner's 
+personal best marathon time as a proxy for their marathon skill, we
+estimate $\beta_1$, $\beta_2$, $\beta_3$ and $\sigma$ using ordinary least
+squares on all finishing Rio marathoners.  We exclude the Hahner/Kim
+twins and Luik triplets from the sample so that that our estimates are not affected by any intentional twin/triplet finish.  
+For a simulated marathon, we draw a runner's
 time from an estimated distribution and condition on the runner's
-personal best marathon time.  Once a race is simulated for all
+personal best marathon time and age.  Once a race is simulated for all
 runners, twins and triplets included, we record both the time between
-Anne and Lisa Hahner's simulated finishes and the difference in their
+Anna and Lisa Hahner's simulated finishes and the difference in their
 simulated ranks.  We then simulate a new race---drawing a new set of
@@
   linear model that predicts a runner's finishing time $Y_i$ based on
-  her pre-Olympic personal best time $X_i$ and her best time squared.
+  her pre-Olympic personal best time $X_i$, her best time squared $X^2_{i}$, and her age $Z_i$.
 \item Extract the resulting coefficient vector, estimated covariance
@@
 \item For each simulated rate, draw intercept and slope estimates
-  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
+  $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and $\tilde{\beta}_3$,
   respectively, from a multivariate normal distribution with mean
@@
   equal to the previously estimated covariance matrix.
-\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
-  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
+\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and
+  $\tilde{\beta}_3$, for each runner draw an error $\tilde{e}$ from a
   normal distribution with mean zero and a standard deviation equal to
@@
   $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
-  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
+  \tilde{\beta}_2\,X_i^2 + \tilde{\beta}_3\,Z_i +  \tilde{e}$.
 \item Eliminate each runner from the simulated race with a probability
@@
     \includegraphics[width=\textwidth,
-    keepaspectratio]{simulated_time.pdf}
+    keepaspectratio]{simulated_time_with_age.pdf}
     \caption{Simulated differences in finishing times}
@@
   \begin{subfigure}{.45\textwidth}
-    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_with_age.pdf}
     \caption{Simulated differences in places}
@@
     \includegraphics[width=\textwidth,
-    keepaspectratio]{simulated_time_half.pdf}
+    keepaspectratio]{simulated_time_half_with_age.pdf}
     \caption{Simulated differences in finishing times}
-    \label{fig:simulatedfinishtimes}
+    \label{fig:simulatedfinishtimes_half}
   \end{subfigure}
   \begin{subfigure}{.45\textwidth}
-    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half_with_age.pdf}
     \caption{Simulated differences in places}
-    \label{fig:simulatedranks}
+    \label{fig:simulatedranks_half}
   \end{subfigure}

### 2017-02-22 1-create.R

@@ -1,5 +1,6 @@
 library(tidyverse)
 library(rvest)
+library(lubridate)
 
 from_web = FALSE
 

### 2017-02-22 2-simulate.R

@@ -27,7 +27,7 @@ ggplot(aes(y=FINAL, x=PB), data = df) +
     ylab("Olympic finishing time, in seconds\n") + 
     xlab("\nSeconds") + 
     ##  ggtitle("Distribution of Y | X") +
-    scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
+    scale_color_manual(values = c("red", "orange", "blue", "gray"), name = "") +
     scale_shape_manual(values = c(19, 15), guide = FALSE) +
     theme_bw() +
     theme(legend.position = c(.8,.2))
@@ -48,7 +48,7 @@ ggplot(aes(y = FINAL, x = SPLIT_HALF), data = df2use) +
     ylab("Olympic finishing time, in seconds\n") + 
     xlab("\nSeconds") + 
     ##  ggtitle("Distribution of Y | X") +
-    scale_color_manual(values = c("orange", "red", "blue", "gray"), name = "") +
+    scale_color_manual(values = c("red", "orange", "blue", "gray"), name = "") +
     theme_bw() +
     theme(legend.position = c(0.8, 0.2))
 dev.off()
@@ -65,14 +65,14 @@ pdf("plots/studentized_residuals.pdf", height = 5, width = 5)
 ggplot(dta, aes(STUD_RESID, PERCENTILE)) + 
   geom_point(size = 1, colour = "gray") + 
   geom_point(data = na.omit(dta), aes( col = TWINS)) + 
-  geom_text(data = dta[dta$TWINS == "Hahner twins",], aes(label = print_pct(PERCENTILE)), nudge_x = -.4, colour = "orange") +
-  geom_text(data = dta[dta$TWINS == "Kim twins",], aes(label = print_pct(PERCENTILE)), nudge_x = .4, colour = "red") +
+  geom_text(data = dta[dta$TWINS == "Hahner twins",], aes(label = print_pct(PERCENTILE)), nudge_x = -.4, colour = "red") +
+  geom_text(data = dta[dta$TWINS == "Kim twins",], aes(label = print_pct(PERCENTILE)), nudge_x = .4, colour = "orange") +
   geom_text(data = dta[dta$TWINS == "Luik triplets",], aes(label = print_pct(PERCENTILE)), nudge_x = .4, colour = "blue") +
   scale_x_continuous(breaks = -10:10) +
   ##  ggtitle("Distribution of Studentized Residuals") +
   ylab("Percentile\n") +
[…]

### 2017-02-22 seconds-behind.R

@@ -28,8 +28,8 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     pch2use[dta2use$BIB == 633  | dta2use$BIB == 634 | dta2use$BIB == 635] <- 1
 
     color2use <- rep("grey50", times = nrow(dta2use))
-    color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- "orange"
-    color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- "red"
+    color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- "red"
+    color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- "orange"
     color2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- "blue"
 
     points (x = rep(dists[match(col2use, cols2use)], times = nrow(dta2use)), y = dta2use[, col2use] - min(dta2use[, col2use]),
@@ -51,7 +51,7 @@ axis (side = 1, at = dists[c(4,9)] + c(dists[c(5,10)] - dists[c(4,9)])/2 ,
       tick = F)
 
 legend (x = 5, y = 2000,
-        col = c("blue", "orange", "red"),
+        col = c("blue", "red", "orange"),
         pch = 1,
         legend = c("Luik triplets", "Hahner twins", "Kim twins"),
         cex = 0.75,

### 2017-02-22 simhalfwithage.R

@@ -1,0 +1,88 @@
+# Calculate the percent of runners who did not finish.
+# (includes women who did not start)
+df_half <- df[!is.na(df$SPLIT_HALF),] 
+dnf_rate <- sum(is.na(df_half$FINAL))/length(df_half$FINAL)
+
+# Estimate linear relationship between personal best and marathon result, excluding twins.
+m <- lm(FINAL ~ SPLIT_HALF + I(SPLIT_HALF^2) + years2marathon, data = df_half[is.na(df_half$TWINS),])
+
+# Store the total number of runners, N
+N <- length(df_half$SPLIT_HALF)
+
+# Save the vector of personal bests
+x <- df_half$SPLIT_HALF
+age <- df_half$years2marathon
+
+# Create data frame to capture simulation output
+sim_df <- data_frame(ATHLETE = df_half$ATHLETE)
+
+# Set seed
+set.seed(14252345)
+
+# Set total number of sims
+total_sims <- 10000
+
+# Begin simulation
+for (i in 1:total_sims){
+  
+  # Determine whether each runner finishes given the dnf rate
+  dnf <- sample(x = c(F,T), size = N, replace = T, prob = c(1-dnf_rate, dnf_rate) )
+  
+  # Draw each runner's result from a multivariate normal distribution with 
+  # mean Beta and sigma equal to the variance-covariance of Beta
+  # Draw an error term from a normal distibution with standard deviation equal 
+  # to the standard deviation of the model residuals.
+  B <- mvrnorm(n = 1, mu = coef(m), Sigma =  vcov(m))
+  e <- rnorm(n = N, mean = 0, sd = summary(m)$sigma)
+  
+  # Remove the results of those that did not finish
+  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] + age*B[4] + e )
+  
+  # Collect the results
+  sim_df[[paste0(i)]] <- y_hat
+  
+}
+
+sim_df <- sim_df %>% gather(SIM, YHAT, -ATHLETE, convert = T) %>% group_by(SIM) %>% mutate(RANK = min_rank(YHAT))
+
+
+# Compare Anna to Lisa
+#
+anna <- df %>% filter(ATHLETE == "Anna Hahner")
[…]

### 2017-02-22 simwithage.R

@@ -1,0 +1,91 @@
+
+# Calculate the percent of runners who did not finish.
+# (includes women who did not start)
+dnf_rate <- sum(is.na(df$FINAL))/length(df$FINAL)
+
+# Estimate linear relationship between personal best and marathon result, excluding twins.
+m <- lm(FINAL ~ PB + I(PB^2) + years2marathon, data = df[is.na(df$TWINS),])
+
+# Store the total number of runners, N
+N <- length(df$PB)
+
+# Save the vector of personal bests
+x <- df$PB
+age <- df$years2marathon
+
+# Create data frame to capture simulation output
+sim_df <- data_frame(ATHLETE = df$ATHLETE)
+
+# Set seed
+set.seed(14252345)
+
+# Set total number of sims
+total_sims <- 10000
+
+# Begin simulation
+for (i in 1:total_sims){
+  
+  # Determine whether each runner finishes given the dnf rate
+  dnf <- sample(x = c(F,T), size = N, replace = T, prob = c(1-dnf_rate, dnf_rate) )
+  
+  # Draw each runner's result from a multivariate normal distribution with 
+  # mean Beta and sigma equal to the variance-covariance of Beta
+  # Draw an error term from a normal distibution with standard deviation equal 
+  # to the standard deviation of the model residuals.
+  B <- mvrnorm(n = 1, mu = coef(m), Sigma =  vcov(m))
+  e <- rnorm(n = N, mean = 0, sd = summary(m)$sigma)
+
+  
+  # Remove the results of those that did not finish
+  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] + age*B[4] + e )
+
+
+  # Collect the results
+  sim_df[[paste0(i)]] <- y_hat
+  
+}
+
+sim_df <- sim_df %>% gather(SIM, YHAT, -ATHLETE, convert = T) %>% group_by(SIM) %>% mutate(RANK = min_rank(YHAT))
+
+
+# Compare Anna to Lisa
+#
+anna <- df %>% filter(ATHLETE == "Anna Hahner")
+anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
+
+lisa <- df %>% filter(ATHLETE == "Lisa Hahner")
[…]

### 2017-02-23 paper/chance_memo.tex
updated memo

@@
   projected from half vs actual.''
+  \begin{itemize}
+  \item
+  \end{itemize}
 
@@
   finished together?''
+  \begin{itemize}
+  \item We've changed model to include age for the simulations. 
+  \end{itemize}
 
@@
   Hahners were darkest color.''
+   \begin{itemize}
+  \item
+  \end{itemize}
 
 \item ``a similar conclusion is reached?''
-
-  We changed the language in the paper so it uses this phrasing.
+  \begin{itemize}
+  \item We changed the language in the paper so it uses this phrasing.
+  \end{itemize}
 
@@
   different pairs. Not as though runners are drawn with replacement.''
+   \begin{itemize}
+  \item
+  \end{itemize}
 
 \item ``due to subsequent DNF's?''
-
-  We changed the language in the paper so it uses this phrasing.
+   \begin{itemize}
+  \item We changed the language in the paper so it uses this phrasing.
+  \end{itemize}
 
@@
   data from other races.''
+   \begin{itemize}
+  \item WE DID THIS, RIGHT?
+  \end{itemize}
 
 \item There are errors in figure labels, ``5a'' and ``5b''
-
-  We apologize for this oversight and have fixed the associated
+   \begin{itemize}
+  \item  We apologize for this oversight and have fixed the associated
   labeling errors.
+  \end{itemize}
 
@@
   of main interest.''
+   \begin{itemize}
+  \item
+  \end{itemize}
 
 \item ``what are interval widths in seconds? Hard to see.''
+   \begin{itemize}
+  \item
+  \end{itemize}
 
 \item ``Suddenly Latin?''
-
-  We have changed the Latin to, all things equal.
+   \begin{itemize}
+   \item We have changed the Latin to, all things equal. 
+   \end{itemize}
 
 \item ``likely''
+   \begin{itemize}
+  \item
+  \end{itemize}
 

### 2017-02-23 paper/chance_memo.tex
misc edits

--- 2017-02-17 paper/chance_memo.tex
@@
 \usepackage{times}
+\usepackage{xr-hyper}
 \usepackage{url}
 \usepackage{arydshln}
+
+\externaldocument{paper}
 
@@
 
+  The $p$-value for the quadratic term is -1.88, which is not
+  particularly compelling.  We have accordingly dropped the quadratic
+  terms in the scatter plots on p.\ 4 and, implicitly, from the
+  Studentized residual plots on p.\ 6.
+  %% Note: xr labels are
+  %% breaking; looks like some sort of
+  %% label clash with another package.
+  %% Hardcoding here is poor style.
+
+  % > fit <- lm (FINAL ~ PB + I(PB^2), data = df)
+  % > print (summary(fit))
+
+  % Call:
+  % lm(formula = FINAL ~ PB + I(PB^2), data = df)
+
+  % Residuals:
+  % Min      1Q  Median      3Q     Max 
+  % -817.64 -288.70  -92.13  224.05 1319.33
+
+  % Coefficients:
+  % Estimate Std. Error t value Pr(>|t|)   
+  % (Intercept) -1.257e+04  6.417e+03  -1.958  0.05233 . 
+  % PB           3.633e+00  1.336e+00   2.719  0.00744 **
+  % I(PB^2)     -1.309e-04  6.946e-05  -1.884  0.06181 . 
+  % ---
+  % Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+  % Residual standard error: 417.5 on 130 degrees of freedom
+  % (24 observations deleted due to missingness)
+  % Multiple R-squared:  0.6538,	Adjusted R-squared:  0.6484 
+  % F-statistic: 122.7 on 2 and 130 DF,  p-value: < 2.2e-16
+
 \item ``Marathon finish may be linearly related to the fastest
@@
 
+  We have added age to our simulation model, and the results are
+  qualitatively equivalent.  In addition, if one adds age to the
+  linear specification depicted in our basic scatter plot (now without
+  a quadratic term, as noted above), age is a significant predictor of
+  marathon finishing time, even controlling for personal best
+  ($p \approx 2.426$).  Older runners are slower, \emph{ceteris
+    paribus}.
+
+% Call:
+% lm(formula = FINAL ~ PB + years2marathon, data = df)
+
+% Residuals:
+%     Min      1Q  Median      3Q     Max 
+% -779.88 -281.31  -94.71  222.21 1319.29
+
+% Coefficients:
+%                  Estimate Std. Error t value Pr(>|t|)    
+% (Intercept)    -1.366e+03  7.430e+02  -1.838   0.0684 .  
+% PB              1.147e+00  7.222e-02  15.876   <2e-16 ***
+% years2marathon  1.799e+01  7.413e+00   2.426   0.0166 *  
+% ---
+% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
+
+% Residual standard error: 413.9 on 130 degrees of freedom
+%   (23 observations deleted due to missingness)
+% Multiple R-squared:  0.6597,	Adjusted R-squared:  0.6545 
+% F-statistic:   126 on 2 and 130 DF,  p-value: < 2.2e-16
+
 \item ``I printed this out in black and white. Would help see them if
@@
 
+  We have changed the color scheme for the dots in our figures.  The
+  German Hahner twins are now shown in black, the North Korean Kim
+  twins in red, and the Estonian Luik triplets in blue.  Each color is
+  present in its country's respective flag.
+
 \item ``a similar conclusion is reached?''
 
-  We changed the language in the paper so it uses this phrasing.
+  We changed the language in the paper so that it uses this phrasing.
 
@@
 
-  We changed the language in the paper so it uses this phrasing.
+  We changed the language in the paper so that it uses this phrasing.
 
@@
 
-\item There are errors in figure labels, ``5a'' and ``5b''
+  We fully agree with this comment, and indeed the intuition offered
+  by the referee was the motivation behind our simulation.  To make
+  this clear, we added a paragraph after the simulation steps, and
+  this paragraph explains why finishing times (and order) will vary
+  both within and across simulations.  Then, the plots we generated
+  (see Figures 5 and 6) describe how often the Hahner twins finish
+  in close proximity of each other.
+
+  We have thought about historical marathon data but are hesitant to
+  go down that route.  The set of runners who competed in Rio is
+  unique to this race, and marathon finishes are confounded by course
+  and local conditions.
+
+\item There are errors in figure labels ``5a'' and ``5b''
 
@@
 
-  We have changed the Latin to, all things equal.
+  We have changed the Latin text to, all things equal.
 
 \item ``likely''
+
+  We apologize for the confusion here, but we are not sure what the
+  referee is requesting.  In the marked-up PDF document, the word
+  ``likely'' appears over the following sentence: ``Their finish—one
+  second between the two women—was an extremely low probability
+  event.''  We are not sure how to incorporate the referee's request
+  but are happy to do so if given more information.
 

### 2017-02-23 paper/paper.tex
misc edits

@@
 horizontal axes in both plots indicate times, either personal bests or
-half marathon splits, of runners who earned DNF results.
+half marathon splits, of runners who earned DNF
+results.\label{scatterplots}
 
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_no_quadratic.pdf}
     \caption{Based on personal best} 
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half_no_quadratic.pdf}
     \caption{Based on half marathon split} 
@@
 most along the horizontal axes in Figures
-\ref{fig:studentizedresiduals} and \ref{fig:studentizedresiduals_half}.
+\ref{fig:studentizedresiduals} and
+\ref{fig:studentizedresiduals_half}.\label{residualplots}
 
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_no_quadratic.pdf}
     \caption{Based on personal best} 
@@
     \centering
-    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half.pdf}
+    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half_no_quadratic.pdf}
     \caption{Based on half marathon split}
@@
 \ref{fig:residuals} are consistent with our interpretation of the
-Hahner twin's Olympic finishing times as not particularly remarkable.
-With personal best time as a measure of marathon talent as in Figure
-\ref{fig:studentizedresiduals}, the two Hahner residuals are in the
-right tail of the residual distribution; however, their locations are
-not extreme.  One residual is located at the 88th percentile and the
-other at the 93rd.  While these two residuals are in the tail end of
-the residual distribution, they are not the major outliers that would
-lead us to think that the Hahner twin's marathon finishing times were
-extremely slow.  Moreover, the residuals for the Kim twins are
-similarly not remarkable, although they are in the left tail of the
-residual distribution, indicating that the Kims ran faster than one
-would have expected.  If one uses half marathon split times as a
-measure of marathon talent, as in Figure
+Hahner twin's Olympic finishing times as not particularly
+remarkable. With personal best time as a measure of marathon talent as
+in Figure \ref{fig:studentizedresiduals}, the two Hahner residuals are
+in the right tail of the residual distribution; however, their
+locations are not extreme.  One residual is located at the 88th
+percentile and the other at the 93rd.  While these two residuals are
+in the tail end of the residual distribution, they are not the major
+outliers that would lead us to think that the Hahner twin's marathon
+finishing times were extremely slow.  Moreover, the residuals for the
+Kim twins are similarly not remarkable, although they are in the left
+tail of the residual distribution, indicating that the Kims ran faster
+than one would have expected.  If one uses half marathon split times
+as a measure of marathon talent, as in Figure
 \ref{fig:studentizedresiduals_half}, a similar conclusion is reached.
@@
 horizontal axis because their difference in finishing times is one
-second.  However, there are many points above the Hahner's red dot,
+second.  However, there are many points above the Hahner's black dot,
 and this shows that, conditional on an approximate 115 second
@@
 through 15 kilometers at which point Anna took the lead.  Figure
-\ref{fig:secondsbehind} contains two orange dots representing the North
+\ref{fig:secondsbehind} contains two red dots representing the North
 Korean Kim twins, but this is not visually apparent because the Kim
@@
 
-Our visualizations are useful and shed a fair bit of light on the
-Hahner twin's performance in the 2016 women's Olympic marathon.  In
-the interest of increasing precision, we now pose the following
-question: if we take into consideration the Hahner twin's similarities
-in marathon talent as well as natural variation in marathon finishing
-times, what is the probability that the twins would finish the Rio
-race at roughly the same time and/or sequentially?  To answer this
-question, we need to know the counterfactual distribution of potential
-marathon finishing times that would have occurred had the Hahner twins
-independently (in particular, of each other) and repeatedly run the
-Rio marathon, holding constant marathon conditions, the abilities of
-other runners, and so forth.  Access to such a distribution would
-establish the set of potential outcomes that could have occurred on
-August 14, 2016, and we could in principle use this distribution to
-determine the likelihood that a simultaneous finish by the Hahners, or
-at least a near-simultaneous finish, occurred by chance alone.  If
-these twins rarely finish the marathon together in such a
-counterfactual world, then one might be skeptical that their observed
-finish occurred without coordination.
+Our visualizations shed a fair bit of light on the Hahner twin's
+performance in the 2016 women's Olympic marathon.  In the interest of
+increasing precision, we now pose the following question: if we take
+into consideration the twin's similarities in marathon talent as well
+as natural variation in marathon finishing times, what is the
+probability that they would finish the Rio race at roughly the same
+time and/or sequentially?  To answer this question, we need to know
+the counterfactual distribution of potential marathon finishing times
+that would have occurred had the Hahner twins independently (in
+particular, of each other) and repeatedly run the Rio marathon,
+holding constant marathon conditions, the abilities of other runners,
+and so forth.  Access to such a distribution would establish the set
+of potential outcomes that could have occurred on August 14, 2016, and
+we could in principle use this distribution to determine the
+likelihood that a simultaneous finish by the Hahners, or at least a
+near-simultaneous finish, occurred by chance alone.  If these twins
+rarely finish the marathon together in such a counterfactual world,
+then one might be skeptical that their observed finish occurred
+without some degree of coordination.
 
@@
 distribution in order to calculate the likelihood that, for example,
-Lisa and Anna Hahner finished simultaneously.
+Lisa and Anna Hahner finished the Rio race simultaneously.
 
@@
 assume that each runner's marathon time \Yi\ is distributed normally
-with a mean that is a quadratic function of the runner's marathon skill $X_i$ and her age $Z_i$. Hence:
+with a mean that is a quadratic function of the runner's marathon
+skill $X_i$ and her age $Z_i$. Hence:
 $$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + \beta_3\,Z_{i} + e_i, \sigma\right),$$
 %%
-where $N\left(\cdot,\cdot\right)$ denotes a normal distribution. Taking each runner's 
-personal best marathon time as a proxy for their marathon skill, we
-estimate $\beta_1$, $\beta_2$, $\beta_3$ and $\sigma$ using ordinary least
-squares on all finishing Rio marathoners.  We exclude the Hahner/Kim
-twins and Luik triplets from the sample so that that our estimates are not affected by any intentional twin/triplet finish.  
-For a simulated marathon, we draw a runner's
+where $N\left(\cdot,\cdot\right)$ denotes a normal
+distribution. Taking each runner's personal best marathon time as a
+proxy for her underlying marathon skill, we estimate $\beta_1$,
+$\beta_2$, $\beta_3$ and $\sigma$ using ordinary least squares on all
+finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
+triplets from the sample so that that our estimates are not affected
+by the twin/triplet finishes which, theoretically, could reflect the
+results of coordination.  For a simulated marathon, we draw a runner's
 time from an estimated distribution and condition on the runner's
@@
   linear model that predicts a runner's finishing time $Y_i$ based on
-  her pre-Olympic personal best time $X_i$, her best time squared $X^2_{i}$, and her age $Z_i$.
+  her pre-Olympic personal best time $X_i$, her best time squared
+  $X^2_{i}$, and her age $Z_i$.
 \item Extract the resulting coefficient vector, estimated covariance
@@
 \item For each simulated rate, draw intercept and slope estimates
-  $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and $\tilde{\beta}_3$,
-  respectively, from a multivariate normal distribution with mean
-  equal to the previously estimated coefficient vector and covariance
-  equal to the previously estimated covariance matrix.
-\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and
-  $\tilde{\beta}_3$, for each runner draw an error $\tilde{e}$ from a
-  normal distribution with mean zero and a standard deviation equal to
-  the standard deviation of the original regression model's residuals.
-\item For each runner, predict her final result by combining the
-  randomly generated beta coefficients and error terms,
+  $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and
+  $\tilde{\beta}_3$, respectively, from a multivariate normal
+  distribution with mean equal to the previously estimated coefficient
+  vector and covariance equal to the previously estimated covariance
+  matrix.
+\item For each runner, draw an error $\tilde{e}$ from a normal
+  distribution with mean zero and a standard deviation equal to the
+  standard deviation of the original regression model's residuals.
+\item Predict each runner's final result by combining the randomly
+  generated beta coefficients and individual error terms,
   $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
@@
 \end{enumerate}
+
+As these steps illustrate, our simulation repeatedly draws random
+coefficient vectors, and this captures variability in what we know
+about the relationship between runner talent and age and runer finish
+times.  In addition, the simulation draws random disturbances for each
+runner, conditional on our original estimate of regression variance;
+these disturbances capture variability in runner finishing times,
+notwithstanding underlying runner talent.  Importantly, the
+disturbances that we draw are independent across runners.  This means
+that the finishing order among runners for each simulated race will
+vary.
 

### 2017-02-23 2-simulate.R

@@ -4,6 +4,11 @@ library(lubridate)
 library(stringr)
 df <- read.csv("data/times.csv", stringsAsFactors = FALSE) %>% tbl_df()
 
+## set color for twins/triplets
+color_kim <- "red"
+color_hahner <- "black"
+color_luik <- "blue"
+
 # Remove Irvette Van Zyl, since she did not start the race due to
 # injury.  Her bib number is 1172.
 df <- df[df$BIB != 1172,]
@@ -14,27 +19,55 @@ df <- df[df$BIB != 1172,]
 ## add indicator for finishing at least 19 minutes slower than PB
 df <- df %>% mutate(slower19 = (FINAL - PB) / 60 >= 19)
 
+## print basic regression to check for p-value of quadratic term
+fit <- lm (FINAL ~ PB + I(PB^2), data = df)
+print (summary(fit))
+
+## print basic regression with age
+fit <- lm (FINAL ~ PB + years2marathon, data = df)
+print (summary(fit))
+
+
+## make final versus pb scatter plots, with and without quadratic terms
 pdf("plots/scatter_plot.pdf", height = 5,  width = 5)
 ggplot(aes(y=FINAL, x=PB), data = df) + 
     geom_point(size = 1.25, colour = "gray", aes(shape = slower19)) +
     geom_point(data = df[!is.na(df$TWINS),], aes(col = TWINS)) +
     geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = "black", alpha = .5, size = .5, linetype = "dashed", fill = "black", se = FALSE) + 
-    geom_abline(intercept = 0, slope =1) +
+    geom_abline(intercept = 0, slope = 1) +
     geom_rug(data = df[is.na(df$FINAL),], aes(x = PB), colour = "gray") +
-    geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik triplets",], aes(x = PB), colour = "blue") +
+    geom_rug(data = df[is.na(df$FINAL) & df$TWINS == "Luik triplets",], aes(x = PB), colour = color_luik) +
     xlim(8000,12500) +
     ylim(8000, 12500) +
     ylab("Olympic finishing time, in seconds\n") + 
     xlab("\nSeconds") + 
     ##  ggtitle("Distribution of Y | X") +
[…]

### 2017-02-23 paper/paper.tex
updates to simulations

@@
 points in the plots are colored by twin/triplet status, and both plots
-contain second-order polynomial regression lines.  The ticks along the
+contain %second-order polynomial regression lines. 
+a dashed line representing the line of best fit.
+The ticks along the
 horizontal axes in both plots indicate times, either personal bests or
@@
 assume that each runner's marathon time \Yi\ is distributed normally
-with a mean that is a quadratic function of the runner's marathon
+with a mean that is a linear function of the runner's marathon
 skill $X_i$ and her age $Z_i$. Hence:
-$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + \beta_3\,Z_{i} + e_i, \sigma\right),$$
+$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
 %%
@@
   linear model that predicts a runner's finishing time $Y_i$ based on
-  her pre-Olympic personal best time $X_i$, her best time squared
-  $X^2_{i}$, and her age $Z_i$.
+  her pre-Olympic personal best time $X_i$ and her age $Z_i$.
 \item Extract the resulting coefficient vector, estimated covariance
@@
 \item For each simulated rate, draw intercept and slope estimates
-  $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and
-  $\tilde{\beta}_3$, respectively, from a multivariate normal
+  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
+  $\tilde{\beta}_2$, respectively, from a multivariate normal
   distribution with mean equal to the previously estimated coefficient
@@
   generated beta coefficients and individual error terms,
-  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
-  \tilde{\beta}_2\,X_i^2 + \tilde{\beta}_3\,Z_i + \tilde{e}$.
+  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +  \tilde{\beta}_2\,Z_i + \tilde{e}$.
 \item Eliminate each runner from the simulated race with a probability
@@
 coefficient vectors, and this captures variability in what we know
-about the relationship between runner talent and age and runer finish
+about the relationship between runner talent and age and runner finish
 times.  In addition, the simulation draws random disturbances for each

### 2017-02-23 2-simulate.R

@@ -257,133 +257,14 @@ dev.off()
 
 stop()
 
-# Calculate the percent of runners who did not finish.
-# (includes women who did not start)
-dnf_rate <- sum(is.na(df$FINAL))/length(df$FINAL)
 
-# Estimate linear relationship between personal best and marathon result, excluding twins.
-m <- lm(FINAL ~ PB + I(PB^2), data = df[is.na(df$TWINS),])
-
-# Store the total number of runners, N
-N <- length(df$PB)
-
-# Save the vector of personal bests
-x <- df$PB
-
-# Create data frame to capture simulation output
-sim_df <- data_frame(ATHLETE = df$ATHLETE)
-
-# Set seed
-set.seed(14252345)
-
-# Set total number of sims
-total_sims <- 10000
-
-# Begin simulation
-for (i in 1:total_sims){
-  
-  # Determine whether each runner finishes given the dnf rate
-  dnf <- sample(x = c(F,T), size = N, replace = T, prob = c(1-dnf_rate, dnf_rate) )
-  
-  # Draw each runner's result from a multivariate normal distribution with 
-  # mean Beta and sigma equal to the variance-covariance of Beta
-  # Draw an error term from a normal distibution with standard deviation equal 
-  # to the standard deviation of the model residuals.
-  B <- mvrnorm(n = 1, mu = coef(m), Sigma =  vcov(m))
-  e <- rnorm(n = N, mean = 0, sd = summary(m)$sigma)
-  
-  # Remove the results of those that did not finish
-  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] +  e )
-  
-
-  # Collect the results
-  sim_df[[paste0(i)]] <- y_hat
-  
-}
-
-sim_df <- sim_df %>% gather(SIM, YHAT, -ATHLETE, convert = T) %>% group_by(SIM) %>% mutate(RANK = min_rank(YHAT))
+max_x_t <- 2300
+max_y_t <- 500
+max_x_r <- 124
+max_y_r <- 300
 
 
-# Compare Anna to Lisa
-#
-anna <- df %>% filter(ATHLETE == "Anna Hahner")
-anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
-
[…]

### 2017-02-23 seconds-behind.R

@@ -28,8 +28,8 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     pch2use[dta2use$BIB == 633  | dta2use$BIB == 634 | dta2use$BIB == 635] <- 1
 
     color2use <- rep("grey50", times = nrow(dta2use))
-    color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- "red"
-    color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- "orange"
+    color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- "black"
+    color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- "red"
     color2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- "blue"
 
     points (x = rep(dists[match(col2use, cols2use)], times = nrow(dta2use)), y = dta2use[, col2use] - min(dta2use[, col2use]),
@@ -51,7 +51,7 @@ axis (side = 1, at = dists[c(4,9)] + c(dists[c(5,10)] - dists[c(4,9)])/2 ,
       tick = F)
 
 legend (x = 5, y = 2000,
-        col = c("blue", "red", "orange"),
+        col = c("blue", "black", "red"),
         pch = 1,
         legend = c("Luik triplets", "Hahner twins", "Kim twins"),
         cex = 0.75,

### 2017-02-23 sim.R

@@ -1,21 +1,20 @@
+
 # Calculate the percent of runners who did not finish.
 # (includes women who did not start)
-df_half <- df[!is.na(df$SPLIT_HALF),] 
-dnf_rate <- sum(is.na(df_half$FINAL))/length(df_half$FINAL)
+dnf_rate <- sum(is.na(df$FINAL))/length(df$FINAL)
 
 # Estimate linear relationship between personal best and marathon result, excluding twins.
-#m <- lm(FINAL ~ SPLIT_HALF + I(SPLIT_HALF^2) + years2marathon, data = df_half[is.na(df_half$TWINS),])
-m <- lm(FINAL ~ SPLIT_HALF + years2marathon, data = df_half[is.na(df_half$TWINS),])
+#m <- lm(FINAL ~ PB + I(PB^2), data = df[is.na(df$TWINS),])
+m <- lm(FINAL ~ PB, data = df[is.na(df$TWINS),])
 
 # Store the total number of runners, N
-N <- length(df_half$SPLIT_HALF)
+N <- length(df$PB)
 
 # Save the vector of personal bests
-x <- df_half$SPLIT_HALF
-age <- df_half$years2marathon
+x <- df$PB
 
 # Create data frame to capture simulation output
-sim_df <- data_frame(ATHLETE = df_half$ATHLETE)
+sim_df <- data_frame(ATHLETE = df$ATHLETE)
 
 # Set seed
 set.seed(14252345)
@@ -37,8 +36,9 @@ for (i in 1:total_sims){
   e <- rnorm(n = N, mean = 0, sd = summary(m)$sigma)
   
   # Remove the results of those that did not finish
-  #y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] + age*B[4] + e )
-  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + age*B[3] + e )
+  #y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] +  e )
+  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] +  e )
+  
   
   # Collect the results
   sim_df[[paste0(i)]] <- y_hat
@@ -46,7 +46,6 @@ for (i in 1:total_sims){
 }
 
 sim_df <- sim_df %>% gather(SIM, YHAT, -ATHLETE, convert = T) %>% group_by(SIM) %>% mutate(RANK = min_rank(YHAT))
-
 
[…]

### 2017-02-23 simhalf.R

@@ -4,7 +4,8 @@ df_half <- df[!is.na(df$SPLIT_HALF),]
 dnf_rate <- sum(is.na(df_half$FINAL))/length(df_half$FINAL)
 
 # Estimate linear relationship between personal best and marathon result, excluding twins.
-m <- lm(FINAL ~ SPLIT_HALF + I(SPLIT_HALF^2), data = df_half[is.na(df_half$TWINS),])
+# m <- lm(FINAL ~ SPLIT_HALF + I(SPLIT_HALF^2), data = df_half[is.na(df_half$TWINS),])
+m <- lm(FINAL ~ SPLIT_HALF, data = df_half[is.na(df_half$TWINS),])
 
 # Store the total number of runners, N
 N <- length(df_half$SPLIT_HALF)
@@ -35,8 +36,8 @@ for (i in 1:total_sims){
   e <- rnorm(n = N, mean = 0, sd = summary(m)$sigma)
   
   # Remove the results of those that did not finish
-  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] +  e )
-  
+  # y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] +  e )
+  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] +  e )
   
   # Collect the results
   sim_df[[paste0(i)]] <- y_hat
@@ -54,35 +55,42 @@ anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
 lisa <- df %>% filter(ATHLETE == "Lisa Hahner")
 lisa_sim <- sim_df %>% filter(ATHLETE == "Lisa Hahner")
 
-hahner_dist <- abs(anna_sim$YHAT - lisa_sim$YHAT) <= abs(anna$FINAL - lisa$FINAL)
-hahner_dist <- ifelse(is.na(hahner_dist), FALSE, hahner_dist)
-
-hahner_rank <- abs(anna_sim$RANK - lisa_sim$RANK) <= 1
-hahner_rank <- ifelse(is.na(hahner_rank), FALSE, hahner_rank)
-
-pct_less_than_hahner <- sum(hahner_dist)/length(hahner_dist)
-pct_consecutive_hahner <- sum(hahner_rank)/length(hahner_rank)
-
 hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
 
[…]

### 2017-02-23 simwithage.R

@@ -4,7 +4,8 @@
 dnf_rate <- sum(is.na(df$FINAL))/length(df$FINAL)
 
 # Estimate linear relationship between personal best and marathon result, excluding twins.
-m <- lm(FINAL ~ PB + I(PB^2) + years2marathon, data = df[is.na(df$TWINS),])
+# m <- lm(FINAL ~ PB + I(PB^2) + years2marathon, data = df[is.na(df$TWINS),])
+m <- lm(FINAL ~ PB + years2marathon, data = df[is.na(df$TWINS),])
 
 # Store the total number of runners, N
 N <- length(df$PB)
@@ -37,7 +38,8 @@ for (i in 1:total_sims){
 
   
   # Remove the results of those that did not finish
-  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] + age*B[4] + e )
+  # y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + x^2*B[3] + age*B[4] + e )
+  y_hat <- ifelse(dnf, NA, B[1] + x*B[2] + age*B[3] + e )
 
 
   # Collect the results
@@ -56,36 +58,42 @@ anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
 lisa <- df %>% filter(ATHLETE == "Lisa Hahner")
 lisa_sim <- sim_df %>% filter(ATHLETE == "Lisa Hahner")
 
-hahner_dist <- abs(anna_sim$YHAT - lisa_sim$YHAT) <= abs(anna$FINAL - lisa$FINAL)
-hahner_dist <- ifelse(is.na(hahner_dist), FALSE, hahner_dist)
-
-hahner_rank <- abs(anna_sim$RANK - lisa_sim$RANK) <= 1
-hahner_rank <- ifelse(is.na(hahner_rank), FALSE, hahner_rank)
-
-pct_less_than_hahner <- sum(hahner_dist)/length(hahner_dist)
-pct_consecutive_hahner <- sum(hahner_rank)/length(hahner_rank)
-
 hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
 
+pct_less_than_hahner <- sum((hahner_df$time_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
[…]

### 2017-02-24 paper/chance_memo.tex
misc edits
updated histograms, simulations code, paper text, and memo text

@@
 
-  The $p$-value for the quadratic term is -1.88, which is not
+  The $t$-value for the quadratic term is -1.88 and the corresponding $p$-value is .062, which is not
   particularly compelling.  We have accordingly dropped the quadratic
@@
   marathon finishing time, even controlling for personal best
-  ($p \approx 2.426$).  Older runners are slower, \emph{ceteris
+  ($t \approx 2.426$, $p \approx 0.017$).  Older runners are slower, \emph{ceteris
     paribus}.
@@
 
+  We have highlighted the left side bins in red and annotated the plot 
+  with a description of the size of the those bins. This emphasizes the finishes
+  that are the main interest of the result.
+
 \item ``what are interval widths in seconds? Hard to see.''
+
+  We agree that this is hard to see.  The text now points out that bin
+  width is 30 seconds.
 

### 2017-02-24 paper/paper.tex
updated histograms, simulations code, paper text, and memo text

@@
 95\% of the simulations in which Anna Hahner completed the marathon,
-her finishing time was between 9,480 seconds and 10,703 seconds. This
+her finishing time was between 8,369 seconds and 9,999 seconds. This
 is consistent with Anna's observed finishing time in Rio, which was
 9,932 seconds. In fact, our simulations estimate that Anna should have
-finished the Rio marathon in 10,092 seconds, which is slightly more
-than two and one-half minutes slower than her actual time. Lisa
-Hahner's corresponding 95\% interval is 9,427 to 10,619 with an
+finished the Rio marathon in 9,181 seconds, which is slightly more
+than 12 minutes faster than her actual time. Lisa
+Hahner's corresponding 95\% interval is 8,507 to 10,143 with an
 observed finishing time of 9,933 seconds.
@@
 Hahner twins (and same for the Kim twins and Luik triplets). According
-to our simulations, then, neither Hahner twin ran appreciably slowly,
-conditional on personal best times prior to the Rio Olympics.
+to our simulation, then, both Hahner twins ran appreciably slowly,
+conditional on personal best times prior to the Rio Olympics and age.
 
-% 95% simulated interval for Anna Hahner finishing time in seconds: 9480 10703 
-% 95% simulated interval for Lisa Hahner finishing time in seconds: 9427 10619
+% 95% simulated interval for Anna Hahner finishing time in seconds: 8369 9999 
+% 95% simulated interval for Lisa Hahner finishing time in seconds: 8507 10143
+
+% > mean(hahner_df_simwithage$anna_final, na.rm = T)
+% [1] 9181.137
+% > mean(hahner_df_simwithage$lisa_final, na.rm = T)
+% [1] 9319.427
+
+% > (9932 - 9181.137)/60
+% [1] 12.51438
 
@@
 finish within 30 seconds of one another. Moreover, they finished in
-consecutive rank in fewer than 175 of the 10,000 simulated races.  The
+consecutive rank in fewer than 200 of the 10,000 simulated races.  The
 close finish that was observed, where Anna and Lisa crossed the finish

### 2017-02-24 2-simulate.R

@@ -265,6 +265,41 @@ max_y_r <- 300
 
 
 source("sim.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 2.71% of simulations
+# The average simulated time between them was 8.07 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 8486 10133 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 8604 10268 
 source("simhalf.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 2.71% of simulations
+# The average simulated time between them was 5.8 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 9468 10670 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 9387 10586 
 source("simwithage.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 1.82% of simulations
+# The average simulated time between them was 7.87 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 8369 9999 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 8507 10143 
 source("simhalfwithage.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 2.59% of simulations
+# The average simulated time between them was 5.85 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 9493 10717 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 9443 10637 
+
+
+# Save simulation results
+save(hahner_df_sim, 
+     hahner_df_simhalf, 
+     hahner_df_simwithage, 
+     hahner_df_simhalfwithage, 
+     file = "sim_results.RData")
+
+
+mean(hahner_df_simwithage$anna_final, na.rm = T)
+mean(hahner_df_simwithage$lisa_final, na.rm = T)
+mean(hahner_df_simwithage$anna_rank, na.rm = T)
+mean(hahner_df_simwithage$lisa_rank, na.rm = T)
+sum(hahner_df_simwithage$rank_diff == 1, na.rm = T)

### 2017-02-24 sim.R

@@ -1,4 +1,3 @@
-
 # Calculate the percent of runners who did not finish.
 # (includes women who did not start)
 dnf_rate <- sum(is.na(df$FINAL))/length(df$FINAL)
@@ -55,43 +54,51 @@ anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
 lisa <- df %>% filter(ATHLETE == "Lisa Hahner")
 lisa_sim <- sim_df %>% filter(ATHLETE == "Lisa Hahner")
 
-hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
+hahner_df_sim <- data_frame(anna_final = anna_sim$YHAT,
+                            lisa_final = lisa_sim$YHAT,
+                            anna_rank = anna_sim$RANK,
+                            lisa_rank = lisa_sim$RANK,
+                            time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), 
+                            rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
 
-pct_less_than_hahner <- sum((hahner_df$time_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_less_than_min <- sum((hahner_df$time_diff) < 60, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_consecutive <- sum((hahner_df$rank_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
+less_than_hahner <- sum((hahner_df_sim$time_diff) <= 1, na.rm = T ) 
+less_than_min <- sum((hahner_df_sim$time_diff) < 60, na.rm = T )
+consecutive <- sum((hahner_df_sim$rank_diff) <= 1, na.rm = T )
+pct_less_than_hahner <-less_than_hahner / total_sims
+pct_less_than_min <- less_than_min / total_sims
+pct_consecutive <- consecutive / total_sims
 
 if (exists("p")) {rm(p)}
 if (exists("p2")) {rm(p2)}
 
 
 pdf("plots/simulated_time.pdf", height = 5, width = 5)
-p<-ggplot(filter(hahner_df, !is.na(time_diff)), aes(time_diff, fill = time_diff < 60 )) + 
+p<-ggplot(filter(hahner_df_sim, !is.na(time_diff)), aes(time_diff, fill = time_diff < 60 )) + 
[…]

### 2017-02-24 simhalf.R

@@ -55,42 +55,50 @@ anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
 lisa <- df %>% filter(ATHLETE == "Lisa Hahner")
 lisa_sim <- sim_df %>% filter(ATHLETE == "Lisa Hahner")
 
-hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
+hahner_df_simhalf <-  data_frame(anna_final = anna_sim$YHAT,
+                                 lisa_final = lisa_sim$YHAT,
+                                 anna_rank = anna_sim$RANK,
+                                 lisa_rank = lisa_sim$RANK,
+                                 time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), 
+                                 rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
 
-pct_less_than_hahner <- sum((hahner_df$time_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_less_than_min <- sum((hahner_df$time_diff) < 60, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_consecutive <- sum((hahner_df$rank_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
+less_than_hahner <- sum((hahner_df_simhalf$time_diff) <= 1, na.rm = T ) 
+less_than_min <- sum((hahner_df_simhalf$time_diff) < 60, na.rm = T )
+consecutive <- sum((hahner_df_simhalf$rank_diff) <= 1, na.rm = T )
+pct_less_than_hahner <-less_than_hahner / total_sims
+pct_less_than_min <- less_than_min / total_sims
+pct_consecutive <- consecutive / total_sims
 
 if (exists("p")) {rm(p)}
 if (exists("p2")) {rm(p2)}
 
 pdf("plots/simulated_time_half.pdf", height = 5, width = 5)
-p <- ggplot(filter(hahner_df, !is.na(time_diff)), aes(time_diff, fill = time_diff < 60 )) + 
+p <- ggplot(filter(hahner_df_simhalf, !is.na(time_diff)), aes(time_diff, fill = time_diff < 60 )) + 
   geom_histogram(breaks = seq(0,max_x_t, 30), colour = "black") +
[…]

### 2017-02-24 simhalfwithage.R

@@ -56,42 +56,50 @@ anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
 lisa <- df %>% filter(ATHLETE == "Lisa Hahner")
 lisa_sim <- sim_df %>% filter(ATHLETE == "Lisa Hahner")
 
-hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
+hahner_df_simhalfwithage <- data_frame(anna_final = anna_sim$YHAT,
+                                       lisa_final = lisa_sim$YHAT,
+                                       anna_rank = anna_sim$RANK,
+                                       lisa_rank = lisa_sim$RANK,
+                                       time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), 
+                                       rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
 
-pct_less_than_hahner <- sum((hahner_df$time_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_less_than_min <- sum((hahner_df$time_diff) < 60, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_consecutive <- sum((hahner_df$rank_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
+less_than_hahner <- sum((hahner_df_simhalfwithage$time_diff) <= 1, na.rm = T ) 
+less_than_min <- sum((hahner_df_simhalfwithage$time_diff) < 60, na.rm = T )
+consecutive <- sum((hahner_df_simhalfwithage$rank_diff) <= 1, na.rm = T )
+pct_less_than_hahner <-less_than_hahner / total_sims
+pct_less_than_min <- less_than_min / total_sims
+pct_consecutive <- consecutive / total_sims
 
 if (exists("p")) {rm(p)}
 if (exists("p2")) {rm(p2)}
 
 pdf("plots/simulated_time_half_with_age.pdf", height = 5, width = 5)
-p <- ggplot(filter(hahner_df, !is.na(time_diff)), aes(time_diff, fill = time_diff < 60 )) + 
+p <- ggplot(filter(hahner_df_simhalfwithage, !is.na(time_diff)), aes(time_diff, fill = time_diff < 60 )) + 
   geom_histogram(breaks = seq(0,max_x_t, 30), colour = "black") +
[…]

### 2017-02-24 simwithage.R

@@ -1,4 +1,3 @@
-
 # Calculate the percent of runners who did not finish.
 # (includes women who did not start)
 dnf_rate <- sum(is.na(df$FINAL))/length(df$FINAL)
@@ -58,42 +57,50 @@ anna_sim <- sim_df %>% filter(ATHLETE == "Anna Hahner")
 lisa <- df %>% filter(ATHLETE == "Lisa Hahner")
 lisa_sim <- sim_df %>% filter(ATHLETE == "Lisa Hahner")
 
-hahner_df <- data_frame(time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
+hahner_df_simwithage <-  data_frame(anna_final = anna_sim$YHAT,
+                                    lisa_final = lisa_sim$YHAT,
+                                    anna_rank = anna_sim$RANK,
+                                    lisa_rank = lisa_sim$RANK,
+                                    time_diff = abs(anna_sim$YHAT - lisa_sim$YHAT), 
+                                    rank_diff =  abs(anna_sim$RANK - lisa_sim$RANK))
 
-pct_less_than_hahner <- sum((hahner_df$time_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_less_than_min <- sum((hahner_df$time_diff) < 60, na.rm = T ) / length(na.omit(hahner_df$time_diff))
-pct_consecutive <- sum((hahner_df$rank_diff) <= 1, na.rm = T ) / length(na.omit(hahner_df$time_diff))
+less_than_hahner <- sum((hahner_df_simwithage$time_diff) <= 1, na.rm = T ) 
+less_than_min <- sum((hahner_df_simwithage$time_diff) < 60, na.rm = T )
+consecutive <- sum((hahner_df_simwithage$rank_diff) <= 1, na.rm = T )
+pct_less_than_hahner <-less_than_hahner / total_sims
+pct_less_than_min <- less_than_min / total_sims
+pct_consecutive <- consecutive / total_sims
 
 if (exists("p")) {rm(p)}
 if (exists("p2")) {rm(p2)}
 
 pdf("plots/simulated_time_with_age.pdf", height = 5, width = 5)
-p <- ggplot(filter(hahner_df, !is.na(time_diff)), aes(time_diff, fill = time_diff < 60 )) + 
[…]

### 2017-02-24 paper/chance_memo.tex
misc edits, almost ready for resubmission

@@
 
-  The $t$-value for the quadratic term is -1.88 and the corresponding $p$-value is .062, which is not
-  particularly compelling.  We have accordingly dropped the quadratic
-  terms in the scatter plots on p.\ 4 and, implicitly, from the
-  Studentized residual plots on p.\ 6.
+  The $t$-value for the quadratic term is -1.88 and the corresponding
+  $p$-value is 0.062, which is not particularly compelling.  We have
+  accordingly dropped the quadratic terms in the scatter plots on p.\
+  4 and, implicitly, from the Studentized residual plots on p.\ 6.  In
+  terms of the latter suggestion about, Figure 1b contains a plot of
+  half marathon split times versus actual finishing times.
   %% Note: xr labels are
@@
   marathon finishing time, even controlling for personal best
-  ($t \approx 2.426$, $p \approx 0.017$).  Older runners are slower, \emph{ceteris
-    paribus}.
+  ($t \approx 2.426$, $p \approx 0.017$).  Older runners are slower,
+  all things equal.
 
@@
 
+  We agree with the thrust of this comment, and our manuscript now
+  emphasizes the lack of independence.  Insofar as dependence in
+  sampling typically diminishes variance, our lack of independence in
+  the pair plots should make our results conservative. This means that
+  that the observed Hahner and Kim compression in final results,
+  conditional on prior best marathon times, is even more unusual.
+
 \item ``due to subsequent DNF's?''
@@
 
-  We have highlighted the left side bins in red and annotated the plot 
-  with a description of the size of the those bins. This emphasizes the finishes
-  that are the main interest of the result.
+  We agree with this suggestion.  Accordingly, in the histograms that
+  describe simulation results, we have highlighted left side bins in
+  red and annotated the plots with descriptions of the sizes of key
+  areas.  This emphasizes the finishes that are most important for our
+  results.
 

### 2017-02-24 paper/paper.tex
misc edits, almost ready for resubmission

@@
 several things: personal best marathon time prior to the 2016 Olympic
-games; split times from the Rio marathon course at 5 kilometers, 10
-kilometers, and so forth; and, finishing time. We cannot directly
-observe the effort that an individual put into the race, and we do not
-know why some runners have DNF results.  Some runners may have injured
-themselves on the course and accordingly dropped out, and others may
-have dropped out, uninjured, in anticipation of an unsatisfactory
-result. Of the 156 marathon starters, 133 completed the race and 23
-DNFed at various locations throughout the course. The overall DNF rate
-was thus $\frac{23}{156} \approx 0.15$, and the relatively small
-sample size at our disposal means that a 95\% confidence for this rate
-is fairly wide, namely, $\left(0.098, 0.22\right)$.
+games; age; split times from the Rio marathon course at 5 kilometers,
+10 kilometers, and so forth; and, overall finishing time. We cannot
+directly observe the effort that an individual put into the race, and
+we do not know why some runners have DNF results.  Some runners may
+have injured themselves on the course and accordingly dropped out, and
+others may have dropped out, uninjured, in anticipation of an
+unsatisfactory result. Of the 156 marathon starters, 133 completed the
+race and 23 DNFed at various locations throughout the course. The
+overall DNF rate was thus $\frac{23}{156} \approx 0.15$, and the
+relatively small sample size at our disposal yields a relatively wide
+95\% confidence for this rate, namely, $\left(0.098, 0.22\right)$.
 
@@
 hand-in-hand with wide grins.  Thus, our investigation of the charges
-that Kurschilgen offered will distinguish between a slow finish versus
-a simultaneous finish.
+that Kurschilgen offered distinguishes between a slow finish and a
+simultaneous finish.
 
@@
 
-One way that we might assess whether an athlete's Rio finish was
-unusual---or, say, whether two athletes' finishes were jointly
-unusual---is by comparing a runner's observed Olympic women's marathon
-finishing time on August 14, 2016, with a measure of her underlying
-marathon talent.  For the latter we use prior personal best marathon
-times.  This is a natural measure of marathon talent, we believe, but
-it is not entirely free of complications.  For example, personal best
-times are potentially confounded by the marathon courses on which they
-were set; some courses, like the Berlin marathon, are known for
-relatively fast times.  In addition, personal best times may be
-confounded by conditions, like weather, that varied across races.
-Finally, personal best times may not capture Olympic raceday
-idiosyncrasies that could affect individual runners.  For example, a
-runner could have woken up with a minor cold on the morning of August
-14, 2016.  With these concerns in mind, we use a runner's half
-marathon split time from the Olympic marathon as a secondary measure
-of athletic skill in marathoning.  Our two measures of marathon talent
-are featured in the visualizations below.
+One way that we might assess whether a women marathoner's Rio finish
+was unusual---or, say, whether two finishes were jointly unusual---is
+by comparing a runner's observed Olympic finishing time on August 14,
+2016, with a measure of her underlying marathon talent.  For the
+latter we use prior personal best marathon times.  This is a natural
+measure of marathon talent, we believe, but it is not entirely free of
+complications.  For example, personal best times are potentially
+confounded by the marathon courses on which they were set; some
+courses, like the Berlin marathon, are known for relatively fast
+times.  In addition, personal best times may be confounded by
+conditions, like weather, that varied across races.  Finally, personal
+best times may not capture Olympic raceday idiosyncrasies that could
+affect individual runners.  For example, a runner could have woken up
+with a minor cold on the morning of August 14, 2016.  With these
+concerns in mind, we use a runner's half marathon split time from the
+Olympic marathon as a secondary measure of athletic skill in
+marathoning.  Our two measures of marathon talent are featured in the
+visualizations below.
 
@@
 contain %second-order polynomial regression lines.
-a dashed line representing the line of best fit.
-The ticks along the
-horizontal axes in both plots indicate times, either personal bests or
-half marathon splits, of runners who earned DNF
+dashed lines representing least squares regression fits.  The ticks
+along the horizontal axes in both plots indicate times, either
+personal bests or half marathon splits, of runners who earned DNF
 results.\label{scatterplots}
@@
 \ref{fig:residuals}.  This figure presents cumulative distributions of
-the polynomial regression residuals from the regression models
-displayed in the two panels of Figure \ref{fig:scatter}.  In
-particular, we calculated the residuals from Figures
-\ref{fig:45degreeplot} and \ref{fig:45degreeplot_half}, Studentized
-them, and then arranged resulting Studentized residuals from least to
-most along the horizontal axes in Figures
-\ref{fig:studentizedresiduals} and
+the residuals from the regression models displayed in the two panels
+of Figure \ref{fig:scatter}.  In particular, we calculated the
+residuals from Figures \ref{fig:45degreeplot} and
+\ref{fig:45degreeplot_half}, Studentized them, and then arranged
+resulting Studentized residuals from least to most along the
+horizontal axes in Figures \ref{fig:studentizedresiduals} and
 \ref{fig:studentizedresiduals_half}.\label{residualplots}
@@
 in the right tail of the residual distribution; however, their
-locations are not extreme.  One residual is located at the 88th
-percentile and the other at the 93rd.  While these two residuals are
-in the tail end of the residual distribution, they are not the major
+locations are not extreme.  One residual is located at the 87th
+percentile and the other at the 92rd.  While these two residuals are
+in the tail end of the residual distribution, they are not major
 outliers that would lead us to think that the Hahner twin's marathon
 finishing times were extremely slow.  Moreover, the residuals for the
-Kim twins are similarly not remarkable, although they are in the left
-tail of the residual distribution, indicating that the Kims ran faster
-than one would have expected.  If one uses half marathon split times
-as a measure of marathon talent, as in Figure
-\ref{fig:studentizedresiduals_half}, a similar conclusion is reached.
+Kim twins are similarly not remarkable; these two residuals are in the
+left tail of the residual distribution, indicating that the Kims ran
+faster than one would have expected.  Finally, if one relies on half
+marathon split times as measures of marathon talent, as in Figure
+\ref{fig:studentizedresiduals_half}, similar conclusions follows.
 Neither Hahner twin had a finishing time that was particularly unusual
@@
 115 second personal best differences had finishing time differences of
-1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
-Figure \ref{fig:diffdiffscatter} are not independent, but they provide
-a sense of the dispersion in finishing time differences that one can
-expect conditional on differences in personal best times.
+1,000 seconds, i.e., in excess of 15 minutes.  The points in Figure
+\ref{fig:diffdiffscatter} are not independent, but they provide a
+rough sense of the dispersion in finishing time differences between
+runners that one might expect conditional on differences in personal
+best times.
 
@@
 
-Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
-of the Kim twins, and it describes each Olympic runner's status at
-various split times on the marathon course.  Each dot in the figure
-depicts a recorded split and the number of seconds each runner was
-behind the race leader at the time.  There are more dots at earlier
-splits due to subsequent DNFs.  The usual color scheme applies here,
-and the Estonian triplet DNF prior to the half marathon split is
-evident. As noted in the figure, Lisa Hahner was ahead of her sister
-through 15 kilometers at which point Anna took the lead.  Figure
-\ref{fig:secondsbehind} contains two red dots representing the North
-Korean Kim twins, but this is not visually apparent because the Kim
-twins had identical split times during the entire marathon.
+Figure \ref{fig:secondsbehind} describes each Olympic runner's status
+at various split times on the marathon course.  Each dot in the
+figure---color as before---depicts a recorded split and the number of
+seconds each runner was behind the race leader at the time.  There are
+more dots at earlier splits due to subsequent DNFs.
 
@@
 \end{figure}
+
+The Estonian triplet DNF prior to the half marathon split is evident
+in Figure \ref{fig:secondsbehind} , which also shows that Lisa Hahner
+was ahead of her sister through 15 kilometers.  The figure contains
+two red dots representing the North Korean Kim twins, but this is not
+visually apparent because the Kim twins had identical split times
+during the entire marathon.
 
@@
 distribution. Taking each runner's personal best marathon time as a
-proxy for her underlying marathon skill, we estimate $\beta_1$,
-$\beta_2$, $\beta_3$ and $\sigma$ using ordinary least squares on all
+proxy for her underlying marathon skill, we estimate $\beta_0$,
+$\beta_1$, $\beta_2$, and $\sigma$ using ordinary least squares on
 finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
 triplets from the sample so that that our estimates are not affected
-by the twin/triplet finishes which, theoretically, could reflect the
-results of coordination.  For a simulated marathon, we draw a runner's
+by the twin/triplet finishes which, theoretically, could reflect
+runner coordination.  For a simulated marathon, we draw a runner's
 time from an estimated distribution and condition on the runner's
@@
 these disturbances capture variability in runner finishing times,
-notwithstanding underlying runner talent.  Importantly, the
-disturbances that we draw are independent across runners.  This means
-that the finishing order among runners for each simulated race will
-vary.
+notwithstanding underlying marathon talent as proxied for by prior
+best finishing time.  Importantly, the disturbances that we draw are
+independent across runners.  Consequently, for each simulated race the
+finishing order among runners will vary.
 
-From our simulations we can generate intervals which describe the
-extent of the variability in marathon finishing times. For example, in
-95\% of the simulations in which Anna Hahner completed the marathon,
-her finishing time was between 8,369 seconds and 9,999 seconds. This
-is consistent with Anna's observed finishing time in Rio, which was
-9,932 seconds. In fact, our simulations estimate that Anna should have
+From our simulations we generate intervals which describe the extent
+of the variability in marathon finishing times. For example, in 95\%
+of the simulations in which Anna Hahner completed the marathon, her
+finishing time was between 8,369 seconds and 9,999 seconds. This is
+consistent with Anna's observed finishing time in Rio, which was 9,932
+seconds. In fact, our simulations estimate that Anna should have
 finished the Rio marathon in 9,181 seconds, which is slightly more
-than 12 minutes faster than her actual time. Lisa
-Hahner's corresponding 95\% interval is 8,507 to 10,143 with an
-observed finishing time of 9,933 seconds.
+than 12 minutes faster than her actual time. Lisa Hahner's
+corresponding 95\% interval is 8,507 to 10,143 with an observed
+finishing time of 9,933 seconds.
 
@@
 Hahner twins (and same for the Kim twins and Luik triplets). According
-to our simulation, then, both Hahner twins ran appreciably slowly,
-conditional on personal best times prior to the Rio Olympics and age.
+to our simulation, then, both Hahner twins did not run appreciably
+slowly, conditional on personal best times prior to the Rio Olympics
+and age.
 
@@
 With an eye on the matter of simultaneous finishing, Figure
-\ref{fig:simdiff} contains two histograms based on our simulated race
+\ref{fig:simdiff} contains two histograms based on simulated race
 results. Figure \ref{fig:simulatedfinishtimes} is a histogram which
@@
   \centering
-  \caption{Distribution of simulated results for the Hahner twins
-    based on personal best times}
+  \caption{Distribution of Hahner twin results in 10,000 simulated
+    marathons, based on personal best times}
   \label{fig:simdiff}
@@
     keepaspectratio]{simulated_time_with_age.pdf}
-    \caption{Simulated differences in finishing times}
+    \caption{Differences in Hahner finishing times}
     \label{fig:simulatedfinishtimes}
@@
     \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_with_age.pdf}
-    \caption{Simulated differences in places}
+    \caption{Differences in Hahner twin places}
     \label{fig:simulatedranks}
@@
 that a simultaneous finish in the Rio marathon would be very rare if
-Anna and Lisa had run independently of each other. For example, in
-fewer than 300 of the 10,000 simulated races did Anna and Lisa Hahner
-finish within 30 seconds of one another. Moreover, they finished in
-consecutive rank in fewer than 200 of the 10,000 simulated races.  The
-close finish that was observed, where Anna and Lisa crossed the finish
-line one after the other, would have been highly unlikely if the two
-German twins had raced independently of each other.
+Anna and Lisa had run independently. For example, in fewer than 300 of
+10,000 simulated races did Anna and Lisa Hahner finish within 30
+seconds of one another, and in fewer than 600 did the Hahner twins
+finish within a minute of each other.  The histogram area associated
+with this result is depicted in red in Figure
+\ref{fig:simulatedfinishtimes}.  Moreover, the Hahner twins finished
+in consecutive rank in fewer than 200 of 10,000 simulated races; the
+red zone in Figure \ref{fig:simulatedranks} presents this visually.
+The close finish that was observed in Rio, where Anna and Lisa crossed
+the finish line one after the other, would have been highly unlikely
+if the two German twins had raced independently of each other.
 
-Parallel with our prior analyses, we repeated our simulations but this
-time used half marathon splits as the predicting variable in our
-simulations. The results of this are in Figure \ref{fig:simdiffhalf}.
-While this use of marathon splits reduces the variation in the
-predicted outcome of each runner and therefore reduces the expected
-distance between Anna and Lisa Hahner, even with half marathon split
-as a measure of ability it is still quite rare for the twins to finish
-simultaneously or consecutively.
+Parallel with our prior analyses, we repeated our simulations using
+half marathon splits as the predicting variable in our simulations;
+results are in Figure \ref{fig:simdiffhalf}.  While this use of
+marathon splits reduces the variation in the predicted outcome of each
+runner and therefore reduces the expected distance between Anna and
+Lisa Hahner, even with half marathon split as a measure of ability it
+is still quite rare for the twins to finish simultaneously or
+consecutively.
 
@@
   \centering
-  \caption{Distribution of simulated results for the Hahner twins
-    based on half marathon splits}
+  \caption{Distribution of Hahner twin results in 10,000 simulated
+    marathons, based on half marathon splits}
   \label{fig:simdiffhalf} 
@@
 
-An important point regarding our simulations is the way that we
-handled DNFs. As our description above indicates, we assumed that DNF
-probabilities are the same for all runners and that the likelihood of
-a DNF is a not function of a runner's anticipated marathon finishing
-time. The rug marks in Figure \ref{fig:45degreeplot} suggest that
-runners with better personal best times may be more likely to DNF than
-other runners, all things equal. We suspect that this occurs because
-some better runners may expend excessive energy trying to achieve a
-good result in the marathon and in so doing injure or exhaust
-themselves; lesser runners, in contrast, may be content to finish
+The way that we handled DNFs in our simulations is notable. As our
+description above indicates, we assumed that DNF probabilities are the
+same for all runners and that the likelihood of a DNF is not a
+function of a runner's anticipated marathon finishing time. The rug
+marks in Figure \ref{fig:45degreeplot} suggest that runners with
+better personal best times may be more likely to DNF than other
+runners, all things equal. We suspect that this occurs because some
+better runners may expend excessive energy trying to achieve a good
+result in the marathon and in so doing injure or exhaust themselves;
+lesser runners, in contrast, may be content to finish
 respectably. Regardless of the validity of this conjecture, Figure

### 2017-02-24 seconds-behind.R

@@ -1,5 +1,12 @@
 library(dplyr)
 
+## set color for twins/triplets
+color_kim <- "red"
+color_hahner <- "black"
+color_luik <- "blue"
+
+
+
 dta <- read.csv (file = "data/times.csv", header = TRUE)
 
 pdf (file = "plots/seconds-behind.pdf", width = 5, height = 5)
@@ -28,9 +35,9 @@ for (col2use in cols2use <- grep("SPLIT|FINAL", colnames(dta))) {
     pch2use[dta2use$BIB == 633  | dta2use$BIB == 634 | dta2use$BIB == 635] <- 1
 
     color2use <- rep("grey50", times = nrow(dta2use))
-    color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- "black"
-    color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- "red"
-    color2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- "blue"
+    color2use[dta2use$BIB == 748 | dta2use$BIB == 749] <- color_hahner
+    color2use[dta2use$BIB == 1137 | dta2use$BIB == 1138] <- color_kim
+    color2use[dta2use$BIB == 633 | dta2use$BIB == 634 | dta2use$BIB == 635] <- color_luik
 
     points (x = rep(dists[match(col2use, cols2use)], times = nrow(dta2use)), y = dta2use[, col2use] - min(dta2use[, col2use]),
            col = color2use,
@@ -51,9 +58,9 @@ axis (side = 1, at = dists[c(4,9)] + c(dists[c(5,10)] - dists[c(4,9)])/2 ,
       tick = F)
 
 legend (x = 5, y = 2000,
-        col = c("blue", "black", "red"),
+        col = c(color_hahner, color_kim, color_luik),
         pch = 1,
-        legend = c("Luik triplets", "Hahner twins", "Kim twins"),
+        legend = c("Hahner twins", "Kim twins", "Luik triplets"),
         cex = 0.75,
         bty = "n"
         )

### 2017-02-24 simhalfwithage.R

@@ -80,7 +80,7 @@ p <- ggplot(filter(hahner_df_simhalfwithage, !is.na(time_diff)), aes(time_diff, 
   ylab("Count\n") +
   ylim(0, max_y_t) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
+  annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 minute:  ", less_than_min), size = 5, col = "red") +
   theme_bw()
 print(p)
 dev.off()

### 2017-02-24 simwithage.R

@@ -81,7 +81,7 @@ p <- ggplot(filter(hahner_df_simwithage, !is.na(time_diff)), aes(time_diff, fill
      ylab("Count\n") +
      ylim(0, max_y_t) + 
      scale_fill_manual(values = c(NA, "red"), guide = F) +
-     annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
+     annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 minute:  ", less_than_min), size = 5, col = "red") +
      theme_bw()
 print(p)
 dev.off()

### 2017-02-24 2-analyze.R
changed plotting paramerters and split the simulate section into two parts: "analyze" and "simulate"

@@ -257,49 +257,3 @@ dev.off()
 
 stop()
 
-
-max_x_t <- 2300
-max_y_t <- 500
-max_x_r <- 124
-max_y_r <- 300
-
-
-source("sim.r")
-# The race was simulated 10000 times.
-# The Hahner twins place consecutively in 2.71% of simulations
-# The average simulated time between them was 8.07 minutes
-# 95% simulated interval for Anna Hahner finishing time in seconds: 8486 10133 
-# 95% simulated interval for Lisa Hahner finishing time in seconds: 8604 10268 
-source("simhalf.r")
-# The race was simulated 10000 times.
-# The Hahner twins place consecutively in 2.71% of simulations
-# The average simulated time between them was 5.8 minutes
-# 95% simulated interval for Anna Hahner finishing time in seconds: 9468 10670 
-# 95% simulated interval for Lisa Hahner finishing time in seconds: 9387 10586 
-source("simwithage.r")
-# The race was simulated 10000 times.
-# The Hahner twins place consecutively in 1.82% of simulations
-# The average simulated time between them was 7.87 minutes
-# 95% simulated interval for Anna Hahner finishing time in seconds: 8369 9999 
-# 95% simulated interval for Lisa Hahner finishing time in seconds: 8507 10143 
-source("simhalfwithage.r")
-# The race was simulated 10000 times.
-# The Hahner twins place consecutively in 2.59% of simulations
-# The average simulated time between them was 5.85 minutes
-# 95% simulated interval for Anna Hahner finishing time in seconds: 9493 10717 
-# 95% simulated interval for Lisa Hahner finishing time in seconds: 9443 10637 
-
-
-# Save simulation results
-save(hahner_df_sim, 
-     hahner_df_simhalf, 
-     hahner_df_simwithage, 
-     hahner_df_simhalfwithage, 
-     file = "sim_results.RData")
-
-
-mean(hahner_df_simwithage$anna_final, na.rm = T)
-mean(hahner_df_simwithage$lisa_final, na.rm = T)
-mean(hahner_df_simwithage$anna_rank, na.rm = T)
-mean(hahner_df_simwithage$lisa_rank, na.rm = T)
-sum(hahner_df_simwithage$rank_diff == 1, na.rm = T)

### 2017-02-24 3-simulate.R

@@ -1,0 +1,64 @@
+library(MASS)
+library(tidyverse)
+library(lubridate)
+library(stringr)
+
+df <- read.csv("data/times.csv", stringsAsFactors = FALSE) %>% tbl_df()
+
+# Remove Irvette Van Zyl, since she did not start the race due to
+# injury.  Her bib number is 1172.
+df <- df[df$BIB != 1172,]
+
+
+max_x_t <- 2300
+max_x_r <- 124
+
+max_y_t <- 325
+max_y_r <- 200
+
+max_y_t_half <- 500
+max_y_r_half <- 300
+
+text_y <- .70
+text_x <- .60 
+
+
+source("sim.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 1.54% of simulations
+# The average simulated time between them was 8.07 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 8486 10133 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 8604 10268 
+source("simhalf.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 2.71% of simulations
+# The average simulated time between them was 5.8 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 9468 10670 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 9387 10586 
+source("simwithage.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 1.82% of simulations
+# The average simulated time between them was 7.87 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 8369 9999 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 8507 10143 
+source("simhalfwithage.r")
+# The race was simulated 10000 times.
+# The Hahner twins place consecutively in 2.59% of simulations
+# The average simulated time between them was 5.85 minutes
+# 95% simulated interval for Anna Hahner finishing time in seconds: 9493 10717 
+# 95% simulated interval for Lisa Hahner finishing time in seconds: 9443 10637 
+
+
+# Save simulation results
+save(hahner_df_sim, 
+     hahner_df_simhalf, 
+     hahner_df_simwithage, 
+     hahner_df_simhalfwithage, 
+     file = "sim_results.RData")
+
+
[…]

### 2017-02-24 sim.R

@@ -79,7 +79,7 @@ p<-ggplot(filter(hahner_df_sim, !is.na(time_diff)), aes(time_diff, fill = time_d
   ylab("Count\n") +
   ylim(0, max_y_t) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_t, x = text_x*max_x_t, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
   theme_bw()
 print(p)
 dev.off()
@@ -92,7 +92,7 @@ p2<-ggplot(filter(hahner_df_sim, !is.na(rank_diff)), aes(rank_diff, fill = rank_
   ylim(0, max_y_r) + 
   xlim(0, max_x_r) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_r, x = 80, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_r, x = text_x*max_x_r, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
   theme_bw()
 print(p2)
 dev.off()

### 2017-02-24 simhalf.R

@@ -77,9 +77,9 @@ p <- ggplot(filter(hahner_df_simhalf, !is.na(time_diff)), aes(time_diff, fill = 
   geom_histogram(breaks = seq(0,max_x_t, 30), colour = "black") +
   xlab("\nTime difference, in seconds") +
   ylab("Count\n") +
-  ylim(0, max_y_t) + 
+  ylim(0, max_y_t_half) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_t_half, x = text_x*max_x_t, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
   theme_bw()
 print(p)
 dev.off()
@@ -89,10 +89,10 @@ p2 <- ggplot(filter(hahner_df_simhalf, !is.na(rank_diff)), aes(rank_diff, fill =
   geom_histogram(binwidth=1, center=0, colour = "black") +
   xlab("\nRank difference") +
   ylab("Count\n") +
-  ylim(0, max_y_r) + 
+  ylim(0, max_y_r_half) + 
   xlim(0, max_x_r) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_r, x = 80, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_r_half, x = text_x*max_x_r, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
   theme_bw()
 print(p2)
 dev.off()

### 2017-02-24 simhalfwithage.R

@@ -78,9 +78,9 @@ p <- ggplot(filter(hahner_df_simhalfwithage, !is.na(time_diff)), aes(time_diff, 
   geom_histogram(breaks = seq(0,max_x_t, 30), colour = "black") +
   xlab("\nTime difference, in seconds") +
   ylab("Count\n") +
-  ylim(0, max_y_t) + 
+  ylim(0, max_y_t_half) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 minute:  ", less_than_min), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_t_half, x = text_x*max_x_t, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
   theme_bw()
 print(p)
 dev.off()
@@ -90,10 +90,10 @@ p2 <- ggplot(filter(hahner_df_simhalfwithage, !is.na(rank_diff)), aes(rank_diff,
   geom_histogram(binwidth=1, center=0, colour = "black") +
   xlab("\nRank difference") +
   ylab("Count\n") +
-  ylim(0, max_y_r) + 
+  ylim(0, max_y_r_half) + 
   xlim(0, max_x_r) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_r, x = 80, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_r_half, x = text_x*max_x_r, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
   theme_bw()
 print(p2)
 dev.off()

### 2017-02-24 simwithage.R

@@ -81,7 +81,7 @@ p <- ggplot(filter(hahner_df_simwithage, !is.na(time_diff)), aes(time_diff, fill
      ylab("Count\n") +
      ylim(0, max_y_t) + 
      scale_fill_manual(values = c(NA, "red"), guide = F) +
-     annotate("text", y = .8*max_y_t, x = 1500, label = paste0("< 1 minute:  ", less_than_min), size = 5, col = "red") +
+     annotate("text", y = text_y*max_y_t, x = text_x*max_x_t, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
      theme_bw()
 print(p)
 dev.off()
@@ -94,7 +94,7 @@ p2 <- ggplot(filter(hahner_df_simwithage, !is.na(rank_diff)), aes(rank_diff, fil
   ylim(0, max_y_r) + 
   xlim(0, max_x_r) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = .8*max_y_r, x = 80, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_r, x = text_x*max_x_r, label = paste0("Consecutive finishes:  ", consecutive), size = 5, col = "red") +
   theme_bw()
 print(p2)
 dev.off()

### 2017-02-24 paper/paper.tex
a couple changes

@@
 consistent with Anna's observed finishing time in Rio, which was 9,932
-seconds. In fact, our simulations estimate that Anna should have
-finished the Rio marathon in 9,181 seconds, which is slightly more
-than 12 minutes faster than her actual time. Lisa Hahner's
-corresponding 95\% interval is 8,507 to 10,143 with an observed
+seconds. Nonetheless, her finishing time was on the slower end of this interval,.   
+In fact, our simulations estimate that, given Anna's personal best marathon time
+and her age, she would be expected to finish the Rio marathon in 9,181 seconds, which is slightly more
+than 12 minutes faster than her actual time.  This is similar to her sister's result. Lisa Hahner's
+corresponding 95\% interval is 8,507 to 10,143 with a mean of 9,319.  Her observed
 finishing time of 9,933 seconds.

### 2017-02-24 paper/paper.tex
a couple final edits

@@
 locations are not extreme.  One residual is located at the 87th
-percentile and the other at the 92rd.  While these two residuals are
+percentile and the other at the 92nd.  While these two residuals are
 in the tail end of the residual distribution, they are not major
@@
 assume that each runner's marathon time \Yi\ is distributed normally
-with a mean that is a linear function of the runner's marathon
-skill $X_i$ and her age $Z_i$. Hence:
+with a mean that is a function of the runner's marathon
+skill - which we assume can be determined by a linear combination of her personal best marathon time $X_i$ and her age $Z_i$. Hence:
 $$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
@@
 where $N\left(\cdot,\cdot\right)$ denotes a normal
-distribution. Taking each runner's personal best marathon time as a
-proxy for her underlying marathon skill, we estimate $\beta_0$,
+distribution. We estimate $\beta_0$,
 $\beta_1$, $\beta_2$, and $\sigma$ using ordinary least squares on
@@
 consistent with Anna's observed finishing time in Rio, which was 9,932
-seconds. Nonetheless, her finishing time was on the slower end of this interval,.   
+seconds. Nonetheless, her finishing time was on the slower end of this interval.   
 In fact, our simulations estimate that, given Anna's personal best marathon time
@@
 The bottom line here is that our simulated marathon finishes are
-consistent with the Hahner twins' finishing times. And note that the
+consistent with the Hahner twins' finishing times in that their finishes were inside
+the bounds of traditional 95\% confidence intervals. And note that the
 regression model underlying the simulations was estimated without the

### 2017-02-25 paper/chance_memo.tex
edits prior to submitting revision

@@
   \textbf{To:} &  Scott Evans, editor, \emph{CHANCE}\\
-  \textbf{From:} & David Cottrell and Michael C.\ Herron, Dartmouth College\\
+  \textbf{From:} & David Cottrell and Michael C.\ Herron\\
   \textbf{Date:} & \today \\
@@
   accordingly dropped the quadratic terms in the scatter plots on p.\
-  4 and, implicitly, from the Studentized residual plots on p.\ 6.  In
-  terms of the latter suggestion about, Figure 1b contains a plot of
-  half marathon split times versus actual finishing times.
+  4 and, implicitly, from the Studentized residual plots on p.\ 6.
+  Regarding the latter suggestion concerned half marathon splits and
+  observed finishing times, Figure 1b contains such a plot.
   %% Note: xr labels are
@@
   all things equal.
+
+  The blindfolding question is addressed in Figures 5b and 6b.  In
+  short, a consecutive Hahner twin finish assuming independence (that
+  is, blindfoldedness) would have been very unlikely.
 

### 2017-02-25 paper/paper.tex
edits prior to submitting revision

@@
 Kurschilgen averred that the twins slowed down so as to finish
-simultaneously and create a spectacle which would ``generate media
-attention.''  Kurschilgen justified his charge with the fact that the
-twins ran the Rio marathon over 18 minutes slower their personal best
-times prior to the Olympics.  Kurschilgen's accusations were denied by
-the Hahner twins, perhaps not surprisingly, who claimed that their
-joint finish was simply a fortuitous coincidence.
+simultaneously and create a spectacle, and he justified his charge by
+noting that the Hahners ran the Rio marathon over 18 minutes slower their
+personal best times prior to the Olympics.  Kurschilgen's accusations
+were denied by the Hahner twins, perhaps not surprisingly, who claimed
+that their joint finish was simply a fortuitous coincidence.
 
@@
 have injured themselves on the course and accordingly dropped out, and
-others may have dropped out, uninjured, in anticipation of an
-unsatisfactory result. Of the 156 marathon starters, 133 completed the
+others may have stopped running, uninjured, in anticipation of an
+unsatisfactory result. Of 156 marathon starters, 133 completed the
 race and 23 DNFed at various locations throughout the course. The
@@
 Our research design is twofold.  First, we present visualizations that
-describe various features of the 2016 women's Olympic marathon, and an
-important element in our visualizations is the difference between
-runners' Rio times and their prior personal best marathon times.  Our
+describe various features of the 2016 women's Olympic marathon, and
+among other things our visualizations feature difference between
+runners' Rio times and their prior personal best marathon times.  The
 visualizations suggest that the Hahner twin's pace in the marathon was
@@
 Olympic marathon as a secondary measure of athletic skill in
-marathoning.  Our two measures of marathon talent are featured in the
-visualizations below.
+marathoning.
 
-Figure \ref{fig:scatter} contains two plots, both of which describe
-how finishing times from the Rio women's marathon varied as a function
-of athletes' personal best times and half marathon split times.  The
-points in the plots are colored by twin/triplet status, and both plots
+Figure \ref{fig:scatter} contains two plots describing how finishing
+times from the Rio women's marathon varied as a function of athletes'
+personal best times and half marathon split times.  The points in the
+plots are colored by twin/triplet status, and both plots
 contain %second-order polynomial regression lines.
@@
 middle of the distribution of points and therefore imply that the
-twins do not seem to have slowed down in any unusual way.  This is
-consistent with the previous Figure \ref{fig:45degreeplot} and
-inconsistent with accusations made against the Hahner twins, at least
-the part of the accusation that focused on their overall pace.
+twins do not seem to have slowed down dramatically after their reached
+the halfway point of the Rio marathon.  This is consistent with the
+aforementioned Figure \ref{fig:45degreeplot} and inconsistent with
+accusations made against the Hahner twins, at least the part of the
+accusation that focused on their overall pace.
 
@@
 their official finishing times were separated by one second.  Is such
-a 115 to one compression typical among pairs of runners?  Moreover,
+a 115 to one compression typical among pairs of runners?  Relatedly,
 are there other pairs of Olympic marathoners who had a difference
@@
 differences in personal best times, and pairs of twins/triplets are
-identified by the same color scheme we used earlier.  Consider first
+identified by the same color scheme we used earlier (recall that only
+two of the Luik triplets finished the Rio marathon).  Consider first
 the Hahner twins.  The two German women are effectively located on the
@@
 and this shows that, conditional on an approximate 115 second
-difference in personal best times, most marathoners did not have close
-finishing times like the Hahners.  Some pairs of runners with around
-115 second personal best differences had finishing time differences of
-1,000 seconds, i.e., in excess of 15 minutes.  The points in Figure
-\ref{fig:diffdiffscatter} are not independent, but they provide a
-rough sense of the dispersion in finishing time differences between
-runners that one might expect conditional on differences in personal
-best times.
+difference in personal best times, most marathoner pairs did not have
+close finishing times like the Hahners.  Some pairs of runners with
+around 115 second personal best differences had finishing time
+differences of 1,000 seconds, i.e., in excess of 15 minutes.  The
+points in Figure \ref{fig:diffdiffscatter} are not independent, but
+they provide a rough sense of the dispersion in finishing time
+differences between runners that one might expect conditional on
+differences in personal best times.
 
@@
 at various split times on the marathon course.  Each dot in the
-figure---color as before---depicts a recorded split and the number of
+figure---colored as before---depicts a recorded split and the number of
 seconds each runner was behind the race leader at the time.  There are
@@
 visually apparent because the Kim twins had identical split times
-during the entire marathon.
+during the entire marathon.  This explains why we used the term
+``lockstep'' to refer to the Kim twin's marathon pace.
 
@@
 assume that each runner's marathon time \Yi\ is distributed normally
-with a mean that is a function of the runner's marathon
-skill - which we assume can be determined by a linear combination of her personal best marathon time $X_i$ and her age $Z_i$. Hence:
+with a mean that is a function of the runner's marathon skill, which
+we assume is a linear combination of her personal best marathon time
+$X_i$ and her age $Z_i$:
 $$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
@@
   matrix, and estimated regression variance from this model.
-\item For each simulated rate, draw intercept and slope estimates
+\item For each simulated race, draw intercept and slope estimates
   $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
@@
   standard deviation of the original regression model's residuals.
+  This step requires 156 draws from a normal distribution, one draw
+  per marathoner.
 \item Predict each runner's final result by combining the randomly
@@
 As these steps illustrate, our simulation repeatedly draws random
-coefficient vectors, and this captures variability in what we know
+coefficient vectors, and this captures uncertainty in what we know
 about the relationship between runner talent and age and runner finish
-times.  In addition, the simulation draws random disturbances for each
-runner, conditional on our original estimate of regression variance;
-these disturbances capture variability in runner finishing times,
-notwithstanding underlying marathon talent as proxied for by prior
-best finishing time.  Importantly, the disturbances that we draw are
-independent across runners.  Consequently, for each simulated race the
-finishing order among runners will vary.
+times.  In addition, for each simulated race our simulation draws
+random disturbances for each runner, conditional on the original
+estimate of regression variance; these disturbances capture
+variability in runner finishing times, notwithstanding age and
+underlying marathon talent.  Importantly, the disturbances that we
+draw are independent across runners.  Consequently, for each simulated
+race the finishing order among runners will vary.
 
@@
 consistent with Anna's observed finishing time in Rio, which was 9,932
-seconds. Nonetheless, her finishing time was on the slower end of this interval.   
-In fact, our simulations estimate that, given Anna's personal best marathon time
-and her age, she would be expected to finish the Rio marathon in 9,181 seconds, which is slightly more
-than 12 minutes faster than her actual time.  This is similar to her sister's result. Lisa Hahner's
-corresponding 95\% interval is 8,507 to 10,143 with a mean of 9,319.  Her observed
-finishing time of 9,933 seconds.
+seconds. Nonetheless, Anna's finishing time was on the slower end of
+this interval.  In fact, our simulations estimate that, given Anna's
+personal best marathon time and her age, she would be expected to
+finish the Rio marathon in 9,181 seconds, which is slightly more than
+12 minutes faster than her actual time.  This is similar to her
+sister's result. Lisa Hahner's corresponding 95\% interval is 8,507 to
+10,143 with a mean of 9,319.  Her actual finishing time was 9,933
+seconds.
 
 The bottom line here is that our simulated marathon finishes are
-consistent with the Hahner twins' finishing times in that their finishes were inside
-the bounds of traditional 95\% confidence intervals. And note that the
-regression model underlying the simulations was estimated without the
-Hahner twins (and same for the Kim twins and Luik triplets). According
-to our simulation, then, both Hahner twins did not run appreciably
-slowly, conditional on personal best times prior to the Rio Olympics
-and age.
+consistent with the Hahner twins' finishing times in that their
+finishes were inside the bounds of traditional 95\% intervals. And
+note that the regression model underlying the simulations was
+estimated without the Hahner twins (and same for the Kim twins and
+Luik triplets). According to our simulation, then, both Hahner twins
+did not run appreciably slowly, conditional on personal best times
+prior to the Rio Olympics and age.
 
@@
 finish within a minute of each other.  The histogram area associated
-with this result is depicted in red in Figure
+with this latter result is depicted in red in Figure
 \ref{fig:simulatedfinishtimes}.  Moreover, the Hahner twins finished
@@
 appreciably during the Rio marathon. Their times were not fast, but
-they were within reason for runners of the Hahners' abilities.
+they were within reason for runners of the Hahners' abilities and age.
 Locally, though, we find that the Hahner's finish was in fact
-contrived. Their finish---one second between the two women---was an
-extremely low probability event. Compared to their differences in
-talent, the Hahner twins difference in finishing times was unusually
-compressed. This is evidence that their finishing had elements of
-intentionality---perhaps at the last minute but intentionality
-nonetheless.
+contrived. Their finish---consecutive, with one second between the two
+women---was a rather low probability event. Compared to their
+differences in talent, that is, the Hahner twins difference in
+finishing times was unusually compressed. This is evidence that their
+finishing had elements of intentionality---perhaps at the last minute
+but intentionality nonetheless.
 
@@
 and compared to the doping allegations that presently surround
-endurance sports in general, what the Hahners did seems relatively
-tame. Still, it might have behooved them to be a bit more open about
-their end-of-race tactics, and to this end we have shown here how a
-simple data analysis can shed light on claims about racing results.
-Should the 2020 Summer Olympics feature twins and triplets again, we
-look forward to a comparative analysis using the techniques
-illustrated here.
+endurance sports in general, what the Hahners appear to have done
+seems relatively tame. Still, it might have behooved them to have been
+a bit more open about their end-of-race tactics, and to this end we
+have shown here how a simple data analysis can shed light on claims
+about racing results.  Should the 2020 Summer Olympics again feature
+marathoning twins and triplets, we look forward to a comparative
+analysis using the techniques illustrated here.
 

### 2017-02-25 simhalfwithage.R

@@ -80,7 +80,7 @@ p <- ggplot(filter(hahner_df_simhalfwithage, !is.na(time_diff)), aes(time_diff, 
   ylab("Count\n") +
   ylim(0, max_y_t_half) + 
   scale_fill_manual(values = c(NA, "red"), guide = F) +
-  annotate("text", y = text_y*max_y_t_half, x = text_x*max_x_t, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
+  annotate("text", y = text_y*max_y_t_half, x = text_x*max_x_t, label = paste0("< 1 minute:  ", less_than_min), size = 5, col = "red") +
   theme_bw()
 print(p)
 dev.off()

### 2017-02-25 simwithage.R

@@ -81,7 +81,7 @@ p <- ggplot(filter(hahner_df_simwithage, !is.na(time_diff)), aes(time_diff, fill
      ylab("Count\n") +
      ylim(0, max_y_t) + 
      scale_fill_manual(values = c(NA, "red"), guide = F) +
-     annotate("text", y = text_y*max_y_t, x = text_x*max_x_t, label = paste0("< 1 min:  ", less_than_min), size = 5, col = "red") +
+     annotate("text", y = text_y*max_y_t, x = text_x*max_x_t, label = paste0("< 1 minute:  ", less_than_min), size = 5, col = "red") +
      theme_bw()
 print(p)
 dev.off()

### 2017-02-25 paper/chance_memo.tex
minor memo typo

@@
   referee is requesting.  In the marked-up PDF document, the word
-  ``likely'' appears over the following sentence: ``Their finish—one
-  second between the two women—was an extremely low probability
+  ``likely'' appears over the following sentence: ``Their finish---one
+  second between the two women---was an extremely low probability
   event.''  We are not sure how to incorporate the referee's request

### 2017-03-03 paper/paper.tex
edits for chance

@@
 marathoners ran slower in Rio compared to their prior personal bests.
-The Hahner twins were definitely on the slow side, well above the 45-degree line,
-but a number of runners had even greater differences between
-their Olympic times and their personal bests than the Hahners.  These
-runners are denoted with squares in Figure \ref{fig:45degreeplot}, and
-there are 13 such symbols, highlighting approximately 10\% of the
-racers who completed the Rio marathon.  Moreover, the figure shows
-that there were two women who had personal best times slightly faster
-than the Hahner's and yet finished after the German women.  Although
-Figure \ref{fig:45degreeplot} suggests that the Hahner twins were
-slower than one would have expected given their previous best marathon
-times, it is not consistent with the accusation that they dramatically
-slowed down in the Rio marathon.
+The Hahner twins were definitely on the slow side, well above the
+45-degree line, but a number of runners had even greater differences
+between their Olympic times and their personal bests than the Hahners.
+These runners are denoted with squares in Figure
+\ref{fig:45degreeplot}, and there are 13 such symbols, highlighting
+approximately 10\% of the racers who completed the Rio marathon.
+Moreover, the figure shows that there were two women who had personal
+best times slightly faster than the Hahners and yet finished after the
+two German women.  Although Figure \ref{fig:45degreeplot} suggests
+that the Hahner twins were slower than one would have expected given
+their previous best marathon times, it is not consistent with the
+accusation that they dramatically slowed down in the Rio marathon.
 
@@
 horizontal axis because their difference in finishing times is one
-second.  However, there are many points above the Hahner's black dot,
+second.  However, there are many points above the Hahners' black dot,
 and this shows that, conditional on an approximate 115 second
@@
 they were within reason for runners of the Hahners' abilities and age.
-Locally, though, we find that the Hahner's finish was in fact
+Locally, though, we find that the Hahners' finish was in fact probably
 contrived. Their finish---consecutive, with one second between the two
@@
 Our goal is not to speak to the question of whether the Hahner twins
-should or should not have enjoyed a somewhat contrived moment of
-Olympic glory. Neither was in contention for a podium finish in Rio,
-and compared to the doping allegations that presently surround
+should or should not have enjoyed what some might call an artificial
+moment of Olympic glory. Neither was in contention for a podium finish
+in Rio, and compared to the doping allegations that presently surround
 endurance sports in general, what the Hahners appear to have done
@@
 analysis using the techniques illustrated here.
-
-% \section*{The probability of an unintentional simultaneous finish}
-
-% Thomas Kurschilgen's accusations of an intentional finish stems from
-% two observations. First, the Hahner twins finished slower than he
-% expected given respective personal bests. Second, the twins finished
-% together at the exactly same time. Kurschilgen clearly believed that
-% neither of these events would have occurred had both Hahner run the
-% marathon independently, absent coordination. By his logic, the German
-% twins should have run faster and not have finished simultaneously.
-
-% On the other hand, only a handful of runners completed the Rio
-% marathon with times that were faster than their recorded best times.
-% Thus, 18 minutes behind a personal record may not be the outlier that
-% Kurschilgen claimed it to be. Moreover, there is good reason to
-% suspect that an unintended simultaneous finish was more likely for the
-% Hahner twin than it would be for any other set of runners. After all,
-% these two women are twins with presumably similar abilities. Not only
-% do they train together, but their pre-Rio best times are less than two
-% minutes apart. While Anna may be slightly faster than Lisa measured by
-% personal bests, we might expect the difference in their Rio result to
-% be just as close as the differences in their recorded bests. And given
-% random variation in finishing times, a simultaneous finish might not
-% be out of the ordinary.
-
-% Even though they seem similar, finishing at very similar times and
-% finishing together are different phenomena. If, for example, both
-% Hahners were of similar ability to each other and also to many other
-% runners, then we might expect similar finishing times yet not
-% necessarily similar placements. The latter will be a function of the
-% extent to which all runners on the Rio course have similar talent
-% levels.  This point is an important one and will be evident in the
-% results that follow.
-
-% To test the assertion that the Hahner twins paced themselves to finish
-% at the same time and with back-to-back placements, we need to compute
-% the probability that such a result would have occurred
-% unintentionally. Hence we need to know the probability distribution of
-% Anna and Lisa's finishing times if their runs had been independent of
-% each other. The challenge, of course, is that this distribution is
-% unknown.
-
-% \section*{Modeling}
-
-% The Hahner twins either acted independently $I$ or coordinated their
-% times $C$; these are the two possible states of the race
-% $\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
-% times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
-% that they ran independently, as they say they did. Hence, we are
-% looking to determine,
-
-% $$P(\psi = I \mid y_A  \cap y_L )$$
-
-% However, to determine this, we need to have some understanding of a
-% likelihood function that specifies Rio finishing times. We want to
-% know the likelihood of Anna's and Lisa's final times given
-% independence $P(y_A \cap y_L \mid \psi = I )$. We can estimate this
-% function with a few assumptions. First, we assume that under
-% independence, any given runner's final time $Y_i$ is conditional on
-% his/her running ability plus noise. Specifically, we assume that the
-% $Y_i$ is a linear function of the runner's ability $X_i$ plus a
-% normally distributed error term $e_i \sim N(0, \sigma)$. Second, we
-% assume that every runner shares the same linear relationship - meaning
-% the slope and intercept remain constant across runners. Third, we
-% assume that the error term is drawn from a common distribution across
-% runners. Hence, luck and misfortune are drawn from the same
-% distribution. Therefore,
-
-% $$Y_i \sim N(\beta_0 + X_{i}\beta_1 + e_i, \sigma)$$
-
-% We also assume that a runner's ability $X_i$ can be measured precisely
-% by her best marathon performance leading up to the Olympics.
-
-% %% Any measurement error must therefore be negligible.
-
-% If Anna and Lisa intentionally slowed down as a result of coordination
-% then we would likely observe $y_A > E(Y_A \mid \psi = I) $ and
-% $y_L > E(Y_L\mid \psi = I )$. In other words, the final times that
-% Anna and Lisa recorded in the race would be greater than we would
-% expect if they had run independently.
-
-% Moreover, if Anna and Lisa coordinated to finish simultaneously, then
-% the difference between the two sisters' final times would be less that
-% the expected difference had they run independently. Therefore, under a
-% coordinated finish we would expect $\left|y_A - y_L\right| < E(\left|Y_A - Y_L\right| \mid \psi = I )$.
-
-% \section*{Did Anna and Lisa intentionally slow down?}
-
-% According to Kurschilgen, Anna and Lisa underperformed in the Rio marathon.
-He claimed that because their goal was to finish simultaneously rather than finish at their fastest pace, their times were slower than they otherwise would have been.
-He claimed that the twins were simply trading speed for a photo-finish.
-
-% If this were the case, the function generating Anna and Lisa's final times would deviate from the function that generated everyone else's final times.
-Given everyone else would draw their times from the independent distribution  $Y_i \sim N(\beta_0 + X_{i}\beta_1 + e, \sigma)$,  Anna and Lisa would draw from a distribution of times that are slower in expectation.
-Hence, they would lie well-above the line that links a runner's performance in Rio to their previous best performance.
-
-% \begin{figure}[!ht]
-%   \caption{Relationship between Personal Best and Result  -- caption?}
-%   \label{fig:scatter}
-%   \begin{subfigure}{.5\textwidth}
-%     \centering
-%     \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
-%     \caption{Rio finishing times and personal best times}
-%     \label{fig:45degreeplot}
-%   \end{subfigure}
-%   \begin{subfigure}{.5\textwidth}
-%     \centering
-%     \includegraphics[width=.975\textwidth, keepaspectratio]{studentized_residuals.pdf}
-%     \caption{Studentized residuals}
-%     \label{fig:studentizedresiduals}
-%   \end{subfigure}
-% \end{figure}
-
-% Figure \ref{fig:45degreeplot} displays the relationship between a
-% runner's personal best (horizontal axis) and her final result
-% (vertical axis).  The 133 finishers are plotted as grey points, and
-% the 23 DNFs are depicted below along the horizontal axis with
-% tick-marks.  The dashed black line with grey confidence intervals
-% represent ordinarily least squares estimates of the linear
-% relationship between racer's finishing and personal best times.
-
-% One can see in Figure \ref{fig:scatter} that there is a clear
-% relationship between a runner's personal best time and her performance
-% in the Olympics.  While there is a significant amount of noise in the
-% relationship, the Hahner twins' dots are well above the regression's
-% fitted line notwithstanding this point.  They finished the Rio
-% marathon much slower than they should have, per their pre-Olympic
-% personal best times.  In fact, if we sort studentized residual, we can
-% see in Figure \ref{fig:studentizedresiduals} that the Hahner twins are
-% in the tail-end of the residual distribution.  Anna Hahner's residual
-% deviation is greater than 91.7\% of the runners who completed the
-% marathon while Lisa Hahner's deviation is greater than 86.5\%.  Hence,
-% the twins appear to have finished at a much slower pace than expected,
-% which is what we might expect if they were coordinating their runs.
-% The North Korean Kim twins are also in the tail of the residual
-% distribution---but the left tail.  These twins finished much faster
-% than expected.
-
-% \section*{Did Anna and Lisa intentionally finish together?}
-
-% Anna and Lisa finished the marathon at an unusually slow rate, which is what we might expect if the twins intentionally slowed their pace to finish simultaneously.
-However, the observation of a slow finish does not necessarily imply that it was an intentional result .
-It only suggests that Kurschilgen's claim regarding this issue holds up in the data.
-
-% Perhaps it would be easier to refute the claim if we could show that the Twins's back-to-back finish was not such an unusual event.
-Given that Anna and Lisa were so close in their personal best times, it may seem reasonable to expect that they would also be close in their finishing times.
-Hence, one way to determine if the twins's finish was intentional is to quantify the likelihood that such a close finish would occur by chance given the differences in their ability.
-
-% We can attempt to quantify this likelihood by looking at how differences in the personal best time of two runners translate into differences in Olympic times .
-If Anna and Lisa's finish was unintentional we'd expect that such a close finish would be unlikely given their difference in personal best times.
-Yet, how does the difference in personal best times relate to the difference in finishing times?
-
-% One way we can estimate this relationship is to look at all of the differences that we observe in the marathon.
-We can take the difference in personal best time for every combination of runners in the data and look at the difference in their final times.
-We should expect that compared to the difference in their personal best times, the difference in Anna and Lisa's final times should be commensurate with the conditional difference of all other runners.
-%This needs a better explanation.
-
-% Therefore, we find all combinations of runners in the marathon and calculate two quantities: 1) the difference in their personal best time and 2) the difference in their final time.
-We plot the relationship between the two quantities in Figure \ref{fig:diffdiffscatter}.
-All 8,778 dyadic combinations of runners are displayed by grey points in the scatter plot.
-The solid black line indicates a one-to-one relationship, where every point would fall if the difference in final times were systematically equivalent to the difference in personal bests.
-
-% % \begin{figure}[!ht]
-% %  \caption{Relationship between Difference in Personal Best and Difference in Result}
-% %  \label{fig:diffdiffscatter}
-% %  \centering
-% %  \begin{subfigure}{.5\textwidth}
-% %  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
-% %  \end{subfigure}
-% % \end{figure}
-
-% We can see from the yellow dot that the Hahner Twins are just below this one-to-one line.
-Hence, the twins' final times were closer to each other than their personal best times; but not by much.
-In fact, this might be an expected difference if we expected there to be a one-to-one relationship.
-However, although there is a clear relationship between the difference in final time and the difference in personal best, it is not a strong one-to-one relationship.
-We've added a dotted smoother to display a cubic splines estimate of the conditional averages.
-While the Hahner twins may be close to the one-to-one line, they are much further away from the average set of runners.
-Hence, we would expect for any pair of runners, the difference in their final results should be much greater than their difference in their personal best.
-However, the Hahner twins do not fit this expectation.
-Yet, how different from this expectation are they?
-
-% One way we can quantify the degree to which the twins are outliers is by comparing their difference in difference to every other pair of runner's difference-in-differences.
-We calculate the difference-in-differences for each pair of runners by simply subtracting their difference in result (y-axis) from their difference in personal best (x-axis).
-Therefore positive difference-in-differences mean that the runners ended up further apart than their difference in personal bests.
-Negative difference-in-differences mean that the runners end up closer together.
-Hence, the Hahner twins have a difference-in-difference of -114 seconds.
-They finished 114 seconds closer together than the distance between their personal bests.
-
-% We arrange all 8,778 difference-in-differences from small to large in the first plot in Figure \ref{fig:diffdiff}.
-The Hahner twins are denoted by the yellow point.
-Using this plot we can visualize where the twins fall within the distribution of difference-in-differences.
-It is clear that they are closer to the lower tail of the distribution.
-However, compared to all other difference-in-differences, their result is not that extreme.
-Nearly 21\% of the difference-in-dfferences were less than the that of the twins'.
-Hence, we'd expect at least one in five pairs of runners to have a reduction in difference at least as great as the twins.
-
-% However, if we limit the observations to only those runners who had personal best times that were at least as close to each other as the Hahner twins, we get a different perspective.
-Among those runners,  the Hahner twins reduced the distance between them to the greatest extent.
-There was no other pair of runners whose reduced the difference in the final time relative to their difference in personal bests more than them.
-
-% \begin{figure}[!ht]
-%  \caption{How Hahner Twins Rank in Difference-in-Differences}
-%  \label{fig:diffdiff}
-%  \begin{subfigure}{.5\textwidth}
-%  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_1.pdf}
-%  \end{subfigure}
-%  \begin{subfigure}{.5\textwidth}
-%  \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_2.pdf}
-%  \end{subfigure}
-% \end{figure}
 

### 2017-03-03 2-analyze.R

@@ -247,7 +247,7 @@ ggplot(df2c, aes(PERCENTILE,diff_in_diff)) +
   geom_point(data = na.omit(df2c), aes( col = twins)) + 
   geom_text(data = df2c[df2c$twins == "Hahner twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_y = 100, colour = color_hahner) +
   geom_text(data = df2c[df2c$twins == "Kim twins",], aes(label = paste(round(PERCENTILE,1),"%")), nudge_y = 100, colour = color_kim) +
-  ggtitle("Distribution of Difference in Differences\n(Diads where difference in PB < Hahner's)") +
+  ggtitle("Distribution of Difference in Differences\n(Dyads where difference in PB < Hahner's)") +
   ylab("Difference in Difference (sec)\n") +
   xlab("\nPercentile") + 
   scale_color_manual(values = c(color_hahner, color_kim, color_luik, "gray"), name = "") +
@@ -255,5 +255,5 @@ ggplot(df2c, aes(PERCENTILE,diff_in_diff)) +
   theme(legend.position = c(.85,.15))
 dev.off()
 
-stop()
+
 
